### 2020-03-24 draft.tex
ajout .tex

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={Numerical investigation of two-phase flow in high-permeability porous media Effect of permeability variation on traction terms},
            pdfauthor={Maxime Cochennec¹; Hossein Davarzani¹* (h.davarzani@brgm.fr); Yohan Davit²; Ioannis Ignatiadis¹; Michel Quintard²},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{longtable}}{}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{setspace}
\usepackage{cleveref}
\usepackage{xcolor}
\doublespacing

\title{Numerical investigation of two-phase flow in high-permeability porous
media Effect of permeability variation on traction terms}
\author{Maxime Cochennec¹ \and Hossein Davarzani¹* (h.davarzani@brgm.fr) \and Yohan Davit² \and Ioannis Ignatiadis¹ \and Michel Quintard²}
\date{\today{}}

\begin{document}
\maketitle

¹French Geological Survey, Orléans, France ²Institut de Mécanique des
Fluides de Toulouse, Toulouse, France

*To whom correspondence should be addressed

\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

In this paper we study the effects of varying permeability on traction
terms exerted at the different boundaries for two-phase flow in
Hele-Shaw cell. The traction terms, or surface force, are. surface
integrals that are necessary to close the continuous macroscopic
momentum balance equations for two-phase flow in porous media. These
terms are usually modeled from experimental results or obtained by
solving a closure problem (a set of boundary value problems to solve on
a representative periodic unit-cell). Here these terms are calculated
directly by means of direct numerical simulations, and the value of
these terms on each interface is studied as a function of the varying
permeability and capillary number. We focus on two-phase displacement in
high-permeability porous media for which the characteristic flow regimes
implies a non-negligible extent of the fluid-fluid interface and thus;

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{two-phase-flow-in-high-permeability-porous-media}{%
\subsection{Two-phase flow in high-permeability porous
media}\label{two-phase-flow-in-high-permeability-porous-media}}

An accurate description of two-phase flow in high-permeability porous
media is of major importance for several practical applications. One can
mention, among others, soil remediation in gravely soils
{[}\protect\hyperlink{ref-fetter2017contaminant}{1}{]}, nuclear safety
{[}\protect\hyperlink{ref-clavier2017modeling}{2}{]} or hydrodynamic of
catalytic fixed bed reactors
{[}\protect\hyperlink{ref-Santos1991}{3}{]}. However, most of the
literature is dedicated to two-phase flow in low-permeability porous
media.

Due to the larger pore size, two-phase flow in high-permeability porous
media (hereafter, high-permeability porous media refers to media for
which the characteristic particle diameter is about one millimeter and
above) results in a complex interaction between capillary, gravity and
viscous forces {[}\protect\hyperlink{ref-davit2018one}{4}{]}. Because of
this complex interplay, the characteristic flow regimes are very
different from those observed for surface tension dominated flow
{[}\protect\hyperlink{ref-dullien2012porous}{5}{]}. For low-permeability
medium, if the surface tension between the fluids is not too low and the
viscosity not too high, that is the capillary number is low (usually
inferior to \(10^{-3}\)), the fluid repartition pattern is described as
two independent flow streams separate by a multitude of stable meniscus
at steady state, as illustrated in \cref{fig:fluidPatterns} (a). In this
case, for a completely wet medium, the wetting phase flows in the small
pores while the non-wetting phase tends to occupy the larger pores. For
high permeability porous media, the fluid velocity can be high enough
that the viscous forces dominate over capillary forces (alike gravity
and inertial effects may become important if Bond and Reynolds numbers
are high, respectively). Then, the fluids repartition patterns can take
two forms, either the non-wetting phase is continuous, see for example
\cref{fig:fluidPatterns} (b), or is flowing as droplets or ganglia as in
\cref{fig:fluidPatterns} (c), while the wetting phase is flowing as a
film in contact with the solid. In both cases, the two fluids occupy
most of the pores at the same time and the non-wetting phase flows at
the center of the pores surrounded by the wetting phase. One can note
that these flow regimes are equivalent to those observed in two-phase
flows in a tube (slug flow, bubbly flow). Strictly speaking, these
different regimes must be considered when attempting to describe
two-phase flows with continuous macroscopic equations. Indeed, it has
been shown on numerous works that the overall flow depends on the flow
regimes
{[}\protect\hyperlink{ref-Avraam1995a}{6},\protect\hyperlink{ref-armstrong2016beyond}{7}{]}.
In the first glance, one would consider that the exchange terms between
the fluid phases as negligible compared to their counterpart between the
fluid phases and the solid phase for surface-tension dominated flwo and
that the extent of the fluid-fluid interface is small, as in
\cref{fig:fluidPatterns} (a). On the other hand, and this is what we are
interested in here, this is not necessarily the case for regimes
specific to flow in high-permeability porous media for which the extent
of the fluid-fluid interface is large. This is important because, as we
will see in the next section, these exchange terms between phases are
the basis of any attempt to establish continuous relationships on a
macroscopic scale starting from the pore scale.

\begin{figure}
\hypertarget{fig:fluidPatterns}{%
\centering
\includegraphics{figures/pdf/dessin.pdf}
\caption{Illustration of possible fluids dispatching in a 2D porous
network with solid phase in black, fluid 1, that stands for the
non-wetting phase (gray) and the wetting fluid (white), (a) the two
fluids are flowing in different channels separate by numerous meniscus
(b) wetting and non-wetting fluid are flowing together in most of the
pores as two continuous streams and (c) both fluids are flowing together
in most of the pores and the non-wetting phase is discontinuous -
Adapted from
{[}\protect\hyperlink{ref-dullien2012porous}{5}{]}}\label{fig:fluidPatterns}
}
\end{figure}

\hypertarget{continuous-model}{%
\subsection{Continuous model}\label{continuous-model}}

Here we present two different approaches to treat the continuous
macroscopic momentum equations for two-phase flow. Although
fundamentally identical, these approaches were developed for the needs
of two different specialties, namely, the study of groundwater flows and
the study of the hydrodynamic of chemical reactors.

\hypertarget{porous-media-approach}{%
\subsubsection{Porous media approach}\label{porous-media-approach}}

The ubiquitous continuous model used to describe two-phase flows in
soils is based on an extension of the Darcy equation (creeping saturated
single-phase flow in porous media). The whole model, also known as
Muskat equations
{[}\protect\hyperlink{ref-wyckoff1936flow}{8},\protect\hyperlink{ref-muskat1938flow}{9}{]},
reads

\[
0=\frac{\partial\varepsilon S_{i}}{\partial t}+\nabla\cdot\mathbf{U}_{i},\quad i=o,w,
\] \{\#eq:darcyContinuity\}

\[
\mathbf{U}_{i}=-\frac{1}{\mu_{i}}\mathbf{K}_{i}\cdot(\nabla P_{i}-\rho_{i}\mathbf{g}),\quad i=o,w,
\] \{\#eq:darcyMomentum\}

\[
1=S_{w}+S_{o}
\] \{\#eq:darcySaturation\}

\[
\mathbf{K}_{w}=\mathbf{K}k_{rw}(S_{w}),\qquad\mathbf{K}_{o}=\mathbf{K}k_{ro}(S_{w}),
\] \{\#eq:darcyConstitutiveRel\}

\[
P_{c}(S_{w})=P_{o}-P_{w}.
\] \{\#eq:darcyPc\}

In these equations, the subscripts \(o\) and \(w\) refer to the
non-wetting and wetting phase, respectively. \(S_{i}\) is the saturation
of phase \(i\), \(\varepsilon\) is the medium porosity,
\(\mathbf{U}_{i}\) is the superficial mean velocity of fluid \(i\),
\(P_i\) is the intrinsic mean pressure of fluid \(i\) and \(\mathbf{K}\)
is the absolute permeability tensor. The generalization toward two-phase
flows involves the introduction of the relative permeability terms
\(k_{ri}\) which account for one phase for the inaccessible void space
occupied by the other fluid
{[}\protect\hyperlink{ref-dullien2012porous}{5}{]}. To close the set of
the macroscopic equations, the relative permeabilities are assumed to be
non-linear functions of the saturation and a relation between the
macroscopic pressure of each fluid has to be furnished. This relation,
is known as the capillary pressure relation and, as for the relative
permeabilities, is supposed to vary non-linearly with the saturation
only {[}\protect\hyperlink{ref-leverett1941capillary}{10}{]}.

A significant amount of work attempted to make improvements to the
Muskat equations (e.g.~including moving contact line
{[}\protect\hyperlink{ref-kalaydjian1992dynamic}{11}--\protect\hyperlink{ref-barenblatt2003mathematical}{13}{]}
or take into account the trapped phases
{[}\protect\hyperlink{ref-hilfer1998macroscopic}{14}{]}). Here we would
stress that the concept of relative permeabilities into the Muskat
equations is dedicated to independent flow pathways only
{[}\protect\hyperlink{ref-blunt2017multiphase}{15}{]} and the underlying
assumption that the two streams do not interfere with each other.
However, in case that the area of the fluid-fluid interface is not
negligible compared to the solid-fluid interface, as for flow regimes
\cref{fig:fluidPatterns} (b) and \cref{fig:fluidPatterns} (c), one would
include supplementary terms dedicated to the momentum exchange across
the fluid-fluid interface due to normal and shear stresses exerted on it
(i.e.~coupling between the phases) into the macroscopic momentum
equations. In fact, it has been theoretically established that the
macroscopic equations of motion obtained by averaging of the Stokes
equations contain momentum exchange terms between fluids
{[}\protect\hyperlink{ref-MARLE1982643}{16}--\protect\hyperlink{ref-kalaydjian1990origin}{19}{]}.
The volume averaged momentum equation for the \(i\)-phase read

\[
0=-\varepsilon_{i}\nabla P_{i}+\varepsilon_{i}\rho_{i}\mathbf{g}+\frac{1}{V}\int_{A_{is}+A_{ij}}\mathbf{n}_{i}\cdot\left(-\mathbf{I}p_{i}+\mu_{i}\left(\nabla\mathbf{u}_{i}+\left(\nabla\mathbf{u}_{i}\right)^{T}\right)\right)\mathrm{d}A.
\] \{\#eq:VAmomentum\}

The next step in the upscaling methods literature is to express the
boundaries integral of exchange between the phases in terms of averaged
quantity through the resolution of boundary value problems on a
representative periodic unit-cell. We need to go no further here but we
insist that, as shown in {[}\protect\hyperlink{ref-Lasseux1996}{20}{]},
once the closure problems have been solved, the macroscopic equations of
motion can be put into a form analogous to the generalized Darcy's law,

\[
\mathbf{U}_{i}=-\frac{1}{\mu_{i}}\mathbf{K}_{ii}^{*}\cdot(\nabla P_{i}-\rho_{i}\mathbf{g})-\frac{1}{\mu_{j}}\mathbf{K}_{ij}^{*}\cdot(\nabla P_{j}-\rho_{j}\mathbf{g}),\quad i,j=o,w\:\mathrm{and}\:i\neq j,
\] \{\#eq:darcyCross\}

in which \(\mathbf{K}_{ij}^{*}\) are the coupled relative permeability
tensors.

Many studies have attempted to calculate coupled permeabilities
experimentally based on the +\textbf{???}, but before we turn to present
results from the literature, we discuss the chemical engineering point
of view on the continuous model.

\hypertarget{chemical-engineering-approach}{%
\subsubsection{Chemical engineering
approach}\label{chemical-engineering-approach}}

CFD modeling applied to the hydrodynamics of trickle-bed reactors (TBR)
is in most cases focused on an Euler-Euler flow formulation. The two
fluids are thus treated as two interpenetrating continua. The volume
averaged momentum equation for the two fluids to solve, taking into
account that the inertial effects are not negligible here, read

\[
\rho_{i}\varepsilon_{i}\left(\frac{\partial\mathbf{U}_{i}}{\partial t}+\mathbf{U}_{i}\cdot\nabla\mathbf{U}_{i}\right)=-\varepsilon_{i}\nabla P_{i}+\varepsilon_{i}\rho_{i}\mathbf{g}+\mathbf{F}_{mi}+\varepsilon_{i}\nabla\cdot\boldsymbol{\tau}_{i}+\nabla\cdot\mathbf{R}_{i},
\] \{\#eq:trickleMomentum\}

where \(\varepsilon_{i}\) is the volume fraction of phase \(i\), the
velocity and the pressure are averaged quantities identical to the
generalized Darcy equations, \(\boldsymbol{\tau}_{i}\) denotes the
viscous stress tensor and \(\mathbf{R}_{i}\) is the pseudo-turbulence
stress. The vectors \(\mathbf{F}_{mi}\) are the momentum transfer terms
at the interface for the \(i\)-phase. These terms have to be expressed
in terms of constitutive relation to close the set of equations, and the
main difference between the existing continuous models is the choice of
correlations used for these terms. We stress here that the two equations
+\textbf{???} and +\textbf{???} are equivalents (except for the
non-linear inertial terms).

\hypertarget{modeling-of-the-momentum-exchange-between-fluids}{%
\subsubsection{Modeling of the momentum exchange between
fluids}\label{modeling-of-the-momentum-exchange-between-fluids}}

Here we present different estimation of the momentum transfer terms
(i.e.~coupled relative permeabilities in the porous media framework)
between fluids. Based on the generalized Darcy law with cross terms
+\textbf{???}, experimental works computed the coupled transport
coefficient through steady-state cocurent flow in sandpack with
alternatively one fluid, then the other, which is submitted to a null
pressure gradient. This protocol was used with oil and water in a
cylindric sandpack
{[}\protect\hyperlink{ref-zarcone1994determination}{21}{]} and the
authors found a negligible effect of the coupled permeabilities in the
overall flow. With the same protocol, with oil and water in a
2D-sandpack Dullien and Dong
{[}\protect\hyperlink{ref-Dullien1996}{22}{]} found that the coupled
permeabilities are important since they can contribute at best to 35\%
of the effective permeability. Alternatively, authors imposed
{[}\protect\hyperlink{ref-ramakrishnan2015measurement}{23}{]} a null
displacement and measured the induced pressure drop, with air and water
in a Berea sandstone core they found that the coupled transport
coefficients must not be overlooked for intermediate saturations.
Several authors conducted experiments with two different set-ups as
proposed by Rose {[}\protect\hyperlink{ref-Rose1988}{24}{]}, e.g.
{[}\protect\hyperlink{ref-bentsen1993use}{25}{]}, in which the authors
made cocurrent and countercurrent experiments with water and oil in a
sandpack and found that coupled permeabilities reach, at least, 15\% of
the effective permeability value. However, it was pointed out that the
saturation between the two sets of experiments can be very different and
therefore the computed relative permeabilities can not be safely
compared {[}\protect\hyperlink{ref-langaas2001numerical}{26}{]}. The
effect of the non-wetting phase connectivity on the transport parameters
was extensively studied in {[}\protect\hyperlink{ref-Avraam1995}{27}{]},
the authors performed steady-state cocurrent two-phase flow in 2D-micro
model experiments and found that the contribution of the coupled
permeabilities on the flow is non-negligible and depend on the flow
regimes.

Several numerical studies investigate the share of the coupled
permeabilities in the effective permeability. The seminal work of
Rothman {[}\protect\hyperlink{ref-Rothman1990}{28}{]} examined the
question by conducting two-phase flow simulations in simple geometries
with the immiscible lattice-gas method. The author found non-negligible
participation of the coupled permeabilities by applying the volume force
alternatively on each fluid. Numerous authors used the lattice-Boltzmann
method, such as {[}\protect\hyperlink{ref-Li2005}{29}{]}, in which the
authors examined the value of the coupled permeabilities as a function
of the saturation in a 3D-sphere pack and they found results in
agreement with Rothman's results. In
{[}\protect\hyperlink{ref-Yiotis2007}{30}{]} the authors used a lattice
Boltzmann method in 2D and 3D pore networks and found a non-wetting
apparent relative permeability greater than unity when the wetting fluid
is more viscous than the non-wetting fluid. This result is a
manifestation of the lubrification effect, also observed in experiments
{[}\protect\hyperlink{ref-odeh1959effect}{31}{]}, and arising because of
a strongly hydraulic coupling between the fluids. Recently, in
{[}\protect\hyperlink{ref-shams2018study}{32}{]}, the authors have used
a Volume Of Fluid method to study the transport coefficients of fluid
layers in non-circular capillary tubes. Based on analogy with a model
Couette flow the authors have derived simple relations that can predict
with good accuracy the transport coefficients of fluid layers.

\hypertarget{outline-of-the-study}{%
\subsection{Outline of the study}\label{outline-of-the-study}}

In this study, we investigate the effect of the varying intrinsic
permeability and capillary number on the momentum transfer terms through
direct numerical simulations of two-phase flow in a model
high-permeability 2D-porous media. To do so we compute the surface
integrals in +\textbf{???} on each boundaries of the system. The idea of
solving a depth-averaged flow in a Hele-Shaw allows us to introduce an
additional permeability term and thus to vary the intrinsic permeability
of the strucure without having to modify the geometry per se.

\hypertarget{theoretical-background}{%
\section{Theoretical background}\label{theoretical-background}}

\hypertarget{macroscopic-quantities}{%
\subsection{Macroscopic quantities}\label{macroscopic-quantities}}

In the following, the superficial mean of a physical quantity
\(\theta_i\) associated with the \(i\)-phase is given by

\[
\left\langle \theta_{i}\right\rangle =\frac{1}{V}\int_{V_{i}}\theta_{i}\:\mathrm{d}V.
\] \{\#eq:superficialMean\}

Macroscopic quantities such as saturation and intrinsic mean pressure
for two fluids \(i=w,o\) read

\[
S_i = \frac{1}{V} \int_{V}\chi_{i}\;\mathrm{d}V
\] \{\#eq:saturation\}

and

\[
P_{i}=\frac{\int_{V}p\chi_{i}\;\mathrm{d}V}{\int_{V}\chi_{i}\;\mathrm{d}V},
\] \{\#eq:intrinsicMeanPressure\}

respectively. In these equations \(V\) is the volume accessible to the
fluids, that is the pore volume, and \(\chi_i\) is the phase indicator
of the i-fluid (scalar function that takes the value 1 in the fluid and
0 elsewhere). The quantity \(\int_{V}\chi_{i}\;\mathrm{d}V\) is the
volume fraction of the i-fluid that is \(\varepsilon_{i}\) as already
introduced in the previous section.

Then we can introduce the dynamic capillary pressure defined as the
difference of the intrinsic mean pressures,

\[
P_{c}=P_{o}-P_{w}.
\] \{\#eq:darcyPc\}

\hypertarget{traction-terms}{%
\subsection{Traction terms}\label{traction-terms}}

As previously said, an essential step in order to derive some
macroscopic law at macroscale from the microscale is to use averaged
quantities to express the momentum transfer terms between phases that
arising because of the averaging theorem application
{[}\protect\hyperlink{ref-Whitaker2013}{33}{]}. Remember that these
closure relations are not presented here but rather we leave intact the
surface integrals, the averaged momentum balance equations +\textbf{???}
for two fluids \(w\) and \(o\), are recast as

\[
0=-\varepsilon_{w}\nabla P_{w}+\mathbf{T}_{wo}+\mathbf{T}_{ws}, 
\]

\[
0=-\varepsilon_{o}\nabla P_{o}+\mathbf{T}_{ow}+\mathbf{T}_{os},
\]

where \(\varepsilon_{i}\) and \(P_{i}\) are the macroscopic quantities
defined above, \(\mathbf{T}_{ij}\) the traction (name given to surface
forces in the Boundary Integral Methods literature in which these
integrals are central) exerted by the \(i\)-fluid on the \(j\)-fluid and
\(\mathbf{T}_{is}\) the traction exerted by the \(i\)-fluid on the solid
surfaces {[}\protect\hyperlink{ref-Kalaydjian1987}{34}{]}. The traction
terms are written as

\[
\mathbf{T}_{ij}=\frac{1}{V}\int_{A_{ij}}\mathbf{n}_{ij}\cdot\left(-p_{i}\mathbf{I}+2\mu_{i}\mathbf{e}_{i}\right)\mathrm{d}A,
\]

where \(\mathbf{n}_{ij}\) is the normal vector at the interface and
pointing toward the j-phase, \(p_{i}\) the pointwise pressure of the
i-fluid and
\(\mathbf{e}_{i}=\frac{1}{2}\left(\nabla\mathbf{u}_{i}+\left(\nabla\mathbf{u}_{i}\right)^{T}\right)\)
the rate-of-strain tensor for Newtonian fluids with \(\mathbf{u}_{i}\)
the pointwise velocity of the i-fluid.

\hypertarget{theory-of-two-phase-displacement-in-hele-shaw-cell}{%
\subsection{Theory of two-phase displacement in Hele-Shaw
cell}\label{theory-of-two-phase-displacement-in-hele-shaw-cell}}

Two-phase flow between two parallel plates under study here is depicted
in \cref{fig:modelStudy}. This illustration represents a cocurrent flow
of two fluids inside a Hele-Shaw cell parallel to the \(x-y\) plane,
with the \(z\)-axis perpendicular to the plates, and for which \(h\) is
the aperture between the plates. A solid obstacle of circular
cross-section is sandwiched between the two plates. Tangential (cell
plane) stress, as well as stress arising from the perpendicular
confinement, need to be taking into account to properly describe the
flow. The important parameter here are the aspect ratio \(h/l\) where
\(l\) is some characteristic transverse length (hereafter \(l\) is the
cell width as shown in \cref{fig:modelStudy}) and the capillary number.

\begin{figure}
\hypertarget{fig:modelStudy}{%
\centering
\includegraphics{figures/pdf/dessin_courbure.pdf}
\caption{Illustation of confined cocurrent two-phase flow between two
parallel plates and around cylinder obstacles.}\label{fig:modelStudy}
}
\end{figure}

Beside the interface curvature in the Hele-Shaw cell plane, the meniscus
in the perpendicular plane must be taken into account. Therefore, the
pressure jump at equilibrium across the interface reads

\[
\Delta p = \gamma \left(\frac{2}{h}+\kappa_{\parallel}\right),
\] \{\#eq:dP\}

where \(\gamma\) denotes the surface tension and \(\kappa_{\parallel}\)
the curvature in the Hele-Shaw plane
{[}\protect\hyperlink{ref-saffman1958penetration}{35}{]}. The
+\textbf{???} was completed in
{[}\protect\hyperlink{ref-park1984two}{36}{]} by an additional term that
pertains for dynamic film formation and a corrective \(\pi/4\) factor
for the curvature in the cell plane.

To compute the traction exerted by the fluids on the cell plates we need
to consider that the velocity profile in the \(z\)-direction is well
described by the Poiseuille flow profile,

\[
u=-\frac{1}{2\mu}\frac{dp}{dx}z(z-h).
\] \{\#eq:poiseuilleProfile\}

Then the friction force exerted on the bottom plate is given by

\[
\mu\frac{\partial u}{\partial z}\mid_{z=0}=-\frac{h}{2}\frac{dp}{dx}.
\] \{\#eq:poiseuilleFriction\}

Thus, the traction \(T_{i,pl}^x\) is obtained by multiplying the
friction force by two times the surface wetted by the fluid \(i\).

We can show that the average velocity taken over the interval between
the plates is written as follows

\[
\mathbf{U}(x,y,0)=-\frac{h^{2}}{12\mu}\nabla_{\parallel}p.
\] \{\#eq:darcyHeleShaw\}

There is therefore a direct analogy between flows in a 2D porous medium
and a flow in a Hele-Shaw cell, as can be seen in +\textbf{???}, where
the permeability is given by \(h^{2}/12\)
{[}\protect\hyperlink{ref-saffman1958penetration}{35}{]}.

\hypertarget{direct-numerical-simulations}{%
\section{Direct numerical
simulations}\label{direct-numerical-simulations}}

\hypertarget{equations}{%
\subsection{Equations}\label{equations}}

\hypertarget{level-set-model}{%
\subsubsection{Level Set model}\label{level-set-model}}

We use the Level Set method to capture the moving free interface between
the fluids. In this framework the fluid phases are identified with a
level set (scalar) function that goes smoothly from 0 to 1 across the
fluid interface and the interface is implicitely defined as the
iso-level \(\phi=0.5\). The transport of the level set function \(\phi\)
is governed by

\[
\frac{\partial\phi}{\partial t}+\nabla\cdot(\mathbf{u}\phi)=\tau\nabla\cdot\left(\psi\boldsymbol{\nabla}\phi-\phi(1-\phi)\frac{\boldsymbol{\nabla}\phi}{\vert\nabla\phi\vert}\right).
\] \{\#eq:advecphi\}

In this equation, \(\mathbf{u}\) is the velocity field, \(\tau\) and
\(\psi\) are two numerical parameters that control the diffuse interface
thickness and the amount of reinitialization of the \(\phi\) function,
respectively
{[}\protect\hyperlink{ref-Olsson2005}{37},\protect\hyperlink{ref-Olsson2007}{38}{]}.

\hypertarget{flow-model}{%
\subsubsection{Flow model}\label{flow-model}}

Remember that we need to take into account tangential stress as well
perpendicular stress due to the flow confinement, a full-3D resolution
of the flow equations would be meaningful. In order to avoid solving
such resource-intensive simulations we rather resolved 2D flow equations
plus an additional term taking into account the perpendicular stress.
The velocity and pressure fields are obtained by solving the 2D-modified
Stokes equations since the Reynolds number is assumed to be sufficiently
low so that the flow is creeping, the whole-domain formulation Stokes
and continuity equations read

\[
0=-\nabla p+\mu(\phi)\left(\nabla^{2}\mathbf{u}-k^{2}\mathbf{u}\right)+\mathbf{F}_{st},
\] \{\#eq:StokesMomentum\}

\[
0=\nabla\cdot\mathbf{u},
\] \{\#eq:StokesContinuity\}

where \(k=\frac{\sqrt{12}}{h}\) is an additional term that pertain to a
Darcy-term arising from the depth-averaging of the Stokes equation, as
described in the theoretical background section and \(\mathbf{F}_{st}\)
is the capillary contribution. The capillary contribution is difficult
to compute in volumetric methods such as Level Set and is itself the
object of a large body of work
{[}\protect\hyperlink{ref-Popinet2018}{39}{]}, this term is defined here
as

\[
\mathbf{F}_{st}=\gamma\kappa\delta(\phi)\mathbf{n},
\] \{\#eq:Fst\}

where \(\gamma\) is the surface tension, \(\kappa\) is the interface
mean curvature, \(\mathbf{n}\) denotes the unit normal to the interface
and \(\delta(\phi)\) is the Dirac delta function localized on the
interface {[}\protect\hyperlink{ref-Galusinski2008}{40}{]}. In this
framework the unit normal vector as well as the mean curvature are
obtained with the level set function \(\phi\) as

\[
\mathbf{n}=\frac{\boldsymbol{\nabla}\phi}{\vert\boldsymbol{\nabla}\phi\vert},
\]

\[
\kappa=\frac{\pi}{4}\nabla\cdot\left(\frac{\boldsymbol{\nabla}\phi}{\vert\boldsymbol{\nabla}\phi\vert}\right)+\kappa_{\perp},
\]

respectively, where the \(\pi/4\) correction was derived in
{[}\protect\hyperlink{ref-park1984two}{36}{]} and
\(\kappa_{\perp}=\frac{2}{h}\) is the out-of-plane curvature as
described previously in the theoretical background section.

The material parameters, that is the fluid density and the fluid
viscosity, varies smoothly from one fluid to another

\[
\rho(\phi)=\rho_{o}+(\rho_{w}-\rho_{o})\phi,
\]

\[
\mu(\phi)=\mu_{o}+(\mu_{w}-\mu_{o})\phi.
\]

\hypertarget{dimensionless-equations-and-relevant-dimensionless-numbers}{%
\subsubsection{Dimensionless equations and relevant dimensionless
numbers}\label{dimensionless-equations-and-relevant-dimensionless-numbers}}

If one define the following dimensionless variables

\[
\mathbf{u}=\mathbf{u}'\times U,\;p=p'\times\frac{\mu U}{l},\;\kappa=\kappa'\times l,\;\mathbf{x}=\mathbf{x}'\times l,
\]

where \(U\) is a reference velocity, \(\mu\) a reference viscosity and
\(l\) a reference length, Stokes and continuity equations can be recast
as

\[
0=-\nabla'p'+\frac{\mu(\phi)}{\mu}\nabla'^{2}\mathbf{u}'+\frac{\gamma}{\mu U}\kappa'\nabla'\phi,
\]

\[
0=\nabla'\cdot\mathbf{u}'.
\]

One can notice the two following dimensionless numbers

\[
\bar{\mu}=\frac{\mu(\phi)}{\mu},\;Ca^{-1}=\frac{\gamma}{\mu U},
\]

the viscosity ratio and the capillary number, respectively, which add up
to the already identified aspect ratio \(h/l\).

\hypertarget{test-case}{%
\subsection{Test case}\label{test-case}}

\hypertarget{geometry-and-boundary-conditions}{%
\subsubsection{Geometry and boundary
conditions}\label{geometry-and-boundary-conditions}}

We study a model system consisting of an array of cylinders in a thin
square cuboid, resembling a Hele-Shaw-cell with obstacles, as shown in
\cref{fig:model}. The two fluids are injected from the left at a
constant normal velocity. The outlet boundary condition for the flow is
a constant pressure. Initially the wetting fluid occupy the entire model
as depicted in \cref{fig:model}. The different boundary conditions are
listed in \cref{tbl:BC}.

\begin{figure}
\hypertarget{fig:model}{%
\centering
\includegraphics{figures/pdf/model.pdf}
\caption{The geometry used is an array of 5 cylinders in a thin square
cuboid, resembling a Hele-Shaw-cell where both fluids are injecting from
left to right, initially the square cuboïd are saturated with wetting
fluid (in blue), the length of one Unit-Cell (UC) is one
millimetre.}\label{fig:model}
}
\end{figure}

\begin{longtable}[]{@{}llll@{}}
\caption{Boundary conditions used. \label{tbl:BC}}\tabularnewline
\toprule
Boundary & u & p & \(\phi\)\tabularnewline
\midrule
\endfirsthead
\toprule
Boundary & u & p & \(\phi\)\tabularnewline
\midrule
\endhead
\(A_{ws}\) & 0 & - & \(\mathbf{n}\cdot\mathbf{u}\phi=0\)\tabularnewline
Outlet & - & 0 &
\(\mathbf{n}\cdot\boldsymbol{\nabla}\phi=0\)\tabularnewline
Inlet o & \(U_{o}^{x}=\mathrm{cst}\) & - & 0\tabularnewline
Inlet w & \(U_{w}^{x}=\mathrm{cst}\) & - & 1\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{validation}{%
\subsubsection{Validation}\label{validation}}

\hypertarget{results}{%
\section{Results}\label{results}}

The results are given as a function of the dimensionless numbers defined
previously. The reference length \(l\) is the width of the unit cells
(0.5 mm), the reference viscosity is the invading fluid viscosity that
is two times greater (\(\mu_o=2\times 10^{-3}\) Pa.s) than the wetting
fluid viscosity. and the reference velocity is the total velocity, that
is \(U=U_{o}+U_{w}\). In consequence the capillary number and the
dimensionless inlet velocity read

\[
Ca=\frac{(U_{o}+U_{w})\mu_{o}}{\gamma},\;U_{w}^{'}=\frac{U_{w}}{U_{o}+U_{w}}=f_{f},\;U_{o}^{'}=1-f_{f},
\]

respectively, where \(f_{f}\) is the fractional flow. In the following
the viscosity ratio \(\bar{\mu}\) as well as the fractional flow
\(f_{f}\) are kept constant at 2 and 0.25, respectively.

In the following, the intrinsic permeability of the model is varied by
modifying the aperture between the two plates of the cell. We use five
different values for the aperture and the correspondence in terms of
intrinsic permeability is given in \cref{tbl:permeability} from
single-phase simulation. When \(h\rightarrow\infty\), it is expected
that the intrinsic permeability tends towards the permeability obtained
with purely 2D flow equations, this can be verified in
\cref{fig:permeability}. We obtain intrinsic permeability values typical
of high permeability porous media with the selected aperture values.

\begin{longtable}[]{@{}lll@{}}
\caption{Equivalence between the value of the cell plates aperture and
the intrinsic permeability of the model obtained either from the
intrinsic permeability for flow between two plates or from single-phase
simulation in our model. \label{tbl:permeability}}\tabularnewline
\toprule
\(l/h\) & \(h^2/12\) (darcy) & \(k\) Simulation (darcy)\tabularnewline
\midrule
\endfirsthead
\toprule
\(l/h\) & \(h^2/12\) (darcy) & \(k\) Simulation (darcy)\tabularnewline
\midrule
\endhead
0.2 & \(5.21 \times 10^{5}\) & \(1.48 \times 10^{4}\)\tabularnewline
2.0 & \(5.21 \times 10^{3}\) & \(2.85 \times 10^{3}\)\tabularnewline
4.0 & \(1.30 \times 10^{3}\) & \(8.63 \times 10^{2}\)\tabularnewline
8.0 & \(3.26 \times 10^{2}\) & \(2.35 \times 10^{2}\)\tabularnewline
20 & \(5.21 \times 10^{1}\) & \(3.94 \times 10^{1}\)\tabularnewline
\bottomrule
\end{longtable}

\begin{figure}
\hypertarget{fig:permeability}{%
\centering
\includegraphics{figures/pdf/permeability.pdf}
\caption{Intrinsic permeability of the Hele-Shaw cell normalized by the
intrinsic permeability obtained from pure 2D
simulation.}\label{fig:permeability}
}
\end{figure}

\hypertarget{wetting-fluid-saturation-decreases-when-permeability-is-decreasing}{%
\subsection{Wetting fluid saturation decreases when permeability is
decreasing}\label{wetting-fluid-saturation-decreases-when-permeability-is-decreasing}}

Before we turn to the study of traction terms it is worth invesigate the
effect of varying permeability on the fluid repartition and the velocity
fields. Wetting fluid saturation in the third unit-cell (UC3), computed
accordingly to +\textbf{???}, is given as a function of the
dimensionless aperture between the plates in \cref{fig:saturation}.

The wetting fluid saturation non-linearly decreases with \(h^*\) and
stays at moderate values, between 0.6 and 0.3. A smaller capillary
number tends to slightly increase the wetting fluid saturation at
steady-state.

\begin{figure}
\hypertarget{fig:saturation}{%
\centering
\includegraphics{figures/pdf/saturation.pdf}
\caption{Saturation in wetting fluid as a function of the dimensionless
aperture \(h^{*}=h/l\) and for different values of the capillary number
(UC3) at steady-state.}\label{fig:saturation}
}
\end{figure}

Let us take a closer look at the interface position for different values
of \(h^*\) for \(Ca=1\) and \(Ca=0.05\) in \cref{fig:interfaceCa1} and
\cref{fig:interfaceCa005}, respectively. The interface for small
capillary number and large aperture is significantly flatter than for
high capillary number and/or small aperture. One also noted that the
interface is symmetrical when the capillary number is large whereas it
loses this property for a smaller capillary number and for larger
aperture between the plates. On the same plots we show the position of
the interface in a purely 2D case, and one can see that it is very close
to the solution for \(1/h^*=0.2\). The fluid-fluid interface moves
closer to the cylinder for decreasing aperture (i.e.~intrinsic
permeability), this has an impact on the velocity fields, as shown in
\cref{fig:UvelocityCa1LH2} and \cref{fig:UvelocityCa1LH20}, which
represent the dimensionless \(x\)-component of the velocity for
\(Ca=1.0\) and \(1/h^*=2\) and \(1/h^*=20\), respectively. Clearly, what
is happening here is that the velocity profile is close to the
Poiseuille profile as long as the interface is not too close from the
solid. To verify this, we plot in \cref{fig:uprofileCa1} the
dimensionless \(x\)-component of the velocity along the line shown in
\cref{fig:UvelocityCa1LH20}. The profile for the pure 2D solution
(\(h\rightarrow\infty\)) resembles a Poiseuille profile while for
decreasing aperture the maximum velocity is observed between the solid
and the fluid-fluid interface.

\begin{figure}
\hypertarget{fig:interfaceCa1}{%
\centering
\includegraphics{figures/pdf/interfaceCa1.pdf}
\caption{Interface position at steady-state in UC3 for
\(Ca=5\times10^{-2}\).}\label{fig:interfaceCa1}
}
\end{figure}

\begin{figure}
\hypertarget{fig:interfaceCa005}{%
\centering
\includegraphics{figures/pdf/interfaceCa005.pdf}
\caption{Interface position at steady-state in UC3 for
\(Ca=5\times10^{-2}\).}\label{fig:interfaceCa005}
}
\end{figure}

\begin{figure}
\hypertarget{fig:UvelocityCa1LH2}{%
\centering
\includegraphics{figures/png/Uvelocity_Ca1LH2.png}
\caption{Dimensionless \(x\)-component of the velocity in UC3 for
\(Ca=1\) and \(1/h^*=2\).}\label{fig:UvelocityCa1LH2}
}
\end{figure}

\begin{figure}
\hypertarget{fig:UvelocityCa1LH20}{%
\centering
\includegraphics{figures/png/Uvelocity_Ca1LH20.png}
\caption{Dimensionless \(x\)-component of the velocity in UC3 for
\(Ca=1\) and \(1/h^*=20\).}\label{fig:UvelocityCa1LH20}
}
\end{figure}

\begin{figure}
\hypertarget{fig:uprofileCa1}{%
\centering
\includegraphics{figures/pdf/uprofileCa1.pdf}
\caption{Dimensionless \(x\)-velocity profile in UC3 for \(Ca=1\) and
different values of the intrinsic permeability.}\label{fig:uprofileCa1}
}
\end{figure}

\hypertarget{dynamic-capillary-pressure-scale-linearly-with-the-gap-between-hele-shaw-plates-regardless-of-the-capillary-number}{%
\subsection{Dynamic capillary pressure scale linearly with the gap
between Hele-Shaw plates regardless of the capillary
number}\label{dynamic-capillary-pressure-scale-linearly-with-the-gap-between-hele-shaw-plates-regardless-of-the-capillary-number}}

We now consider the dynamic capillary pressure as defined in
+\textbf{???} as a function of \(h^*\). As can be seen on \cref{fig:Pc},
\(P_c\) perfectly scales linearly with \(h^*\), regardless of the
capillary number. One can find on the same plot the theoretical pressure
jump across the fluid-fluid interface due to the out-of-plane meniscus.
This pressure jump closely matches the computed dynamic capillary
pressure and to this point, we can conclude that even for the largest
aperture the difference of mean intrinsic pressure is given by the
meniscus between the two plates. Seeing this, it is worthwhile to
contrast the pressure jump across the interface with the pressure
gradient across the cell. We plot the pressure field and pressure
contour in the third unit-cell (UC3) for \(Ca=1\) and \(1/h^*=2\) on
\cref{fig:pFieldCa1} (a) and \(1/h^*=20\) on \cref{fig:pFieldCa1} (b).
One can note that, although the pressure jump is greater in absolute
terms, it is almost unobservable for the smallest aperture because of
the large pressure gradient.

\begin{figure}
\hypertarget{fig:Pc}{%
\centering
\includegraphics{figures/pdf/Pc.pdf}
\caption{Dynamic capillary pressure in UC3 as a function of \(h^*\) for
different capillary numbers and pressure drop across the fluid-fluid
based on the out-of-plane meniscus at steady-state}\label{fig:Pc}
}
\end{figure}

\begin{figure}
\hypertarget{fig:pFieldCa1}{%
\centering
\includegraphics{figures/pdf/pFieldCa1.pdf}
\caption{Pressure field in UC3 for \(Ca=1\) and (a) \(1/h^*=2\) and (b)
\(1/h^*=20\) and interface position as the iso-\(\phi\)=0.5
contour}\label{fig:pFieldCa1}
}
\end{figure}

\begin{figure}
\hypertarget{fig:pFieldCa001}{%
\centering
\includegraphics{figures/pdf/pFieldCa001.pdf}
\caption{Pressure field in UC3 for \(Ca=1 \times 10^{-2}\) and (a)
\(1/h^*=2\) and (b) \(1/h^*=20\) and interface position as the
iso-\(\phi\)=0.5 contour}\label{fig:pFieldCa001}
}
\end{figure}

\hypertarget{fluid-fluid-traction-increases-faster-compared-to-solid-fluid-traction-when-permeability-is-decreasing}{%
\subsection{Fluid-fluid traction increases faster compared to
solid-fluid traction when permeability is
decreasing}\label{fluid-fluid-traction-increases-faster-compared-to-solid-fluid-traction-when-permeability-is-decreasing}}

Here we plot traction terms as a function of the aperture between the
Hele-Shaw plates. We first plot on \cref{fig:drag_plates} traction
exerted on the Hele-Shaw plates by each fluid, remember that this term
is computed as described in the theoretical background section.
Logically, the traction exerted by the non-wetting fluid on the cell
plates becomes more important for smaller aperture since, as discussed
above, the surface wetted by this fluid becomes larger as the aperture
decreases.

\begin{figure}
\hypertarget{fig:drag_plates}{%
\centering
\includegraphics{figures/pdf/drag_plates.pdf}
\caption{\(x\)-component of the traction on Hele-Shaw cell plates by
each fluid in UC3, as a function of the dimensionless gap and for
different capillary numbers at steady-state.}\label{fig:drag_plates}
}
\end{figure}

In \cref{fig:drag_solid} we plot the sum of traction (\(x\)-component)
exerted on all solid-fluid boundaries as a function of the dimensionless
aperture between the plates and for different capillary numbers. For
\(1/h^*\) greater than 2, the traction scale a bit slower than
\(1/h^2\). Logically, the greater the capillary number, the greater the
traction term.

\begin{figure}
\hypertarget{fig:drag_solid}{%
\centering
\includegraphics{figures/pdf/drag_solid.pdf}
\caption{\(x\)-component of the traction exerted on the solid-fluid
boundaries in UC3 as a function of the dimensionless aperture and for
different capillary numbers at steady-state.}\label{fig:drag_solid}
}
\end{figure}

We compute the fluid-fluid traction \(T_{ij}\) in two steps. The first
step concerns the viscous part of the stress tensor, which is calculated
directly from the velocity gradients available on the interface contour.
The second step concerns the pressure part, which is obtained by
applying the divergence theorem on each phase into UC3. In
\cref{fig:drag_fluidint} we plot the traction on the fluid-fluid
interface \(T_{ow}\) as a function of \(h^*\). From \(1/h^*=2\) the
traction scale a bit faster than \(1/h^2\).

\begin{figure}
\hypertarget{fig:drag_fluidint}{%
\centering
\includegraphics{figures/pdf/drag_fluidint.pdf}
\caption{\(x\)-component of the traction exerted on the fluid-fluid
interface in UC3 as a function of the dimensionless aperture and for
different capillary numbers at steady-state.}\label{fig:drag_fluidint}
}
\end{figure}

\hypertarget{decreasing-the-permeability-does-not-imply-that-the-share-of-the-fluid-fluid-traction-is-negligible-in-the-total-traction}{%
\subsection{Decreasing the permeability does not imply that the share of
the fluid-fluid traction is negligible in the total
traction}\label{decreasing-the-permeability-does-not-imply-that-the-share-of-the-fluid-fluid-traction-is-negligible-in-the-total-traction}}

Let us now sum up all the traction terms and answer the question of the
share of traction exerted on the fluid-fluid boundaries in total
traction. The ratio \(T_{int}^x/T_{s}^x\), were \(s={pl+cyl}\) and
\(int={wo+ow}\), is given as a function of the dimensionless aperture
between the Hele-Shaw plates in \cref{fig:ratioDrag}. The traction
exerted on the fluid-fluid boundary reaches between 10 to 50\% of the
traction exerted on the solid-fluid boundaries. Interestingly, the ratio
varies non-monotonically with \(h^*\), the minimum is reached for the
smallest capillary numbers and for a medium range of the aperture
values.

As the capillary number decreases, variations with the aperture are
amplified. Thus, it is for the lowest capillary number tested that we
obtain both the maximum and minimum values of the ratio
\(T_{int}^x/T_{s}^x\).

\begin{figure}
\hypertarget{fig:ratioDrag}{%
\centering
\includegraphics{figures/pdf/ratioDrag.pdf}
\caption{Ratio of traction on fluid-fluid interface over traction on
solid-fluid interfaces in UC3 as a function of the dimensionless
aperture between the Hele-Shaw plates and for different capillary
numbers at steady-state.}\label{fig:ratioDrag}
}
\end{figure}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

In this work, we investigated the effect of varied permeability on
cocurrent two-phase flow in a Hele-Shaw cell. We focused on the value of
the traction terms, that are necessary to close the macroscopic flow
equations. The idea of solving modified 2D-Stokes equations with a
supplementary Darcy term allowed us to modify the intrinsic permeability
of the structure without having to modify the geometry per se. The
chosen range for intrinsic permeability values is typical of
high-permeability porous media, for which it is known that the flow
regimes are characterized by the large extent of the fluid-fluid
interface the fluids (e.g.~film flow). Through direct numerical
simulation and a Level Set method to handle the free interface between
the fluid we have found that

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-fetter2017contaminant}{}%
{[}1{]} Fetter CW, Boving T, Kreamer D. Contaminant hydrogeology.
Waveland Press; 2017.

\leavevmode\hypertarget{ref-clavier2017modeling}{}%
{[}2{]} Clavier R, Chikhi N, Fichot F, Quintard M. Modeling of inertial
multi-phase flows through high permeability porous media: Friction
closure laws. International Journal of Multiphase Flow 2017;91:243--61.

\leavevmode\hypertarget{ref-Santos1991}{}%
{[}3{]} Santos JM de, Melli TR, Scriven LE. Mechanics of gas-liquid flow
in packed-bed contactors. Annual Review of Fluid Mechanics
1991;23:233--60.
\url{https://doi.org/10.1146/annurev.fl.23.010191.001313}.

\leavevmode\hypertarget{ref-davit2018one}{}%
{[}4{]} Davit Y, Quintard M. One-phase and two-phase flow in highly
permeable porous media. Heat Transfer Engineering 2018:1--19.

\leavevmode\hypertarget{ref-dullien2012porous}{}%
{[}5{]} Dullien FAL. Porous media: Fluid transport and pore structure.
Academic press; 2012.

\leavevmode\hypertarget{ref-Avraam1995a}{}%
{[}6{]} Avraam DG, Payatakes AC. Flow regimes and relative
permeabilities during steady-state two-phase flow in porous media.
Journal of Fluid Mechanics 1995;293:207--36.
\url{https://doi.org/10.1017/S0022112095001698}.

\leavevmode\hypertarget{ref-armstrong2016beyond}{}%
{[}7{]} Armstrong RT, McClure JE, Berrill MA, Rücker M, Schlüter S, Berg
S. Beyond darcy's law: The role of phase topology and ganglion dynamics
for two-fluid flow. Physical Review E 2016;94:043113.

\leavevmode\hypertarget{ref-wyckoff1936flow}{}%
{[}8{]} Wyckoff RD, Botset HG. The flow of gas-liquid mixtures through
unconsolidated sands. Physics 1936;7:325--45.

\leavevmode\hypertarget{ref-muskat1938flow}{}%
{[}9{]} Muskat M. The flow of homogeneous fluids through porous media.
Soil Science 1938;46:169.

\leavevmode\hypertarget{ref-leverett1941capillary}{}%
{[}10{]} Leverett M, others. Capillary behavior in porous solids.
Transactions of the AIME 1941;142:152--69.

\leavevmode\hypertarget{ref-kalaydjian1992dynamic}{}%
{[}11{]} Kalaydjian F-M, others. Dynamic capillary pressure curve for
water/oil displacement in porous media: Theory vs. Experiment. In:. SPE
annual technical conference and exhibition, Society of Petroleum
Engineers; 1992.

\leavevmode\hypertarget{ref-Hassanizadeh1993}{}%
{[}12{]} Hassanizadeh SM, Gray WG. Thermodynamic basis of capillary
pressure in porous media. Water Resources Research 1993;29:3389--405.
\url{https://doi.org/10.1029/93WR01495}.

\leavevmode\hypertarget{ref-barenblatt2003mathematical}{}%
{[}13{]} Barenblatt GI, Patzek TW, Silin DB. The mathematical model of
nonequilibrium effects in water-oil displacement. SPE Journal
2003;8:409--16.

\leavevmode\hypertarget{ref-hilfer1998macroscopic}{}%
{[}14{]} Hilfer R. Macroscopic equations of motion for two-phase flow in
porous media. Physical Review E 1998;58:2090.

\leavevmode\hypertarget{ref-blunt2017multiphase}{}%
{[}15{]} Blunt MJ. Multiphase flow in permeable media: A pore-scale
perspective. Cambridge University Press; 2017.

\leavevmode\hypertarget{ref-MARLE1982643}{}%
{[}16{]} Marle CM. On macroscopic equations governing multiphase flow
with diffusion and chemical reactions in porous media. International
Journal of Engineering Science 1982;20:643--62.
\url{https://doi.org/https://doi.org/10.1016/0020-7225(82)90118-5}.

\leavevmode\hypertarget{ref-Whitaker1986a}{}%
{[}17{]} Whitaker S. Flow in porous media II: The governing equations
for immiscible, two-phase flow. Transport in Porous Media
1986;1:105--25.

\leavevmode\hypertarget{ref-auriault1986remarques}{}%
{[}18{]} Auriault J, Sanchez-Palencia E. Remarques sur la loi de darcy
pour les écoulements biphasiques en milieu poreux. Journal of
Theoretical and Applied Mechanics, Numéro Spécial 1986:p141--156.

\leavevmode\hypertarget{ref-kalaydjian1990origin}{}%
{[}19{]} Kalaydjian F. Origin and quantification of coupling between
relative permeabilities for two-phase flows in porous media. Transport
in Porous Media 1990;5:215--29.

\leavevmode\hypertarget{ref-Lasseux1996}{}%
{[}20{]} Lasseux D, Quintard M, Whitaker S. Determination of
permeability tensors for two-phase flow in homogeneous porous media:
Theory. Transport in Porous Media 1996;24:107--37.

\leavevmode\hypertarget{ref-zarcone1994determination}{}%
{[}21{]} Zarcone C, Lenormand R. Détermination expérimentale du couplage
visqueux dans les écoulements diphasiques en milieu poreux. Comptes
Rendus de L'Académie Des Sciences Série II, Mécanique, Physique, Chimie,
Astronomie 1994;318:1429--35.

\leavevmode\hypertarget{ref-Dullien1996}{}%
{[}22{]} Dullien FAL, Dong M. Experimental determination of the flow
transport coefficients in the coupled equations of two-phase flow in
porous media. Transport in Porous Media 1996;25:97--120.
\url{https://doi.org/10.1007/bf00141264}.

\leavevmode\hypertarget{ref-ramakrishnan2015measurement}{}%
{[}23{]} Ramakrishnan TS, Goode PA. Measurement of off-diagonal
transport coefficients in two-phase flow in porous media. Journal of
Colloid and Interface Science 2015;449:392--8.

\leavevmode\hypertarget{ref-Rose1988}{}%
{[}24{]} Rose W. Measuring transport coefficients necessary for the
description of coupled two-phase flow of immiscible fluids in porous
media. Transport in Porous Media 1988;3:163--71.
\url{https://doi.org/10.1007/BF00820343}.

\leavevmode\hypertarget{ref-bentsen1993use}{}%
{[}25{]} Bentsen RG, Manai AA. On the use of conventional cocurrent and
countercurrent effective permeabilities to estimate the four generalized
permeability coefficients which arise in coupled, two-phase flow.
Transport in Porous Media 1993;11:243--62.

\leavevmode\hypertarget{ref-langaas2001numerical}{}%
{[}26{]} Langaas K, Papatzacos P. Numerical investigations of the steady
state relative permeability of a simplified porous medium. Transport in
Porous Media 2001;45:241--66.

\leavevmode\hypertarget{ref-Avraam1995}{}%
{[}27{]} Avraam DG, Payatakes AC. Generalized relative permeability
coefficients during steady-state two-phase flow in porous media, and
correlation with the flow mechanisms. Transport in Porous Media
1995;20:135--68. \url{https://doi.org/10.1007/bf00616928}.

\leavevmode\hypertarget{ref-Rothman1990}{}%
{[}28{]} Rothman DH. Macroscopic laws for immiscible two-phase flow in
porous media: Results from numerical experiments. Journal of Geophysical
Research 1990;95:8663. \url{https://doi.org/10.1029/jb095ib06p08663}.

\leavevmode\hypertarget{ref-Li2005}{}%
{[}29{]} Li H, Pan C, Miller CT. Pore-scale investigation of viscous
coupling effects for two-phase flow in porous media. Physical Review E
2005;72. \url{https://doi.org/10.1103/physreve.72.026705}.

\leavevmode\hypertarget{ref-Yiotis2007}{}%
{[}30{]} Yiotis AG, Psihogios J, Kainourgiakis ME, Papaioannou A, Stubos
AK. A lattice boltzmann study of viscous coupling effects in immiscible
two-phase flow in porous media. Colloids and Surfaces A: Physicochemical
and Engineering Aspects 2007;300:35--49.
\url{https://doi.org/https://doi.org/10.1016/j.colsurfa.2006.12.045}.

\leavevmode\hypertarget{ref-odeh1959effect}{}%
{[}31{]} Odeh AS, others. Effect of viscosity ratio on relative
permeability (includes associated paper 1496-g) 1959.

\leavevmode\hypertarget{ref-shams2018study}{}%
{[}32{]} Shams M, Raeini AQ, Blunt MJ, Bijeljic B. A study to
investigate viscous coupling effects on the hydraulic conductance of
fluid layers in two-phase flow at the pore level. Journal of Colloid and
Interface Science 2018;522:299--310.

\leavevmode\hypertarget{ref-Whitaker2013}{}%
{[}33{]} Whitaker S. The method of volume averaging. vol. 13. Springer
Science \& Business Media; 2013.

\leavevmode\hypertarget{ref-Kalaydjian1987}{}%
{[}34{]} Kalaydjian F. A macroscopic description of multiphase flow in
porous media involving spacetime evolution of fluid/fluid interface.
Transport in Porous Media 1987;2:537--52.
\url{https://doi.org/10.1007/BF00192154}.

\leavevmode\hypertarget{ref-saffman1958penetration}{}%
{[}35{]} Saffman PG, Taylor GI. The penetration of a fluid into a porous
medium or hele-shaw cell containing a more viscous liquid. Proceedings
of the Royal Society of London Series A Mathematical and Physical
Sciences 1958;245:312--29.

\leavevmode\hypertarget{ref-park1984two}{}%
{[}36{]} Park C-W, Homsy G. Two-phase displacement in hele shaw cells:
Theory. Journal of Fluid Mechanics 1984;139:291--308.

\leavevmode\hypertarget{ref-Olsson2005}{}%
{[}37{]} Olsson E, Kreiss G. A conservative level set method for two
phase flow. Journal of Computational Physics 2005;210:225--46.
\url{https://doi.org/https://doi.org/10.1016/j.jcp.2005.04.007}.

\leavevmode\hypertarget{ref-Olsson2007}{}%
{[}38{]} Olsson E, Kreiss G, Zahedi S. A conservative level set method
for two phase flow ii. Journal of Computational Physics
2007;225:785--807.
\url{https://doi.org/https://doi.org/10.1016/j.jcp.2006.12.027}.

\leavevmode\hypertarget{ref-Popinet2018}{}%
{[}39{]} Popinet S. Numerical models of surface tension. Annual Review
of Fluid Mechanics 2018;50:49--75.
\url{https://doi.org/10.1146/annurev-fluid-122316-045034}.

\leavevmode\hypertarget{ref-Galusinski2008}{}%
{[}40{]} Galusinski C, Vigneaux P. On stability condition for bifluid
flows with surface tension: Application to microfluidics. Journal of
Computational Physics 2008;227:6140--64.
\url{https://doi.org/https://doi.org/10.1016/j.jcp.2008.02.023}.

\end{document}

### 2020-03-30 draft.tex
minor

@@
 above) results in a complex interaction between capillary, gravity and
-viscous forces {[}\protect\hyperlink{ref-davit2018one}{4}{]}. Because of
-this complex interplay, the characteristic flow regimes are very
-different from those observed for surface tension dominated flow
-{[}\protect\hyperlink{ref-dullien2012porous}{5}{]}. For low-permeability
-medium, if the surface tension between the fluids is not too low and the
-viscosity not too high, that is the capillary number is low (usually
-inferior to \(10^{-3}\)), the fluid repartition pattern is described as
-two independent flow streams separate by a multitude of stable meniscus
-at steady state, as illustrated in \cref{fig:fluidPatterns} (a). In this
-case, for a completely wet medium, the wetting phase flows in the small
-pores while the non-wetting phase tends to occupy the larger pores. For
-high permeability porous media, the fluid velocity can be high enough
-that the viscous forces dominate over capillary forces (alike gravity
-and inertial effects may become important if Bond and Reynolds numbers
-are high, respectively). Then, the fluids repartition patterns can take
-two forms, either the non-wetting phase is continuous, see for example
-\cref{fig:fluidPatterns} (b), or is flowing as droplets or ganglia as in
-\cref{fig:fluidPatterns} (c), while the wetting phase is flowing as a
-film in contact with the solid. In both cases, the two fluids occupy
-most of the pores at the same time and the non-wetting phase flows at
-the center of the pores surrounded by the wetting phase. One can note
-that these flow regimes are equivalent to those observed in two-phase
-flows in a tube (slug flow, bubbly flow). Strictly speaking, these
-different regimes must be considered when attempting to describe
-two-phase flows with continuous macroscopic equations. Indeed, it has
-been shown on numerous works that the overall flow depends on the flow
-regimes
-{[}\protect\hyperlink{ref-Avraam1995a}{6},\protect\hyperlink{ref-armstrong2016beyond}{7}{]}.
-In the first glance, one would consider that the exchange terms between
-the fluid phases as negligible compared to their counterpart between the
-fluid phases and the solid phase for surface-tension dominated flwo and
-that the extent of the fluid-fluid interface is small, as in
-\cref{fig:fluidPatterns} (a). On the other hand, and this is what we are
+viscous forces {[}\protect\hyperlink{ref-davit2018one}{4}{]}. For
+low-permeability porous media, the flow is usually dominated by surface
+tension force, i.e.~the capillary number is low (usually inferior to
+\(10^{-3}\)), and the fluid repartition pattern is well described as two
+independent flow streams separate by a multitude of stable meniscus at
+steady state, as illustrated in \cref{fig:fluidPatterns} (a)
+{[}\protect\hyperlink{ref-dullien2012porous}{5}{]}. This regime is
+characterized by the small extent of the fluid-fluid interface, as the
+fluid phases are segregated, with the wetting phase flowing into the
+larger pores while the wetting phase occupies the smaller pores. In
+contrast, for high permeability porous media, the flow is no longer
+capillary-dominated and the viscous forces become significant (also
+gravity and inertial effects may become important if Bond and Reynolds
+numbers are high, respectively). Thus the fluids repartition patterns
+can take two forms, either the non-wetting phase is continuous, see as
+an example \cref{fig:fluidPatterns} (b), or is flowing as droplets or
+ganglia, as in \cref{fig:fluidPatterns} (c). For these two options, the
+wetting phase is flowing as a film in contact with the solid and the
+non-wetting phase flows at the center of the pores surrounded by the
+wetting phase. One can note that these flow regimes are analogous to
+those observed in gas-liquid flows in a tube (annular flow, bubbly flow,
+etc.) {[}\protect\hyperlink{ref-taitel1976model}{6}{]}. Strictly
+speaking, these different regimes must be considered when attempting to
+describe two-phase flows with continuous macroscopic equations. Indeed,
+it has been shown several times that the overall flow properties depend
+on the flow regimes
+{[}\protect\hyperlink{ref-Avraam1995a}{7},\protect\hyperlink{ref-armstrong2016beyond}{8}{]}.
+At first glance, one would consider that the exchange terms between the
+fluids, through their common surfaces, as negligible compared to their
+counterpart between the fluid phases and the solid phase for
+surface-tension dominated flow since the extent of the fluid-fluid
+interface is small. On the other hand, and this is what we are
 interested in here, this is not necessarily the case for regimes
 specific to flow in high-permeability porous media for which the extent
-of the fluid-fluid interface is large. This is important because, as we
-will see in the next section, these exchange terms between phases are
+of the fluid-fluid interface is large. This is important because, as
+discussed in the next section, these exchange terms between phases are
 the basis of any attempt to establish continuous relationships on a
@@
 
-Here we present two different approaches to treat the continuous
-macroscopic momentum equations for two-phase flow. Although
-fundamentally identical, these approaches were developed for the needs
-of two different specialties, namely, the study of groundwater flows and
-the study of the hydrodynamic of chemical reactors.
-
-\hypertarget{porous-media-approach}{%
-\subsubsection{Porous media approach}\label{porous-media-approach}}
-
-The ubiquitous continuous model used to describe two-phase flows in
-soils is based on an extension of the Darcy equation (creeping saturated
-single-phase flow in porous media). The whole model, also known as
-Muskat equations
-{[}\protect\hyperlink{ref-wyckoff1936flow}{8},\protect\hyperlink{ref-muskat1938flow}{9}{]},
+Volume averaged microscopic equation of motion for two Newtonian fluids
+\(i,j=o,w\:\mathrm{and}\:i\neq j\) reads
+
+\[
+\rho_{i}\varepsilon_{i}\left(\frac{\partial\mathbf{U}_{i}}{\partial t}+\mathbf{U}_{i}\cdot\nabla\mathbf{U}_{i}\right)=-\varepsilon_{i}\nabla P_{i}+\varepsilon_{i}\rho_{i}\mathbf{g}+\varepsilon_{i}\nabla\cdot\boldsymbol{\tau}_{i}+(\mathbf{F}_{ij}+\mathbf{F}_{is})
+\] \{\#eq:averagedMomentum\}
+
+where \(\varepsilon_{i}=V_i/V\) is the volume fraction of the
+\(i\)-fluid and \(V\) is a representative elementary volume, \(U_i\) is
+the intrinsic average velocity of the \(i\)-fluid, \(P_i\) is the
+average pressure of the \(i\)-fluid, \(\boldsymbol{\tau}_{i}\) denotes
+the average viscous stress tensor. The last two terms in the left hand
+side denote the force of interaction per unite volume exerted by fluid
+\(i\) upon fluid \(j\) and the force of interaction exerted by fluid
+\(i\) upon the solid phase, respectively
+{[}\protect\hyperlink{ref-Kalaydjian1987}{9}{]}.
+
+From there, there are two possible ways to solve +\textbf{???} and the
+associated equations. In volume averaging litterature, one seeks to
+express the terms of interaction between the phases with averaged
+quantities and to recast equation +\textbf{???} in a form similar to
+Darcy's law (equation for creeping saturated single-phase flow in porous
+media). Indeed, by considering creeping two-phase flow +\textbf{???} can
+be written as
+
+\[
+0=-\varepsilon_{i}\nabla P_{i}+\varepsilon_{i}\rho_{i}\mathbf{g}+\frac{1}{V}\int_{A_{is}+A_{ij}}\mathbf{n}_{i}\cdot\left(-\mathbf{I}p_{i}+\mu_{i}\left(\nabla\mathbf{u}_{i}+\left(\nabla\mathbf{u}_{i}\right)^{T}\right)\right)\mathrm{d}A.
+\] \{\#eq:VAmomentum\}
+
+where \(p_i\) and \(u_i\) are microscopic pressure and velocity fields
+of fluid \(i\), respectively. By expressing the surface integrals in
+term of averaged quantities one can obtain the following momentum
+balance equations,
+
+\[
+\mathbf{U}_{i}=-\frac{1}{\mu_{i}}\mathbf{K}_{ii}^{*}\cdot(\nabla P_{i}-\rho_{i}\mathbf{g})-\frac{1}{\mu_{j}}\mathbf{K}_{ij}^{*}\cdot(\nabla P_{j}-\rho_{j}\mathbf{g}),\quad i,j=o,w\:\mathrm{and}\:i\neq j,
+\] \{\#eq:darcyCross\}
+
+in which \(\mathbf{K}_{ij}^{*}\) are the coupled relative permeability
+tensors that pertain to the interaction between the fluids
+{[}\protect\hyperlink{ref-Whitaker1986a}{10},\protect\hyperlink{ref-Lasseux1996}{11}{]}.
+
+The reader is warned that +\textbf{???} has, to our knowledge, never
+been used in practice to model two-phase subsurface flows. On the
+contrary, the ubiquitous continuous model used to describe two-phase
+flows in soils is based on a direct extension of the Darcy's equation
+and the whole model, also known as Muskat equations
+{[}\protect\hyperlink{ref-wyckoff1936flow}{12},\protect\hyperlink{ref-muskat1938flow}{13}{]},
 reads
@@
 
-In these equations, the subscripts \(o\) and \(w\) refer to the
-non-wetting and wetting phase, respectively. \(S_{i}\) is the saturation
-of phase \(i\), \(\varepsilon\) is the medium porosity,
-\(\mathbf{U}_{i}\) is the superficial mean velocity of fluid \(i\),
-\(P_i\) is the intrinsic mean pressure of fluid \(i\) and \(\mathbf{K}\)
-is the absolute permeability tensor. The generalization toward two-phase
-flows involves the introduction of the relative permeability terms
-\(k_{ri}\) which account for one phase for the inaccessible void space
-occupied by the other fluid
-{[}\protect\hyperlink{ref-dullien2012porous}{5}{]}. To close the set of
-the macroscopic equations, the relative permeabilities are assumed to be
-non-linear functions of the saturation and a relation between the
-macroscopic pressure of each fluid has to be furnished. This relation,
-is known as the capillary pressure relation and, as for the relative
-permeabilities, is supposed to vary non-linearly with the saturation
-only {[}\protect\hyperlink{ref-leverett1941capillary}{10}{]}.
+where \(\mathbf{K}\) is the absolute permeability tensor. The
+generalization toward two-phase flows involves the introduction of the
+relative permeability terms \(k_{ri}\) which account for the division of
+the void space between the fluids
+{[}\protect\hyperlink{ref-dullien2012porous}{5}{]}, thus the relative
+permeability depends (non-linearly) only on the saturation. To close the
+set of the macroscopic equations a constitutive relation between the
+macroscopic pressure of each fluid has to be furnished. This relation is
+known as the capillary pressure relation and, as for the relative
+permeabilities, is supposed to depends non-linearly only on the
+saturation {[}\protect\hyperlink{ref-leverett1941capillary}{14}{]}.
 
@@
 Muskat equations (e.g.~including moving contact line
-{[}\protect\hyperlink{ref-kalaydjian1992dynamic}{11}--\protect\hyperlink{ref-barenblatt2003mathematical}{13}{]}
+{[}\protect\hyperlink{ref-kalaydjian1992dynamic}{15}--\protect\hyperlink{ref-barenblatt2003mathematical}{17}{]}
 or take into account the trapped phases
-{[}\protect\hyperlink{ref-hilfer1998macroscopic}{14}{]}). Here we would
-stress that the concept of relative permeabilities into the Muskat
+{[}\protect\hyperlink{ref-hilfer1998macroscopic}{18}{]}). Here we would
+only stress that the concept of relative permeabilities into the Muskat
 equations is dedicated to independent flow pathways only
-{[}\protect\hyperlink{ref-blunt2017multiphase}{15}{]} and the underlying
-assumption that the two streams do not interfere with each other.
-However, in case that the area of the fluid-fluid interface is not
-negligible compared to the solid-fluid interface, as for flow regimes
-\cref{fig:fluidPatterns} (b) and \cref{fig:fluidPatterns} (c), one would
-include supplementary terms dedicated to the momentum exchange across
-the fluid-fluid interface due to normal and shear stresses exerted on it
-(i.e.~coupling between the phases) into the macroscopic momentum
-equations. In fact, it has been theoretically established that the
-macroscopic equations of motion obtained by averaging of the Stokes
-equations contain momentum exchange terms between fluids
-{[}\protect\hyperlink{ref-MARLE1982643}{16}--\protect\hyperlink{ref-kalaydjian1990origin}{19}{]}.
-The volume averaged momentum equation for the \(i\)-phase read
-
-\[
-0=-\varepsilon_{i}\nabla P_{i}+\varepsilon_{i}\rho_{i}\mathbf{g}+\frac{1}{V}\int_{A_{is}+A_{ij}}\mathbf{n}_{i}\cdot\left(-\mathbf{I}p_{i}+\mu_{i}\left(\nabla\mathbf{u}_{i}+\left(\nabla\mathbf{u}_{i}\right)^{T}\right)\right)\mathrm{d}A.
-\] \{\#eq:VAmomentum\}
-
-The next step in the upscaling methods literature is to express the
-boundaries integral of exchange between the phases in terms of averaged
-quantity through the resolution of boundary value problems on a
-representative periodic unit-cell. We need to go no further here but we
-insist that, as shown in {[}\protect\hyperlink{ref-Lasseux1996}{20}{]},
-once the closure problems have been solved, the macroscopic equations of
-motion can be put into a form analogous to the generalized Darcy's law,
-
-\[
-\mathbf{U}_{i}=-\frac{1}{\mu_{i}}\mathbf{K}_{ii}^{*}\cdot(\nabla P_{i}-\rho_{i}\mathbf{g})-\frac{1}{\mu_{j}}\mathbf{K}_{ij}^{*}\cdot(\nabla P_{j}-\rho_{j}\mathbf{g}),\quad i,j=o,w\:\mathrm{and}\:i\neq j,
-\] \{\#eq:darcyCross\}
-
-in which \(\mathbf{K}_{ij}^{*}\) are the coupled relative permeability
-tensors.
-
-Many studies have attempted to calculate coupled permeabilities
-experimentally based on the +\textbf{???}, but before we turn to present
-results from the literature, we discuss the chemical engineering point
-of view on the continuous model.
-
-\hypertarget{chemical-engineering-approach}{%
-\subsubsection{Chemical engineering
-approach}\label{chemical-engineering-approach}}
-
-CFD modeling applied to the hydrodynamics of trickle-bed reactors (TBR)
-is in most cases focused on an Euler-Euler flow formulation. The two
-fluids are thus treated as two interpenetrating continua. The volume
-averaged momentum equation for the two fluids to solve, taking into
-account that the inertial effects are not negligible here, read
-
-\[
-\rho_{i}\varepsilon_{i}\left(\frac{\partial\mathbf{U}_{i}}{\partial t}+\mathbf{U}_{i}\cdot\nabla\mathbf{U}_{i}\right)=-\varepsilon_{i}\nabla P_{i}+\varepsilon_{i}\rho_{i}\mathbf{g}+\mathbf{F}_{mi}+\varepsilon_{i}\nabla\cdot\boldsymbol{\tau}_{i}+\nabla\cdot\mathbf{R}_{i},
-\] \{\#eq:trickleMomentum\}
-
-where \(\varepsilon_{i}\) is the volume fraction of phase \(i\), the
-velocity and the pressure are averaged quantities identical to the
-generalized Darcy equations, \(\boldsymbol{\tau}_{i}\) denotes the
-viscous stress tensor and \(\mathbf{R}_{i}\) is the pseudo-turbulence
-stress. The vectors \(\mathbf{F}_{mi}\) are the momentum transfer terms
-at the interface for the \(i\)-phase. These terms have to be expressed
-in terms of constitutive relation to close the set of equations, and the
-main difference between the existing continuous models is the choice of
-correlations used for these terms. We stress here that the two equations
-+\textbf{???} and +\textbf{???} are equivalents (except for the
-non-linear inertial terms).
-
-\hypertarget{modeling-of-the-momentum-exchange-between-fluids}{%
-\subsubsection{Modeling of the momentum exchange between
-fluids}\label{modeling-of-the-momentum-exchange-between-fluids}}
-
-Here we present different estimation of the momentum transfer terms
-(i.e.~coupled relative permeabilities in the porous media framework)
-between fluids. Based on the generalized Darcy law with cross terms
-+\textbf{???}, experimental works computed the coupled transport
-coefficient through steady-state cocurent flow in sandpack with
-alternatively one fluid, then the other, which is submitted to a null
-pressure gradient. This protocol was used with oil and water in a
-cylindric sandpack
-{[}\protect\hyperlink{ref-zarcone1994determination}{21}{]} and the
+{[}\protect\hyperlink{ref-blunt2017multiphase}{19}{]} and the underlying
+assumption that the two streams do not interfere with each other, a
+situation far from the regimes previously identified for high
+permeability porous media. Thus +\textbf{???} does not take into account
+the interaction between the fluids as opposed to +\textbf{???}. If the
+former is used to the detriment of the latter, it is mainly because the
+literature on subsurface flow is focused on oil recovery applications,
+for which the permeability is very low and the independent streams
+regime is a good approximation.
+
+Another approach is to implement +\textbf{???} and provide constitutive
+relations for the interaction terms between the phases. These relations
+are usually obtained through interpretation of experimental data. Some
+results, focusing on relation for the interaction between the fluid
+phases, are given in the next section as well as results in literature
+for coupled permeabilities from the porous medium approach.
+
+\hypertarget{modeling-of-the-interaction-between-fluids}{%
+\subsubsection{Modeling of the interaction between
+fluids}\label{modeling-of-the-interaction-between-fluids}}
+
+Based on the generalized Darcy law with coupled terms +\textbf{???},
+experimental works computed the coupled transport coefficient through
+steady-state cocurent flow in sandpack with one fluid, and alternatively
+the other, which is submitted to a null pressure gradient. This protocol
+was used with oil and water in a cylindric sandpack
+{[}\protect\hyperlink{ref-zarcone1994determination}{20}{]} and the
 authors found a negligible effect of the coupled permeabilities in the
@@
 2D-sandpack Dullien and Dong
-{[}\protect\hyperlink{ref-Dullien1996}{22}{]} found that the coupled
+{[}\protect\hyperlink{ref-Dullien1996}{21}{]} found that the coupled
 permeabilities are important since they can contribute at best to 35\%
 of the effective permeability. Alternatively, authors imposed
-{[}\protect\hyperlink{ref-ramakrishnan2015measurement}{23}{]} a null
+{[}\protect\hyperlink{ref-ramakrishnan2015measurement}{22}{]} a null
 displacement and measured the induced pressure drop, with air and water
@@
 Several authors conducted experiments with two different set-ups as
-proposed by Rose {[}\protect\hyperlink{ref-Rose1988}{24}{]}, e.g.
-{[}\protect\hyperlink{ref-bentsen1993use}{25}{]}, in which the authors
+proposed by Rose {[}\protect\hyperlink{ref-Rose1988}{23}{]}, e.g.
+{[}\protect\hyperlink{ref-bentsen1993use}{24}{]}, in which the authors
 made cocurrent and countercurrent experiments with water and oil in a
@@
 therefore the computed relative permeabilities can not be safely
-compared {[}\protect\hyperlink{ref-langaas2001numerical}{26}{]}. The
+compared {[}\protect\hyperlink{ref-langaas2001numerical}{25}{]}. The
 effect of the non-wetting phase connectivity on the transport parameters
-was extensively studied in {[}\protect\hyperlink{ref-Avraam1995}{27}{]},
+was extensively studied in {[}\protect\hyperlink{ref-Avraam1995}{26}{]},
 the authors performed steady-state cocurrent two-phase flow in 2D-micro
@@
 
-Several numerical studies investigate the share of the coupled
+Several numerical studies investigated the share of the coupled
 permeabilities in the effective permeability. The seminal work of
-Rothman {[}\protect\hyperlink{ref-Rothman1990}{28}{]} examined the
+Rothman {[}\protect\hyperlink{ref-Rothman1990}{27}{]} examined the
 question by conducting two-phase flow simulations in simple geometries
@@
 alternatively on each fluid. Numerous authors used the lattice-Boltzmann
-method, such as {[}\protect\hyperlink{ref-Li2005}{29}{]}, in which the
+method, such as {[}\protect\hyperlink{ref-Li2005}{28}{]}, in which the
 authors examined the value of the coupled permeabilities as a function
@@
 agreement with Rothman's results. In
-{[}\protect\hyperlink{ref-Yiotis2007}{30}{]} the authors used a lattice
+{[}\protect\hyperlink{ref-Yiotis2007}{29}{]} the authors used a lattice
 Boltzmann method in 2D and 3D pore networks and found a non-wetting
@@
 manifestation of the lubrification effect, also observed in experiments
-{[}\protect\hyperlink{ref-odeh1959effect}{31}{]}, and arising because of
-a strongly hydraulic coupling between the fluids. Recently, in
-{[}\protect\hyperlink{ref-shams2018study}{32}{]}, the authors have used
+{[}\protect\hyperlink{ref-odeh1959effect}{30}{]}, and arising because of
+strong hydraulic coupling between the fluids. Recently, in
+{[}\protect\hyperlink{ref-shams2018study}{31}{]}, the authors have used
 a Volume Of Fluid method to study the transport coefficients of fluid
@@
 Couette flow the authors have derived simple relations that can predict
-with good accuracy the transport coefficients of fluid layers.
+with good accuracy the transport coefficients, including the coupled
+ones, of fluid layers. We can also refer to recent work in which a whole
+analytical model is given for one-dimensional two-phase flow in coarse
+non-consolidated porous media and which correctly predicts the pressure
+loss in a debris bed
+{[}\protect\hyperlink{ref-clavier2017modeling}{2}{]}.
+
+Several correlations from theorical analysis for the fluid-fluid force
+interaction have been derived in hydrodynamic of TBR litterature. These
+correlations usually relate the momentum exchange between the fluids to
+the flow velocity and the medium properties with Ergun's like relations.
+For example, Attou et al. {[}\protect\hyperlink{ref-Attou1999}{32}{]}
+found that the drag between the fluids can be expressed as follows
+
+\[
+F_{lg} = A_{lg} U_r + B_{lg} U_r^2,
+\]
+
+where the relative velocity is given by \(U_r=U_g-\frac{S_g}{S_l}U_l\)
+and \(A_{lg}\) and \(B_{lg}\) are coefficients that depend on the volume
+fraction of the gas phase and the medium properties. The subscripts
+\(g\) and \(l\) denote the gas phase and the liquid phase, respectively.
+This type of model with explicit account of interaction between fluids,
+see also
+{[}{[}\protect\hyperlink{ref-tung1988hydrodynamic}{33}{]};iliuta2005modelling{]},
+gives good results when compared with experimental results
+{[}\protect\hyperlink{ref-wang2013modelling}{34}{]}.
 
@@
 
-In this study, we investigate the effect of the varying intrinsic
-permeability and capillary number on the momentum transfer terms through
-direct numerical simulations of two-phase flow in a model
-high-permeability 2D-porous media. To do so we compute the surface
-integrals in +\textbf{???} on each boundaries of the system. The idea of
-solving a depth-averaged flow in a Hele-Shaw allows us to introduce an
-additional permeability term and thus to vary the intrinsic permeability
-of the strucure without having to modify the geometry per se.
-
 \hypertarget{theoretical-background}{%
@@
 
-In the following, the superficial mean of a physical quantity
+In the following, the superficial average of a physical quantity
 \(\theta_i\) associated with the \(i\)-phase is given by
@@
 
-Macroscopic quantities such as saturation and intrinsic mean pressure
+Macroscopic quantities such as saturation and intrinsic average pressure
 for two fluids \(i=w,o\) read
@@
 
-As previously said, an essential step in order to derive some
-macroscopic law at macroscale from the microscale is to use averaged
-quantities to express the momentum transfer terms between phases that
-arising because of the averaging theorem application
-{[}\protect\hyperlink{ref-Whitaker2013}{33}{]}. Remember that these
-closure relations are not presented here but rather we leave intact the
-surface integrals, the averaged momentum balance equations +\textbf{???}
-for two fluids \(w\) and \(o\), are recast as
+As said, an essential step in order to derive some macroscopic law at
+macroscale from the microscale is to use averaged quantities to express
+the boundary intergals terms that arising because of the averaging
+theorem application {[}\protect\hyperlink{ref-Whitaker2013}{35}{]}.
+Remember that these closure relations are not presented here but rather
+we leave intact the surface integrals, the averaged momentum balance
+equations +\textbf{???} for two fluids \(w\) and \(o\), are recast as
 
@@
 
-where \(\varepsilon_{i}\) and \(P_{i}\) are the macroscopic quantities
-defined above, \(\mathbf{T}_{ij}\) the traction (name given to surface
-forces in the Boundary Integral Methods literature in which these
-integrals are central) exerted by the \(i\)-fluid on the \(j\)-fluid and
-\(\mathbf{T}_{is}\) the traction exerted by the \(i\)-fluid on the solid
-surfaces {[}\protect\hyperlink{ref-Kalaydjian1987}{34}{]}. The traction
-terms are written as
+where \(\mathbf{T}_{ij}\) are the traction (name given to surface forces
+in the Boundary Integral Methods literature in which these integrals are
+central) exerted by fluid \(i\) upon fluid \(j\) and \(\mathbf{T}_{is}\)
+the traction exerted by fluid \(i\) upon the solid
+{[}\protect\hyperlink{ref-Kalaydjian1987}{9}{]}. The traction terms are
+written as
 
@@
 pointing toward the j-phase, \(p_{i}\) the pointwise pressure of the
-i-fluid and
+fluid \(i\) and
 \(\mathbf{e}_{i}=\frac{1}{2}\left(\nabla\mathbf{u}_{i}+\left(\nabla\mathbf{u}_{i}\right)^{T}\right)\)
 the rate-of-strain tensor for Newtonian fluids with \(\mathbf{u}_{i}\)
-the pointwise velocity of the i-fluid.
+the pointwise velocity of fluid \(i\).
 
@@
 
-Two-phase flow between two parallel plates under study here is depicted
-in \cref{fig:modelStudy}. This illustration represents a cocurrent flow
+Two-phase flow between two parallel plates under study here is
+illustrated in \cref{fig:modelStudy}. This represents a cocurrent flow
 of two fluids inside a Hele-Shaw cell parallel to the \(x-y\) plane,
@@
 
+To compute the traction exerted by the fluids upon the cell plates we
+need to consider that the velocity profile in the \(z\)-direction is
+well described by the Poiseuille flow profile,
+
+\[
+u=-\frac{1}{2\mu}\frac{dp}{dx}z(z-h).
+\] \{\#eq:poiseuilleProfile\}
+
+Then the friction force exerted on the bottom plate is given by
+
+\[
+\mu\frac{\partial u}{\partial z}\mid_{z=0}=-\frac{h}{2}\frac{dp}{dx}.
+\] \{\#eq:poiseuilleFriction\}
+
+Thus, the traction \(T_{i,pl}^x\) is obtained by multiplying the
+friction force by two times the surface wetted by the fluid \(i\).
+
+We note that the average velocity taken over the interval between the
+plates is written as follows
+
+\[
+\mathbf{U}(x,y,0)=-\frac{h^{2}}{12\mu}\nabla_{\parallel}p.
+\] \{\#eq:darcyHeleShaw\}
+
+There is therefore a direct analogy between flows in a 2D porous medium
+and flow in a Hele-Shaw cell, as can be seen in +\textbf{???}, where the
+permeability is given by \(h^{2}/12\)
+{[}\protect\hyperlink{ref-saffman1958penetration}{36}{]}.
+
 Beside the interface curvature in the Hele-Shaw cell plane, the meniscus
@@
 the curvature in the Hele-Shaw plane
-{[}\protect\hyperlink{ref-saffman1958penetration}{35}{]}. The
+{[}\protect\hyperlink{ref-saffman1958penetration}{36}{]}. The
 +\textbf{???} was completed in
-{[}\protect\hyperlink{ref-park1984two}{36}{]} by an additional term that
+{[}\protect\hyperlink{ref-park1984two}{37}{]} by an additional term that
 pertains for dynamic film formation and a corrective \(\pi/4\) factor
-for the curvature in the cell plane.
-
-To compute the traction exerted by the fluids on the cell plates we need
-to consider that the velocity profile in the \(z\)-direction is well
-described by the Poiseuille flow profile,
-
-\[
-u=-\frac{1}{2\mu}\frac{dp}{dx}z(z-h).
-\] \{\#eq:poiseuilleProfile\}
-
-Then the friction force exerted on the bottom plate is given by
-
-\[
-\mu\frac{\partial u}{\partial z}\mid_{z=0}=-\frac{h}{2}\frac{dp}{dx}.
-\] \{\#eq:poiseuilleFriction\}
-
-Thus, the traction \(T_{i,pl}^x\) is obtained by multiplying the
-friction force by two times the surface wetted by the fluid \(i\).
-
-We can show that the average velocity taken over the interval between
-the plates is written as follows
-
-\[
-\mathbf{U}(x,y,0)=-\frac{h^{2}}{12\mu}\nabla_{\parallel}p.
-\] \{\#eq:darcyHeleShaw\}
-
-There is therefore a direct analogy between flows in a 2D porous medium
-and a flow in a Hele-Shaw cell, as can be seen in +\textbf{???}, where
-the permeability is given by \(h^{2}/12\)
-{[}\protect\hyperlink{ref-saffman1958penetration}{35}{]}.
+for the curvature in the cell plane. The reader is warnes that while
+dynamic film formation along the plates is an important phenomenom
+regarding the flow {[}\protect\hyperlink{ref-jackson2015dynamic}{38}{]},
+this requires dedicated study and is beyond the scope of this article.
+Thus, we consider in the following that the invading fluid is in contact
+with the cell plates, i.e.~presence of a triple line.
 
@@
 respectively
-{[}\protect\hyperlink{ref-Olsson2005}{37},\protect\hyperlink{ref-Olsson2007}{38}{]}.
-
-\hypertarget{flow-model}{%
-\subsubsection{Flow model}\label{flow-model}}
+{[}\protect\hyperlink{ref-Olsson2005}{39},\protect\hyperlink{ref-Olsson2007}{40}{]}.
+
+\hypertarget{flow-equations}{%
+\subsubsection{Flow equations}\label{flow-equations}}
 
@@
 object of a large body of work
-{[}\protect\hyperlink{ref-Popinet2018}{39}{]}, this term is defined here
+{[}\protect\hyperlink{ref-Popinet2018}{41}{]}, this term is defined here
 as
@@
 and \(\delta(\phi)\) is the Dirac delta function localized on the
-interface {[}\protect\hyperlink{ref-Galusinski2008}{40}{]}. In this
+interface {[}\protect\hyperlink{ref-Galusinski2008}{42}{]}. In this
 framework the unit normal vector as well as the mean curvature are
@@
 respectively, where the \(\pi/4\) correction was derived in
-{[}\protect\hyperlink{ref-park1984two}{36}{]} and
+{[}\protect\hyperlink{ref-park1984two}{37}{]} and
 \(\kappa_{\perp}=\frac{2}{h}\) is the out-of-plane curvature as
@@
 
-\hypertarget{validation}{%
-\subsubsection{Validation}\label{validation}}
-
 \hypertarget{results}{%
@@
 intrinsic permeability for flow between two plates or from single-phase
-simulation in our model. \label{tbl:permeability}}\tabularnewline
+simulation. \label{tbl:permeability}}\tabularnewline
 \toprule
@@
 0.2 & \(5.21 \times 10^{5}\) & \(1.48 \times 10^{4}\)\tabularnewline
-2.0 & \(5.21 \times 10^{3}\) & \(2.85 \times 10^{3}\)\tabularnewline
-4.0 & \(1.30 \times 10^{3}\) & \(8.63 \times 10^{2}\)\tabularnewline
-8.0 & \(3.26 \times 10^{2}\) & \(2.35 \times 10^{2}\)\tabularnewline
+2 & \(5.21 \times 10^{3}\) & \(2.85 \times 10^{3}\)\tabularnewline
+4 & \(1.30 \times 10^{3}\) & \(8.63 \times 10^{2}\)\tabularnewline
+8 & \(3.26 \times 10^{2}\) & \(2.35 \times 10^{2}\)\tabularnewline
 20 & \(5.21 \times 10^{1}\) & \(3.94 \times 10^{1}\)\tabularnewline
@@
 \centering
-\includegraphics{figures/pdf/drag_plates.pdf}
+\includegraphics{figures/pdf/dragPlates.pdf}
 \caption{\(x\)-component of the traction on Hele-Shaw cell plates by
@@
 \centering
-\includegraphics{figures/pdf/drag_solid.pdf}
+\includegraphics{figures/pdf/dragSolid.pdf}
 \caption{\(x\)-component of the traction exerted on the solid-fluid
@@
 \centering
-\includegraphics{figures/pdf/drag_fluidint.pdf}
+\includegraphics{figures/pdf/dragFluid.pdf}
 \caption{\(x\)-component of the traction exerted on the fluid-fluid
@@
 
-\hypertarget{decreasing-the-permeability-does-not-imply-that-the-share-of-the-fluid-fluid-traction-is-negligible-in-the-total-traction}{%
-\subsection{Decreasing the permeability does not imply that the share of
-the fluid-fluid traction is negligible in the total
-traction}\label{decreasing-the-permeability-does-not-imply-that-the-share-of-the-fluid-fluid-traction-is-negligible-in-the-total-traction}}
+\hypertarget{decreasing-the-intrinsic-permeability-does-not-imply-that-the-share-of-the-fluid-fluid-traction-is-negligible-in-the-total-traction}{%
+\subsection{Decreasing the intrinsic permeability does not imply that
+the share of the fluid-fluid traction is negligible in the total
+traction}\label{decreasing-the-intrinsic-permeability-does-not-imply-that-the-share-of-the-fluid-fluid-traction-is-negligible-in-the-total-traction}}
 
@@
 share of traction exerted on the fluid-fluid boundaries in total
-traction. The ratio \(T_{int}^x/T_{s}^x\), were \(s={pl+cyl}\) and
-\(int={wo+ow}\), is given as a function of the dimensionless aperture
-between the Hele-Shaw plates in \cref{fig:ratioDrag}. The traction
-exerted on the fluid-fluid boundary reaches between 10 to 50\% of the
-traction exerted on the solid-fluid boundaries. Interestingly, the ratio
-varies non-monotonically with \(h^*\), the minimum is reached for the
-smallest capillary numbers and for a medium range of the aperture
-values.
-
-As the capillary number decreases, variations with the aperture are
-amplified. Thus, it is for the lowest capillary number tested that we
-obtain both the maximum and minimum values of the ratio
+traction. The ratio \(T_{int}^x/T_{s}^x\) is given as a function of the
+dimensionless aperture between the Hele-Shaw plates in
+\cref{fig:ratioDrag}. The traction exerted on the fluid-fluid boundary
+reaches between 10 and 50\% of the traction exerted on the solid-fluid
+boundaries. Interestingly, the ratio varies non-monotonically with
+\(h^*\), the minimum is reached for the smallest capillary numbers and a
+medium range of the aperture values.
+
+Variations with aperture between plates are amplified as the capillary
+number decreases. Thus, it is for the lowest capillary number tested
+that we obtain both the maximum and minimu m values of the ratio
 \(T_{int}^x/T_{s}^x\).
@@
 \includegraphics{figures/pdf/ratioDrag.pdf}
-\caption{Ratio of traction on fluid-fluid interface over traction on
-solid-fluid interfaces in UC3 as a function of the dimensionless
-aperture between the Hele-Shaw plates and for different capillary
-numbers at steady-state.}\label{fig:ratioDrag}
+\caption{Ratio of traction (\(x\)-component) on fluid-fluid interface
+over traction on solid-fluid interfaces in UC3 as a function of the
+dimensionless aperture between the Hele-Shaw plates and for different
+capillary numbers at steady-state.}\label{fig:ratioDrag}
 }
@@
 supplementary Darcy term allowed us to modify the intrinsic permeability
-of the structure without having to modify the geometry per se. The
-chosen range for intrinsic permeability values is typical of
+of the structure without having to modify the geometry per se. The range
+of intrinsic permeability values investigated is typical of
 high-permeability porous media, for which it is known that the flow
 regimes are characterized by the large extent of the fluid-fluid
-interface the fluids (e.g.~film flow). Through direct numerical
-simulation and a Level Set method to handle the free interface between
-the fluid we have found that
+interface (e.g.~film flow). With direct numerical simulations we found
+that
+
+\hypertarget{appendix}{%
+\section{Appendix}\label{appendix}}
+
+\hypertarget{validiation-against-explicit-tracking-interface-method}{%
+\subsection{Validiation against explicit tracking interface
+method}\label{validiation-against-explicit-tracking-interface-method}}
+
+The code is validated by comparison with a Boundary-Element Method,
+which relies on a surface discretization of the interface and a
+pseudo-analytical formulation in the bulk of the phases. This allows us
+to precisely locate the interface, even in the case of very thin film
+flow, and to carefully analyze the choice of parameters in Eq.
+{[}eq:advecPhi{]}.
+
+The test case resembles the study case as it consists of the injection
+of a fluid into a Hele-Shaw initially fully saturated with a flowing
+fluid. The dynamic viscosity of the fluids is equal and the model is
+illustrated in \cref{fig:testCase}.
+
+\begin{figure}
+\hypertarget{fig:testCase}{%
+\centering
+\includegraphics{figures/pdf/TestCase_validation.pdf}
+\caption{The test case consists in the injection of fluid 1 into a Hele
+Shaw cell initially fully saturated with a flowing fluid 2, inlet
+boundary conditions are constant velocity along the \(x\)-axis, the
+outlet boundary condition is}\label{fig:testCase}
+}
+\end{figure}
 
@@
 
+\leavevmode\hypertarget{ref-taitel1976model}{}%
+{[}6{]} Taitel Y, Dukler AE. A model for predicting flow regime
+transitions in horizontal and near horizontal gas-liquid flow. AIChE
+Journal 1976;22:47--55.
+
 \leavevmode\hypertarget{ref-Avraam1995a}{}%
-{[}6{]} Avraam DG, Payatakes AC. Flow regimes and relative
+{[}7{]} Avraam DG, Payatakes AC. Flow regimes and relative
 permeabilities during steady-state two-phase flow in porous media.
@@
 \leavevmode\hypertarget{ref-armstrong2016beyond}{}%
-{[}7{]} Armstrong RT, McClure JE, Berrill MA, Rücker M, Schlüter S, Berg
+{[}8{]} Armstrong RT, McClure JE, Berrill MA, Rücker M, Schlüter S, Berg
 S. Beyond darcy's law: The role of phase topology and ganglion dynamics
@@
 
+\leavevmode\hypertarget{ref-Kalaydjian1987}{}%
+{[}9{]} Kalaydjian F. A macroscopic description of multiphase flow in
+porous media involving spacetime evolution of fluid/fluid interface.
+Transport in Porous Media 1987;2:537--52.
+\url{https://doi.org/10.1007/BF00192154}.
+
+\leavevmode\hypertarget{ref-Whitaker1986a}{}%
+{[}10{]} Whitaker S. Flow in porous media II: The governing equations
+for immiscible, two-phase flow. Transport in Porous Media
+1986;1:105--25.
+
+\leavevmode\hypertarget{ref-Lasseux1996}{}%
+{[}11{]} Lasseux D, Quintard M, Whitaker S. Determination of
+permeability tensors for two-phase flow in homogeneous porous media:
+Theory. Transport in Porous Media 1996;24:107--37.
+
 \leavevmode\hypertarget{ref-wyckoff1936flow}{}%
-{[}8{]} Wyckoff RD, Botset HG. The flow of gas-liquid mixtures through
+{[}12{]} Wyckoff RD, Botset HG. The flow of gas-liquid mixtures through
 unconsolidated sands. Physics 1936;7:325--45.
@@
 \leavevmode\hypertarget{ref-muskat1938flow}{}%
-{[}9{]} Muskat M. The flow of homogeneous fluids through porous media.
+{[}13{]} Muskat M. The flow of homogeneous fluids through porous media.
 Soil Science 1938;46:169.
@@
 \leavevmode\hypertarget{ref-leverett1941capillary}{}%
-{[}10{]} Leverett M, others. Capillary behavior in porous solids.
+{[}14{]} Leverett M, others. Capillary behavior in porous solids.
 Transactions of the AIME 1941;142:152--69.
@@
 \leavevmode\hypertarget{ref-kalaydjian1992dynamic}{}%
-{[}11{]} Kalaydjian F-M, others. Dynamic capillary pressure curve for
+{[}15{]} Kalaydjian F-M, others. Dynamic capillary pressure curve for
 water/oil displacement in porous media: Theory vs. Experiment. In:. SPE
@@
 \leavevmode\hypertarget{ref-Hassanizadeh1993}{}%
-{[}12{]} Hassanizadeh SM, Gray WG. Thermodynamic basis of capillary
+{[}16{]} Hassanizadeh SM, Gray WG. Thermodynamic basis of capillary
 pressure in porous media. Water Resources Research 1993;29:3389--405.
@@
 \leavevmode\hypertarget{ref-barenblatt2003mathematical}{}%
-{[}13{]} Barenblatt GI, Patzek TW, Silin DB. The mathematical model of
+{[}17{]} Barenblatt GI, Patzek TW, Silin DB. The mathematical model of
 nonequilibrium effects in water-oil displacement. SPE Journal
@@
 \leavevmode\hypertarget{ref-hilfer1998macroscopic}{}%
-{[}14{]} Hilfer R. Macroscopic equations of motion for two-phase flow in
+{[}18{]} Hilfer R. Macroscopic equations of motion for two-phase flow in
 porous media. Physical Review E 1998;58:2090.
@@
 \leavevmode\hypertarget{ref-blunt2017multiphase}{}%
-{[}15{]} Blunt MJ. Multiphase flow in permeable media: A pore-scale
+{[}19{]} Blunt MJ. Multiphase flow in permeable media: A pore-scale
 perspective. Cambridge University Press; 2017.
 
-\leavevmode\hypertarget{ref-MARLE1982643}{}%
-{[}16{]} Marle CM. On macroscopic equations governing multiphase flow
-with diffusion and chemical reactions in porous media. International
-Journal of Engineering Science 1982;20:643--62.
-\url{https://doi.org/https://doi.org/10.1016/0020-7225(82)90118-5}.
-
-\leavevmode\hypertarget{ref-Whitaker1986a}{}%
-{[}17{]} Whitaker S. Flow in porous media II: The governing equations
-for immiscible, two-phase flow. Transport in Porous Media
-1986;1:105--25.
-
-\leavevmode\hypertarget{ref-auriault1986remarques}{}%
-{[}18{]} Auriault J, Sanchez-Palencia E. Remarques sur la loi de darcy
-pour les écoulements biphasiques en milieu poreux. Journal of
-Theoretical and Applied Mechanics, Numéro Spécial 1986:p141--156.
-
-\leavevmode\hypertarget{ref-kalaydjian1990origin}{}%
-{[}19{]} Kalaydjian F. Origin and quantification of coupling between
-relative permeabilities for two-phase flows in porous media. Transport
-in Porous Media 1990;5:215--29.
-
-\leavevmode\hypertarget{ref-Lasseux1996}{}%
-{[}20{]} Lasseux D, Quintard M, Whitaker S. Determination of
-permeability tensors for two-phase flow in homogeneous porous media:
-Theory. Transport in Porous Media 1996;24:107--37.
-
 \leavevmode\hypertarget{ref-zarcone1994determination}{}%
-{[}21{]} Zarcone C, Lenormand R. Détermination expérimentale du couplage
+{[}20{]} Zarcone C, Lenormand R. Détermination expérimentale du couplage
 visqueux dans les écoulements diphasiques en milieu poreux. Comptes
@@
 \leavevmode\hypertarget{ref-Dullien1996}{}%
-{[}22{]} Dullien FAL, Dong M. Experimental determination of the flow
+{[}21{]} Dullien FAL, Dong M. Experimental determination of the flow
 transport coefficients in the coupled equations of two-phase flow in
@@
 \leavevmode\hypertarget{ref-ramakrishnan2015measurement}{}%
-{[}23{]} Ramakrishnan TS, Goode PA. Measurement of off-diagonal
+{[}22{]} Ramakrishnan TS, Goode PA. Measurement of off-diagonal
 transport coefficients in two-phase flow in porous media. Journal of
@@
 \leavevmode\hypertarget{ref-Rose1988}{}%
-{[}24{]} Rose W. Measuring transport coefficients necessary for the
+{[}23{]} Rose W. Measuring transport coefficients necessary for the
 description of coupled two-phase flow of immiscible fluids in porous
@@
 \leavevmode\hypertarget{ref-bentsen1993use}{}%
-{[}25{]} Bentsen RG, Manai AA. On the use of conventional cocurrent and
+{[}24{]} Bentsen RG, Manai AA. On the use of conventional cocurrent and
 countercurrent effective permeabilities to estimate the four generalized
@@
 \leavevmode\hypertarget{ref-langaas2001numerical}{}%
-{[}26{]} Langaas K, Papatzacos P. Numerical investigations of the steady
+{[}25{]} Langaas K, Papatzacos P. Numerical investigations of the steady
 state relative permeability of a simplified porous medium. Transport in
@@
 \leavevmode\hypertarget{ref-Avraam1995}{}%
-{[}27{]} Avraam DG, Payatakes AC. Generalized relative permeability
+{[}26{]} Avraam DG, Payatakes AC. Generalized relative permeability
 coefficients during steady-state two-phase flow in porous media, and
@@
 \leavevmode\hypertarget{ref-Rothman1990}{}%
-{[}28{]} Rothman DH. Macroscopic laws for immiscible two-phase flow in
+{[}27{]} Rothman DH. Macroscopic laws for immiscible two-phase flow in
 porous media: Results from numerical experiments. Journal of Geophysical
@@
 \leavevmode\hypertarget{ref-Li2005}{}%
-{[}29{]} Li H, Pan C, Miller CT. Pore-scale investigation of viscous
+{[}28{]} Li H, Pan C, Miller CT. Pore-scale investigation of viscous
 coupling effects for two-phase flow in porous media. Physical Review E
@@
 \leavevmode\hypertarget{ref-Yiotis2007}{}%
-{[}30{]} Yiotis AG, Psihogios J, Kainourgiakis ME, Papaioannou A, Stubos
+{[}29{]} Yiotis AG, Psihogios J, Kainourgiakis ME, Papaioannou A, Stubos
 AK. A lattice boltzmann study of viscous coupling effects in immiscible
@@
 \leavevmode\hypertarget{ref-odeh1959effect}{}%
-{[}31{]} Odeh AS, others. Effect of viscosity ratio on relative
+{[}30{]} Odeh AS, others. Effect of viscosity ratio on relative
 permeability (includes associated paper 1496-g) 1959.
@@
 \leavevmode\hypertarget{ref-shams2018study}{}%
-{[}32{]} Shams M, Raeini AQ, Blunt MJ, Bijeljic B. A study to
+{[}31{]} Shams M, Raeini AQ, Blunt MJ, Bijeljic B. A study to
 investigate viscous coupling effects on the hydraulic conductance of
@@
 
+\leavevmode\hypertarget{ref-Attou1999}{}%
+{[}32{]} Attou A, Boyer C, Ferschneider G. Modelling of the
+hydrodynamics of the cocurrent gas--liquid trickle flow through a
+trickle-bed reactor. Chemical Engineering Science 1999;54:785--802.
+\url{https://doi.org/https://doi.org/10.1016/S0009-2509(98)00285-1}.
+
+\leavevmode\hypertarget{ref-tung1988hydrodynamic}{}%
+{[}33{]} Tung V, Dhir V. A hydrodynamic model for two-phase flow through
+porous media. International Journal of Multiphase Flow 1988;14:47--65.
+
+\leavevmode\hypertarget{ref-wang2013modelling}{}%
+{[}34{]} Wang Y, Chen J, Larachi F. Modelling and simulation of
+trickle-bed reactors using computational fluid dynamics: A
+state-of-the-art review. The Canadian Journal of Chemical Engineering
+2013;91:136--80.
+
 \leavevmode\hypertarget{ref-Whitaker2013}{}%
-{[}33{]} Whitaker S. The method of volume averaging. vol. 13. Springer
+{[}35{]} Whitaker S. The method of volume averaging. vol. 13. Springer
 Science \& Business Media; 2013.
 
-\leavevmode\hypertarget{ref-Kalaydjian1987}{}%
-{[}34{]} Kalaydjian F. A macroscopic description of multiphase flow in
-porous media involving spacetime evolution of fluid/fluid interface.
-Transport in Porous Media 1987;2:537--52.
-\url{https://doi.org/10.1007/BF00192154}.
-
 \leavevmode\hypertarget{ref-saffman1958penetration}{}%
-{[}35{]} Saffman PG, Taylor GI. The penetration of a fluid into a porous
+{[}36{]} Saffman PG, Taylor GI. The penetration of a fluid into a porous
 medium or hele-shaw cell containing a more viscous liquid. Proceedings
@@
 \leavevmode\hypertarget{ref-park1984two}{}%
-{[}36{]} Park C-W, Homsy G. Two-phase displacement in hele shaw cells:
+{[}37{]} Park C-W, Homsy G. Two-phase displacement in hele shaw cells:
 Theory. Journal of Fluid Mechanics 1984;139:291--308.
 
+\leavevmode\hypertarget{ref-jackson2015dynamic}{}%
+{[}38{]} Jackson S, Stevens D, Giddings D, Power H. Dynamic-wetting
+effects in finite-mobility-ratio hele-shaw flow. Physical Review E
+2015;92:023021.
+
 \leavevmode\hypertarget{ref-Olsson2005}{}%
-{[}37{]} Olsson E, Kreiss G. A conservative level set method for two
+{[}39{]} Olsson E, Kreiss G. A conservative level set method for two
 phase flow. Journal of Computational Physics 2005;210:225--46.
@@
 \leavevmode\hypertarget{ref-Olsson2007}{}%
-{[}38{]} Olsson E, Kreiss G, Zahedi S. A conservative level set method
+{[}40{]} Olsson E, Kreiss G, Zahedi S. A conservative level set method
 for two phase flow ii. Journal of Computational Physics
@@
 \leavevmode\hypertarget{ref-Popinet2018}{}%
-{[}39{]} Popinet S. Numerical models of surface tension. Annual Review
+{[}41{]} Popinet S. Numerical models of surface tension. Annual Review
 of Fluid Mechanics 2018;50:49--75.
@@
 \leavevmode\hypertarget{ref-Galusinski2008}{}%
-{[}40{]} Galusinski C, Vigneaux P. On stability condition for bifluid
+{[}42{]} Galusinski C, Vigneaux P. On stability condition for bifluid
 flows with surface tension: Application to microfluidics. Journal of

### 2020-04-01 draft.tex
Updates from Overleaf
minor

@@
 \hypersetup{
-            pdftitle={Numerical investigation of two-phase flow in high-permeability porous media Effect of permeability variation on traction terms},
+            pdftitle={Numerical investigation of two-phase flow in high-permeability porous media Effect of permeability variation on the surface force between phases},
             pdfauthor={Maxime Cochennec¹; Hossein Davarzani¹* (h.davarzani@brgm.fr); Yohan Davit²; Ioannis Ignatiadis¹; Michel Quintard²},
@@
 \title{Numerical investigation of two-phase flow in high-permeability porous
-media Effect of permeability variation on traction terms}
+media Effect of permeability variation on the surface force between
+phases}
 \author{Maxime Cochennec¹ \and Hossein Davarzani¹* (h.davarzani@brgm.fr) \and Yohan Davit² \and Ioannis Ignatiadis¹ \and Michel Quintard²}
@@
 
-In this paper we study the effects of varying permeability on traction
-terms exerted at the different boundaries for two-phase flow in
-Hele-Shaw cell. The traction terms, or surface force, are. surface
-integrals that are necessary to close the continuous macroscopic
-momentum balance equations for two-phase flow in porous media. These
-terms are usually modeled from experimental results or obtained by
-solving a closure problem (a set of boundary value problems to solve on
-a representative periodic unit-cell). Here these terms are calculated
-directly by means of direct numerical simulations, and the value of
-these terms on each interface is studied as a function of the varying
-permeability and capillary number. We focus on two-phase displacement in
-high-permeability porous media for which the characteristic flow regimes
-implies a non-negligible extent of the fluid-fluid interface and thus;
+The macroscopic description of two-phase flow in porous media requires
+modeling the force of interaction between the phases, i.e.~the two
+fluids and the solid phase. The flow regimes specific to high
+permeability porous media are characterized by a non-negligible extent
+of the interface between the two fluids, compared to the fluid-solid
+interface. This argues for the systematic introduction of the
+fluid-fluid interaction terms, the traction, in the macroscopic momentum
+equations. However, we do not know whether these terms are always
+non-negligible for this type of flow regime, for example in cases where
+the friction with the solid becomes very large. Here we show that the
+share of the traction between the fluids in the overall pressure drop
+increases when the flow is more confined, and thus the friction with the
+solid phase increases. The idea of solving modified 2D-Stokes equations
+with a supplementary Darcy term allowed us to modify the intrinsic
+permeability of the structure without having to modify the geometry per
+se. Our results demonstrate that the absolute permeability of the
+structure does not prejudge the importance of the traction between
+fluids in the total pressure drop as long as the flow regime is
+characteristic of flow in high permeability porous media.
 
@@
 see also
-{[}{[}\protect\hyperlink{ref-tung1988hydrodynamic}{33}{]};iliuta2005modelling{]},
+{[}\protect\hyperlink{ref-tung1988hydrodynamic}{33},\protect\hyperlink{ref-iliuta2005modelling}{34}{]},
 gives good results when compared with experimental results
-{[}\protect\hyperlink{ref-wang2013modelling}{34}{]}.
+{[}\protect\hyperlink{ref-wang2013modelling}{35}{]}.
 
@@
 \subsection{Outline of the study}\label{outline-of-the-study}}
+
+Up to this point we have seen that the macroscopic momentum equations
+need the modeling of the force of interaction terms between phases. The
+interaction between fluids is neglected in the traditional model used
+for the study of subsurface two-phase flow, while it has been
+experimentally observed that these terms are important when studying
+two-phase flow in high permeability porous media. However, it is not
+known whether these terms are always important when the extent of the
+interface between the fluids is important. For example, do they become
+negligible when friction with the solid phase increases? To what extent?
+To answer this question, we performed direct numerical simulations of
+two-phase flow in a Hele-Shaw cell as a model high permeability porous
+medium. We solved a coupled Level-Set and Stokes equations with an
+additional Darcy-term in the momentum equation that pertain for the
+absolute permeability of the flow between two plates. Thus, the friction
+due to the solid walls is increased by decreasing the aperture between
+the cell plates. This paper discusses the value of the interaction terms
+as a function of the aperture between the plates, i.e.~the absolute
+permeability of the medium, and the capillary number.
 
@@
 macroscale from the microscale is to use averaged quantities to express
-the boundary intergals terms that arising because of the averaging
-theorem application {[}\protect\hyperlink{ref-Whitaker2013}{35}{]}.
+the boundary integrals terms that arising because of the averaging
+theorem application {[}\protect\hyperlink{ref-Whitaker2013}{36}{]}.
 Remember that these closure relations are not presented here but rather
@@
 \[
-u=-\frac{1}{2\mu}\frac{dp}{dx}z(z-h).
+w=-\frac{1}{2\mu}\frac{dp}{dx}z(z-h).
 \] \{\#eq:poiseuilleProfile\}
@@
 permeability is given by \(h^{2}/12\)
-{[}\protect\hyperlink{ref-saffman1958penetration}{36}{]}.
+{[}\protect\hyperlink{ref-saffman1958penetration}{37}{]}.
 
@@
 the curvature in the Hele-Shaw plane
-{[}\protect\hyperlink{ref-saffman1958penetration}{36}{]}. The
+{[}\protect\hyperlink{ref-saffman1958penetration}{37}{]}. The
 +\textbf{???} was completed in
-{[}\protect\hyperlink{ref-park1984two}{37}{]} by an additional term that
+{[}\protect\hyperlink{ref-park1984two}{38}{]} by an additional term that
 pertains for dynamic film formation and a corrective \(\pi/4\) factor
@@
 dynamic film formation along the plates is an important phenomenom
-regarding the flow {[}\protect\hyperlink{ref-jackson2015dynamic}{38}{]},
+regarding the flow {[}\protect\hyperlink{ref-jackson2015dynamic}{39}{]},
 this requires dedicated study and is beyond the scope of this article.
@@
 respectively
-{[}\protect\hyperlink{ref-Olsson2005}{39},\protect\hyperlink{ref-Olsson2007}{40}{]}.
+{[}\protect\hyperlink{ref-Olsson2005}{40},\protect\hyperlink{ref-Olsson2007}{41}{]}.
 
@@
 object of a large body of work
-{[}\protect\hyperlink{ref-Popinet2018}{41}{]}, this term is defined here
+{[}\protect\hyperlink{ref-Popinet2018}{42}{]}, this term is defined here
 as
@@
 and \(\delta(\phi)\) is the Dirac delta function localized on the
-interface {[}\protect\hyperlink{ref-Galusinski2008}{42}{]}. In this
+interface {[}\protect\hyperlink{ref-Galusinski2008}{43}{]}. In this
 framework the unit normal vector as well as the mean curvature are
@@
 respectively, where the \(\pi/4\) correction was derived in
-{[}\protect\hyperlink{ref-park1984two}{37}{]} and
+{[}\protect\hyperlink{ref-park1984two}{38}{]} and
 \(\kappa_{\perp}=\frac{2}{h}\) is the out-of-plane curvature as
@@
 Let us now sum up all the traction terms and answer the question of the
-share of traction exerted on the fluid-fluid boundaries in total
-traction. The ratio \(T_{int}^x/T_{s}^x\) is given as a function of the
-dimensionless aperture between the Hele-Shaw plates in
-\cref{fig:ratioDrag}. The traction exerted on the fluid-fluid boundary
-reaches between 10 and 50\% of the traction exerted on the solid-fluid
-boundaries. Interestingly, the ratio varies non-monotonically with
-\(h^*\), the minimum is reached for the smallest capillary numbers and a
-medium range of the aperture values.
-
-Variations with aperture between plates are amplified as the capillary
-number decreases. Thus, it is for the lowest capillary number tested
-that we obtain both the maximum and minimu m values of the ratio
-\(T_{int}^x/T_{s}^x\).
+share of traction exerted on the fluid-fluid boundary in total traction.
+The traction exerted on the fluid-fluid boundary reaches between 10 and
+50\% of the traction exerted on the solid-fluid boundaries, as can be
+seen in \cref{fig:ratioDrag}, on which we plot the ratio
+\(T_int^x/T_s^x\) as a function of the dimensionless aperture between
+the cell plates. Interestingly, the ratio varies non-monotonically with
+\(h^*\), the minimum is reached for the smallest tested capillary
+numbers and a medium range of the aperture values. Logically the
+traction at the interface between the fluids is important when the
+aperture between the plates is maximum (the solution tends towards the
+2D solution). For a moderate aperture, the proportion of traction at the
+fluid-fluid interface decreases then rises again as the opening and thus
+the absolute permeability of the cell decreases.
+
+The variation of the ratio strongly depends on the capillary number and
+that the smaller the capillary number, the smaller the traction at the
+fluid interface. This is clearly visible in \cref{fig:ratioDragCa} on
+which we plot the ratio \(T_int^x/T_s^x\) as a function of the capillary
+number for different values of the opening between the plates.
 
@@
 
+\begin{figure}
+\hypertarget{fig:ratioDragCa}{%
+\centering
+\includegraphics{figures/pdf/ratioDragCa.pdf}
+\caption{Ratio of traction (\(x\)-component) on fluid-fluid interface
+over traction on solid-fluid interfaces in UC3 as a function of the
+capillary number and for different apertures at
+steady-state.}\label{fig:ratioDragCa}
+}
+\end{figure}
+
 \hypertarget{conclusion}{%
@@
 
-In this work, we investigated the effect of varied permeability on
-cocurrent two-phase flow in a Hele-Shaw cell. We focused on the value of
-the traction terms, that are necessary to close the macroscopic flow
-equations. The idea of solving modified 2D-Stokes equations with a
-supplementary Darcy term allowed us to modify the intrinsic permeability
-of the structure without having to modify the geometry per se. The range
-of intrinsic permeability values investigated is typical of
-high-permeability porous media, for which it is known that the flow
-regimes are characterized by the large extent of the fluid-fluid
-interface (e.g.~film flow). With direct numerical simulations we found
-that
+We investigated the effect of varied permeability on cocurrent two-phase
+flow in a Hele-Shaw cell. We focused on the value of the traction terms,
+that are necessary to close the macroscopic flow equations. The idea of
+solving modified 2D-Stokes equations with a supplementary Darcy term
+allowed us to modify the intrinsic permeability of the structure without
+having to modify the geometry per se. The range of intrinsic
+permeability values investigated is typical of high-permeability porous
+media, for which it is known that the flow regimes are characterized by
+the large extent of the fluid-fluid interface (e.g.~film flow). With
+direct numerical simulations we found that
 
@@
 to precisely locate the interface, even in the case of very thin film
-flow, and to carefully analyze the choice of parameters in Eq.
-{[}eq:advecPhi{]}.
+flow, and to carefully analyze the choice of parameters in the transport
+equation.
 
@@
 
+\leavevmode\hypertarget{ref-iliuta2005modelling}{}%
+{[}34{]} Iliuta I, Larachi F. Modelling the hydrodynamics of gas-liquid
+packed beds via slit models: A review. International Journal of Chemical
+Reactor Engineering 2005;3.
+
 \leavevmode\hypertarget{ref-wang2013modelling}{}%
-{[}34{]} Wang Y, Chen J, Larachi F. Modelling and simulation of
+{[}35{]} Wang Y, Chen J, Larachi F. Modelling and simulation of
 trickle-bed reactors using computational fluid dynamics: A
@@
 \leavevmode\hypertarget{ref-Whitaker2013}{}%
-{[}35{]} Whitaker S. The method of volume averaging. vol. 13. Springer
+{[}36{]} Whitaker S. The method of volume averaging. vol. 13. Springer
 Science \& Business Media; 2013.
@@
 \leavevmode\hypertarget{ref-saffman1958penetration}{}%
-{[}36{]} Saffman PG, Taylor GI. The penetration of a fluid into a porous
+{[}37{]} Saffman PG, Taylor GI. The penetration of a fluid into a porous
 medium or hele-shaw cell containing a more viscous liquid. Proceedings
@@
 \leavevmode\hypertarget{ref-park1984two}{}%
-{[}37{]} Park C-W, Homsy G. Two-phase displacement in hele shaw cells:
+{[}38{]} Park C-W, Homsy G. Two-phase displacement in hele shaw cells:
 Theory. Journal of Fluid Mechanics 1984;139:291--308.
@@
 \leavevmode\hypertarget{ref-jackson2015dynamic}{}%
-{[}38{]} Jackson S, Stevens D, Giddings D, Power H. Dynamic-wetting
+{[}39{]} Jackson S, Stevens D, Giddings D, Power H. Dynamic-wetting
 effects in finite-mobility-ratio hele-shaw flow. Physical Review E
@@
 \leavevmode\hypertarget{ref-Olsson2005}{}%
-{[}39{]} Olsson E, Kreiss G. A conservative level set method for two
+{[}40{]} Olsson E, Kreiss G. A conservative level set method for two
 phase flow. Journal of Computational Physics 2005;210:225--46.
@@
 \leavevmode\hypertarget{ref-Olsson2007}{}%
-{[}40{]} Olsson E, Kreiss G, Zahedi S. A conservative level set method
+{[}41{]} Olsson E, Kreiss G, Zahedi S. A conservative level set method
 for two phase flow ii. Journal of Computational Physics
@@
 \leavevmode\hypertarget{ref-Popinet2018}{}%
-{[}41{]} Popinet S. Numerical models of surface tension. Annual Review
+{[}42{]} Popinet S. Numerical models of surface tension. Annual Review
 of Fluid Mechanics 2018;50:49--75.
@@
 \leavevmode\hypertarget{ref-Galusinski2008}{}%
-{[}42{]} Galusinski C, Vigneaux P. On stability condition for bifluid
+{[}43{]} Galusinski C, Vigneaux P. On stability condition for bifluid
 flows with surface tension: Application to microfluidics. Journal of

### 2020-04-11 article1_cochennec.tex
merge all branches

%% LyX 2.3.4.2 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[12pt,english,review,authoryear,times]{elsarticle}
\usepackage{amsmath}
\usepackage{amsbsy}
\usepackage{fontspec}
\usepackage{unicode-math}
\setmainfont[Mapping=tex-text]{Georgia}
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{4}
\usepackage{booktabs}
\usepackage{setspace}
\usepackage{esint}
\doublespacing

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{lineno,hyperref}
\modulolinenumbers[5]
\usepackage{xcolor}
\usepackage{soul}
\setmainfont{EB Garamond}
\graphicspath{{figures/pdflatex/}{../figures/pdflatex/}}

\makeatother

\usepackage{polyglossia}
\setdefaultlanguage[variant=american]{english}
\begin{document}

\begin{frontmatter}{}

\title{Numerical investigation of two-phase flow in high-permeability porous
media: Effect of permeability variation on the surface force between
phases}

\author{Maxime Cochennec\fnref{myfootnote,myfootnote1}}

\author{Hossein Davarzani\fnref{myfootnote}}

\author{Yohan Davit\fnref{myfootnote1}}

\author{Ioannis Igniatiadis\fnref{myfootnote}}

\author{Michel Quintard\fnref{myfootnote1}}

\fntext[myfootnote]{French Geological Survey}

\fntext[myfootnote1]{Institut de Mécanique des Fluides de Toulouse}
\begin{abstract}
The macroscopic description of two-phase flow in porous media requires
modeling the force of interaction between the phases, i.e. the two
fluids and the solid phase. The flow regimes specific to high permeability
porous media are characterized by a non-negligible extent of the interface
between the two fluids, compared to the fluid-solid interface. This
argues for the systematic introduction of the fluid-fluid interaction
terms, i.e. surface integrals also called traction, in the macroscopic
momentum equations. However, we do not know whether these terms are
always non-negligible for this type of flow regime, for example in
cases where the friction with the solid becomes very large. Here we
assess that the share of the traction between the fluids in the overall
pressure drop increases when the flow is more confined, and thus the
friction with the solid phase increases. The idea of solving modified
2D-Stokes equations with a supplementary Darcy term allowed us to
modify the absolute permeability of the structure without having to
modify the geometry per se. Our results demonstrate that the absolute
permeability of the structure does not prejudge the importance of
the traction between fluids in the total pressure drop as long as
the flow regime is characteristic of flow in high permeability porous
media.
\end{abstract}

\end{frontmatter}{}

\linenumbers

\section{Introduction}

\subsection{Two-phase flow in high-permeability porous media}

An accurate description of two-phase flow in high-permeability porous
media is of major importance for several practical applications. One
can mention, among others, soil remediation in gravely soils \citep{fetter2017contaminant},
nuclear safety \citep{clavier2017modeling} or hydrodynamic of catalytic
fixed bed reactors \citep{Santos1991}. However, most of the literature
is dedicated to two-phase flow in low-permeability porous media.

Due to the larger pore size, two-phase flow in high-permeability porous
media (hereafter, high-permeability porous media refers to media for
which the characteristic particle diameter is about one millimeter
and above) results in a complex interaction between capillary, gravity
and viscous forces \citep{davit2018one}. For low-permeability porous
media, the flow is usually dominated by surface tension force, i.e.
the capillary number is low, usually inferior to $10^{-3}$, and the
fluid repartition pattern is well described as two independent flow
streams separate by a multitude of stable meniscus at steady state,
as illustrated in Fig.~\ref{fig:fluidPatterns} (a) \citep{dullien2012porous}.
This regime is characterized by the small extent of the fluid-fluid
interface, as the fluid phases are segregated, with the wetting phase
flowing into the larger pores while the wetting phase occupies the
smaller pores. In contrast, for high permeability porous media, the
flow is no longer capillary-dominated and the viscous forces become
significant (also gravity and inertial effects may become important
if Bond and Reynolds numbers are high, respectively). Thus the fluids
repartition patterns can take two forms, either the non-wetting phase
is continuous, see as an example Fig.~\ref{fig:fluidPatterns} (b),
or is flowing as droplets or ganglia, as in Fig.~\ref{fig:fluidPatterns}
(c). For these two options, the wetting phase is flowing as a film
in contact with the solid and the non-wetting phase flows at the center
of the pores surrounded by the wetting phase. Strictly speaking, these
different regimes must be considered when attempting to describe two-phase
flows with continuous macroscopic equations. Indeed, it has been shown
several times that the overall flow properties depend on the flow
regimes \citep{Avraam1995a,armstrong2016beyond}. At first glance,
one would consider that the exchange terms between the fluids, through
their common surfaces, as negligible compared to their counterpart
between the fluid phases and the solid phase for surface-tension dominated
flow since the extent of the fluid-fluid interface is small. On the
other hand, and this is what we are interested in here, this is not
necessarily the case for regimes specific to flow in high-permeability
porous media for which the extent of the fluid-fluid interface is
large. This is important because, as discussed in the next section,
these exchange terms between phases are the basis of any attempt to
establish continuous relationships on a macroscopic scale starting
from the pore scale.

\begin{figure}
\begin{centering}
\input{figures/pdflatex/dessin.pdf_tex}
\par\end{centering}
\caption{Illustration of possible fluids dispatching in a 2D porous network
with solid phase in black, fluid 1, that stands for the non-wetting
phase (gray) and the wetting fluid (white), (a) the two fluids are
flowing in different channels separate by numerous meniscus (b) wetting
and non-wetting fluid are flowing together in most of the pores as
two continuous streams and (c) both fluids are flowing together in
most of the pores and the non-wetting phase is discontinuous - Adapted
from \citep{dullien2012porous}\label{fig:fluidPatterns}}
\end{figure}

\subsection{Continuous model}

Volume averaged microscopic equation of motion for two Newtonian fluids
$i,j=o,w$ reads

\begin{equation}
\rho_{i}\varepsilon_{i}\left(\frac{\partial\mathbf{U}_{i}}{\partial t}+\mathbf{U}_{i}\cdot\nabla\mathbf{U}_{i}\right)=-\varepsilon_{i}\nabla P_{i}+\varepsilon_{i}\rho_{i}\mathbf{g}+\varepsilon_{i}\nabla\cdot\boldsymbol{\tau}_{i}+(\mathbf{F}_{ij}+\mathbf{F}_{is}),\;i\neq j,\label{eq:averagedMomentum}
\end{equation}

where $\varepsilon_{i}=V_{i}/V$ is the volume fraction of the $i$-fluid
and $V$ is a representative elementary volume, $U_{i}$ is the intrinsic
average velocity of the $i$-fluid, $P_{i}$ is the intrinsic average
pressure of the $i$-fluid and $\boldsymbol{\tau}_{i}$ denotes the
average viscous stress tensor. The last two terms in the left hand
side denote the force of interaction per unite volume exerted by fluid
$i$ upon fluid $j$ and the force of interaction per unit volume
exerted by fluid $i$ upon the solid phase, respectively \citep{Kalaydjian1987}.

From there, there are two possible ways to adress Eq.~\ref{eq:averagedMomentum}
and the associated equations. In volume averaging litterature, one
seeks to express the terms of interaction between the phases with
averaged quantities and to recast equation Eq.~\ref{eq:averagedMomentum}
in a form similar to Darcy's law (equation for creeping saturated
single-phase flow in porous media). Indeed, by considering creeping
two-phase flow Eq.~\ref{eq:averagedMomentum} can be written as

\begin{equation}
0=-\varepsilon_{i}\nabla P_{i}+\varepsilon_{i}\rho_{i}\mathbf{g}+\frac{1}{V}\int_{A_{is}+A_{ij}}\mathbf{n}_{i}\cdot\left(-\mathbf{I}p_{i}+\mu_{i}\left(\nabla\mathbf{u}_{i}+\left(\nabla\mathbf{u}_{i}\right)^{T}\right)\right)\mathrm{d}A.\label{eq:averagedMomentumIntegral}
\end{equation}

where $p_{i}$ and $u_{i}$ are microscopic pressure and velocity
fields of fluid $i$, respectively. By expressing the surface integrals
in term of averaged quantities one can obtain the following momentum
balance equations,

\begin{equation}
\mathbf{U}_{i}=-\frac{1}{\mu_{i}}\mathbf{K}_{ii}^{*}\cdot(\nabla P_{i}-\rho_{i}\mathbf{g})-\frac{1}{\mu_{j}}\mathbf{K}_{ij}^{*}\cdot(\nabla P_{j}-\rho_{j}\mathbf{g}),\quad i,j=o,w\:\mathrm{and}\:i\neq j,\label{eq:darcyCross}
\end{equation}

in which $\mathbf{K}_{ij}^{*}$ are the coupled relative permeability
tensors that pertain to the interaction between the fluids \citep{Whitaker1986a,Lasseux1996}.

The reader is warned that Eq.~\ref{eq:darcyCross} has, to our knowledge,
never been used in practice to model two-phase subsurface flows. On
the contrary, the ubiquitous continuous model used to describe two-phase
flows in soils is based on a direct extension of the Darcy's equation
and the whole model, also known as Muskat equations \citep{wyckoff1936flow,muskat1938flow},
reads

\begin{subequations}\label{eq:muskatModel}

\begin{equation}
0=\frac{\partial\varepsilon S_{i}}{\partial t}+\nabla\cdot\mathbf{U}_{i},\quad i=o,w,
\end{equation}

\begin{equation}
\mathbf{U}_{i}=-\frac{1}{\mu_{i}}\mathbf{K}_{i}\cdot(\nabla P_{i}-\rho_{i}\mathbf{g}),\quad i=o,w,
\end{equation}

\begin{equation}
1=S_{w}+S_{o}
\end{equation}

\begin{equation}
\mathbf{K}_{w}=\mathbf{K}k_{rw}(S_{w}),\qquad\mathbf{K}_{o}=\mathbf{K}k_{ro}(S_{w}),
\end{equation}

\begin{equation}
P_{c}(S_{w})=P_{o}-P_{w}.
\end{equation}

\end{subequations}

where $\mathbf{K}$ is the absolute permeability tensor. The generalization
toward two-phase flows involves the introduction of the relative permeability
terms $k_{ri}$ which account for the division of the void space between
the fluids \citep{dullien2012porous}, thus the relative permeability
depends (non-linearly) only on the saturation. To close the set of
the macroscopic equations a constitutive relation between the macroscopic
pressure of each fluid has to be furnished. This relation is known
as the capillary pressure relation and, as for the relative permeabilities,
is supposed to depends non-linearly only on the saturation \citep{leverett1941capillary}.

A significant amount of work attempted to make improvements to the
Muskat equations (e.g. including moving contact line \citep{Kalaydjian1987,Hassanizadeh1993,barenblatt2003mathematical}
or take into account the trapped phases \citep{hilfer1998macroscopic}).
Here we would stress that the concept of relative permeabilities into
the Muskat equations is dedicated to independent flow pathways only
\citep{blunt2017multiphase} and the underlying assumption that the
two streams do not interfere with each other, a situation far from
the regimes previously identified for high permeability porous media.
Thus Eq.~\ref{eq:muskatModel}(b) does not take into account the
interaction between the fluids as opposed to Eq.~\ref{eq:darcyCross}.
If the former is used to the detriment of the latter, it is mainly
because the literature on subsurface flow is focused on oil recovery
applications, for which the permeability is very low and the independent
streams regime is a good approximation.

Another approach, widely used in the literature on two-phase flow
in Trickle Bed Reactor (TBR), is to implement Eq.~\ref{eq:averagedMomentum}
and provide constitutive relations for the interaction terms between
the phases. These relations are usually obtained through interpretation
of experimental data. Some results, focusing on relation for the interaction
between the fluid phases, are given in the next section as well as
results in literature for coupled permeabilities from the porous medium
approach.

\subsection{Modeling of the interaction between fluids}

Based on the generalized Darcy law with coupled terms Eq.~\ref{eq:darcyCross},
experimental works computed the coupled transport coefficient through
steady-state cocurent flow in sandpack with one fluid, and alternatively
the other, which is submitted to a null pressure gradient. This protocol
was used with oil and water in a cylindric sandpack \citep{zarcone1994determination}
and the authors found a negligible effect of the coupled permeabilities
in the overall flow. With the same protocol, with oil and water in
a 2D-sandpack \citet{Dullien1996} found that the coupled permeabilities
are important since they can contribute at best to 35\% of the effective
permeability. Alternatively, authors imposed \citep{ramakrishnan2015measurement}
a null displacement and measured the induced pressure drop, with air
and water in a Berea sandstone core they found that the coupled transport
coefficients must not be overlooked for intermediate saturations.
Several authors conducted experiments with two different set-ups as
proposed by \citet{Rose1988}, see for example \citet{bentsen1993use}
who made cocurrent and countercurrent experiments with water and oil
in a sandpack and found that coupled permeabilities reach, at least,
15\% of the effective permeability value. However, it was pointed
out that the saturation between the two sets of experiments can be
very different and therefore the computed relative permeabilities
can not be safely compared \citep{langaas2001numerical}. The effect
of the non-wetting phase connectivity on the transport parameters
was extensively studied in \citep{Avraam1995}, the authors performed
steady-state cocurrent two-phase flow in 2D-micro model experiments
and found that the contribution of the coupled permeabilities on the
flow is non-negligible and depend on the flow regimes.

Several numerical studies investigated the share of the coupled permeabilities
in the effective permeability. The seminal work of \citet{Rothman1990}
examined the question by conducting two-phase flow simulations in
simple geometries with the immiscible lattice-gas method. The author
found non-negligible participation of the coupled permeabilities by
applying the volume force alternatively on each fluid. Numerous authors
used the lattice-Boltzmann method, such as \citep{Li2005}, in which
the authors examined the value of the coupled permeabilities as a
function of the saturation in a 3D-sphere pack and they found results
in agreement with Rothman's results. \citet{Yiotis2007} used a lattice
Boltzmann method in 2D and 3D pore networks and found a non-wetting
apparent relative permeability greater than unity when the wetting
fluid is more viscous than the non-wetting fluid. This result is a
manifestation of the lubrification effect, also observed in experiments
\citep{odeh1959effect}, and arising because of strong hydraulic coupling
between the fluids. Recently, \citet{shams2018study} have used a
Volume Of Fluid method to study the transport coefficients of fluid
layers in non-circular capillary tubes. Based on analogy with a model
Couette flow the authors have derived simple relations that can predict
with good accuracy the transport coefficients, including the coupled
ones, of fluid layers. We can also refer to recent work in which a
whole analytical model is given for one-dimensional two-phase flow
in coarse non-consolidated porous media and which correctly predicts
the pressure loss in a debris bed \citep{clavier2017modeling}.

Several correlations from theorical analysis for the fluid-fluid force
interaction have been derived in hydrodynamic of TBR litterature.
These correlations usually relate the momentum exchange between the
fluids to the flow velocity and the medium properties with Ergun's
like relations. For example, \citet{Attou1999} found that the drag
between the fluids can be expressed as follows

\[
F_{lg}=A_{lg}U_{r}+B_{lg}U_{r}^{2},
\]

where the relative velocity is given by $U_{r}=U_{g}-\frac{S_{g}}{S_{l}}U_{l}$
and$A_{lg}$ and $B_{lg}$ are coefficients that depend on the volume
fraction of the gas phase and the medium properties. The subscripts
$g$ and $l$ denote the gas phase and the liquid phase, respectively.
This type of model with explicit account of interaction between fluids,
see also \citep{Tung1988,iliuta2005modelling}, gives good results
when compared with experimental results \citep{Wang2013}.

\subsection{Outline of the study}

Up to this point we have seen that the macroscopic momentum equations
need the modeling of the force of interaction terms between phases.
The interaction between fluids is neglected in the traditional model
used for the study of subsurface two-phase flow, while it has been
experimentally observed that these terms are important when studying
two-phase flow in high permeability porous media. However, it is not
known whether these terms are always important when the extent of
the interface between the fluids is important. For example, do they
become negligible when friction with the solid phase increases? To
what extent? To answer this question, we performed direct numerical
simulations of two-phase flow in a Hele-Shaw cell as a model high
permeability porous medium. We solved a coupled Level-Set and Stokes
equations with an additional Darcy-term in the momentum equation that
pertain for the absolute permeability of the flow between two plates.
Thus, the friction due to the solid walls is increased by decreasing
the aperture between the cell plates. This paper discusses the value
of the interaction terms as a function of the aperture between the
plates, i.e. the absolute permeability of the medium, and the capillary
number and it is organized as follows. The following section \ref{sec:Theorical-background}
aims to introduce theoretical background about averaged quantities
and two-phase displacement in a Hele-Shaw cell. In section \ref{sec:Direct-numerical-simulations}
we present the numerical methods used to solve the flow and the free
interface position. Finally, section \ref{sec:Results} is devoted
to the results, starting with the influence of the aperture between
the plates on the distribution of fluids and pressure fields, and
ending with the study of the traction terms.

\section{Theorical background\label{sec:Theorical-background}}

\subsection{Macroscopic quantities}

In the following, the superficial average of a physical quantity $\theta_{i}$
associated with the $i$-phase is given by

\begin{equation}
\left\langle \theta_{i}\right\rangle =\frac{1}{V}\int_{V_{i}}\theta_{i}\:\mathrm{d}V.
\end{equation}

Thus saturation and intrinsic average pressure for two fluids $i=w,o$
read

\[
S_{i}=\frac{1}{V}\int_{V}\chi_{i}\;\mathrm{d}V
\]

and

\[
\langle p_{i}\rangle^{i}=\frac{\int_{V}p\chi_{i}\;\mathrm{d}V}{\int_{V}\chi_{i}\;\mathrm{d}V}=P_{i},
\]

respectively. In these equations $V$ is the volume accessible to
the fluids, that is the pore volume, and $\chi_{i}$ is the phase
indicator of the $i$-fluid (scalar function that takes the value
1 in the fluid and 0 elsewhere). The quantity $\int_{V}\chi_{i}\;\mathrm{d}V$
is the volume fraction of the $i$-fluid that is $\varepsilon_{i}$
and the intrinsic average ant the superficial average are linked as
$\left\langle \theta_{i}\right\rangle =\varepsilon_{i}\left\langle \theta_{i}\right\rangle ^{i}$.

Then we can introduce the dynamic capillary pressure defined as the
difference of the intrinsic average pressures,

\begin{equation}
P_{c}=P_{o}-P_{w}.
\end{equation}

\subsection{Traction terms}

An essential step in order to derive some macroscopic law at macroscale
from the microscale is to use averaged quantities to express the exchange
surface integrals terms. Remember that these closure relations are
not presented here but rather we leave intact the surface integrals,
the averaged momentum balance equations Eq.~\ref{eq:averagedMomentumIntegral}
for two fluids $w$ and $o$, are recast as

\begin{subequations}

\begin{equation}
0=-\varepsilon_{w}\nabla P_{w}+\mathbf{T}_{wo}+\mathbf{T}_{ws},
\end{equation}

\begin{equation}
0=-\varepsilon_{o}\nabla P_{o}+\mathbf{T}_{ow}+\mathbf{T}_{os},
\end{equation}

\end{subequations}

where $\mathbf{T}_{ij}$ are the traction (name given to surface forces
in the Boundary Integral Methods literature in which these integrals
are central) exerted by fluid $i$ upon fluid $j$ and $\mathbf{T}_{is}$
the traction exerted by fluid $i$ upon the solid \citep{Kalaydjian1987}.
The traction terms are written as

\begin{equation}
\mathbf{T}_{ij}=\frac{1}{V}\int_{A_{ij}}\mathbf{n}_{ij}\cdot\left(-p_{i}\mathbf{I}+2\mu_{i}\mathbf{e}_{i}\right)\mathrm{d}A,\label{eq:tractionIntegral}
\end{equation}

where $\mathbf{n}_{ij}$ is the normal vector at the interface and
pointing toward the j-phase, $p_{i}$ the pointwise pressure of the
fluid $i$ and$\mathbf{e}_{i}=\frac{1}{2}\left(\nabla\mathbf{u}_{i}+\left(\nabla\mathbf{u}_{i}\right)^{T}\right)$
the rate-of-strain tensor for Newtonian fluids with $\mathbf{u}_{i}$
the pointwise velocity of fluid $i$.

\subsection{Theory of two-phase displacement in Hele-Shaw cell}

Displacement in a Hele-Shaw cell is a popular experimental set-up
used to reproduced 2D potential flow similar to flow in porous media
\citet{guyon1994hydrodynamique}. Two-phase flow into such devices,
as under study here, can be depicted as in Fig.~\ref{fig:modelHeleShaw}.
This represents a cocurrent flow of two fluids inside a Hele-Shaw
cell parallel to the $x-y$ plane, with the $z$-axis perpendicular
to the plates, and for which $h$ is the aperture between the plates
considered small in front of the other dimensions of the cell . A
solid obstacle of circular cross-section is sandwiched between the
two plates. Tangential (cell plane) stress, as well as stress arising
from the perpendicular confinement, need to be taking into account
to properly describe the flow. The important parameter here are the
aspect ratio $h/l$ where $l$ is some characteristic transverse length
(hereafter $l$ is the cell width as shown in Fig.~\ref{fig:modelHeleShaw})
and the capillary number.

\begin{figure}
\begin{centering}
\input{figures/pdflatex/dessin_courbure.pdf_tex}
\par\end{centering}
\caption{Illustation of confined cocurrent two-phase flow between two parallel
plates and around cylinder obstacles.\label{fig:modelHeleShaw}}
\end{figure}

To compute the traction exerted by the fluids upon the cell plates
we consider that the flow velocity is only oriented with the $x$-axis
and that it depends only on the $z$-coordinate, where its profile
follows the Poiseuille's law,

\[
u=-\frac{1}{2\mu}\frac{dp}{dx}z(z-h).
\]

Then the friction force exerted on the bottom plate is given by

\[
\mu\frac{\partial u}{\partial z}\mid_{z=0}=-\frac{h}{2}\frac{dp}{dx}.
\]

Thus, the traction $T_{i,pl}^{x}$ is obtained by multiplying the
friction force by two times the surface wetted by the fluid $i$.

Going beyond the approximation that the velocity depends only on the
$z$-coordinates we note that the average velocity taken over the
interval between the plates is written as follows

\[
\mathbf{U}(x,y,0)=-\frac{h^{2}}{12\mu}\nabla_{\parallel}p.
\]

There is therefore a direct analogy between the flow in a 2D porous
medium and the flow in a Hele-Shaw cell for which the permeability
is given by $h^{2}/12$ \citep{saffman1958penetration}. This allows
us in the following to modify the absolute permeability of the structure
without modifying its overall geometry, just by varying the aperture
between the plates.

Beside the interface curvature in the Hele-Shaw cell plane, the meniscus
in the perpendicular plane must be taken into account. Considering
it as a semicircle of radius $h/2$, the pressure jump at equilibrium
across the interface reads

\begin{equation}
\Delta p=\gamma\left(\frac{2}{h}+\kappa_{\parallel}\right),\label{eq:curvatureHeleShaw}
\end{equation}

where $\gamma$ denotes the surface tension and $\kappa_{\parallel}$
the curvature in the Hele-Shaw plane. Eq.~\ref{eq:curvatureHeleShaw}
was completed by \citet{park1984two} with an additional term that
pertains for dynamic film formation and a corrective $\pi/4$ factor
for the curvature in the cell plane. The reader is warnes that while
dynamic film formation along the plates is an important phenomenom
regarding the flow \citep{jackson2015dynamic}, this question requires
dedicated study and is beyond the scope of this article. Thus, we
consider in the following that the invading fluid is in contact with
the cell plates, i.e. presence of a triple line.-

\section{Direct numerical simulations\label{sec:Direct-numerical-simulations}}

\subsection{Equations}

\subsubsection{Level Set model}

We used the standard Level Set method to capture the moving free interface
between the fluids. The method is part of Eulerian methods which have
the particularity of easily reproducing topological changes of the
phases contrary to Lagrangian methods. As topological changes of the
fluids is not excluded here, this motivated the choice for this type
of method. In this framework the fluid phases are identified with
a level set (scalar) function that goes smoothly from 0 to 1 across
the fluid interface, which is implicitely defined as the iso-level
$\phi=0.5$. Transport of the level set function $\phi$ is governed
by

\begin{equation}
\frac{\partial\phi}{\partial t}+\nabla\cdot(\mathbf{u}\phi)=\tau\nabla\cdot\left(\psi\boldsymbol{\nabla}\phi-\phi(1-\phi)\frac{\boldsymbol{\nabla}\phi}{\vert\nabla\phi\vert}\right).\label{eq:advecPhi}
\end{equation}

In this equation, $\mathbf{u}$ is the velocity field and $\tau$
and $\psi$ are two numerical parameters that control the diffuse
interface thickness and the amount of reinitialization of $\phi$
function, respectively \citep{Olsson2007}.

\subsubsection{Flow equations}

Remember that we need to take into account tangential stress as well
perpendicular stress due to flow confinement, a full-3D resolution
of flow equations would be meaningful. In order to avoid solving such
resource-intensive simulations we rather resolved 2D flow equations
plus an additional term taking into account the perpendicular stress.
Velocity and pressure fields are obtained by solving 2D-modified Stokes
equations since the Reynolds number is assumed to be sufficiently
low so that the flow is creeping. The whole-domain formulation Stokes
and continuity equations read

\begin{subequations}

\begin{equation}
0=-\nabla p+\mu(\phi)\left(\nabla^{2}\mathbf{u}-k^{2}\mathbf{u}\right)+\mathbf{F}_{st},
\end{equation}

\begin{equation}
0=\nabla\cdot\mathbf{u},
\end{equation}

\end{subequations}

where $k=\frac{\sqrt{12}}{h}$ is an additional term that pertain
to a Darcy-term arising from depth-averaging of Stokes equation, as
described in the theoretical background section and $\mathbf{F}_{st}$
is the surface tension fore per unit volume. This term is defined
here as a purely normal contribution, i.e. constant surface tension
along the interface, and reads

\[
\mathbf{F}_{st}=\gamma\kappa\delta(\phi)\mathbf{n},
\]

where $\mathbf{n}$ denotes the unit normal to the interface and $\delta(\phi)$
is the Dirac delta function localized on the interface \citep{Sussman1994}.
In this framework the unit normal vector as well as the mean curvature
are obtained with the level set function $\phi$ as

\begin{equation}
\mathbf{n}=\frac{\boldsymbol{\nabla}\phi}{\vert\boldsymbol{\nabla}\phi\vert},
\end{equation}

\begin{equation}
\kappa=\frac{\pi}{4}\nabla\cdot\left(\frac{\boldsymbol{\nabla}\phi}{\vert\boldsymbol{\nabla}\phi\vert}\right)+\frac{2}{h},
\end{equation}

respectively, where we added the $\pi/4$ factor in the curvature
expression to be consistent with the correction derived by \citet{park1984two}.

The fluid density and the fluid viscosity vary smoothly from one fluid
to another

\[
\rho(\phi)=\rho_{o}+(\rho_{w}-\rho_{o})\phi,
\]

\[
\mu(\phi)=\mu_{o}+(\mu_{w}-\mu_{o})\phi.
\]

\subsubsection{Dimensionless equations and relevant dimensionless numbers}

If one define the following dimensionless variables

\[
\mathbf{u}=\mathbf{u}'\times U,\;p=p'\times\frac{\mu U}{l},\;\kappa=\kappa'\times l,\;\mathbf{x}=\mathbf{x}'\times l,
\]

where $U$ is a reference velocity, $\mu$ a reference viscosity and
$l$ a reference length, Stokes and continuity equations can be recast
as

\[
0=-\nabla'p'+\frac{\mu(\phi)}{\mu}\nabla'^{2}\mathbf{u}'+\frac{\gamma}{\mu U}\kappa'\nabla'\phi,
\]

\[
0=\nabla'\cdot\mathbf{u}'.
\]

One can notice the two following dimensionless numbers

\[
\bar{\mu}=\frac{\mu(\phi)}{\mu},\;Ca^{-1}=\frac{\gamma}{\mu U},
\]

the viscosity ratio and the capillary number, respectively, which
add up to the aspect ratio $h/l$.

In the following, the results are given as a function of

\[
Ca=\frac{(U_{o}+U_{w})\mu_{o}}{\gamma},\qquad U_{w}^{'}=\frac{U_{w}}{U_{o}+U_{w}}=f_{f},\qquad U_{o}^{'}=1-f_{f},
\]

where $f_{f}$ is called fractional flow, the reference viscosity
is the invading fluid viscosity ($\mu_{o}=2\times\mu_{w}=2\times10^{-3}$
Pa.s) and the reference velocity is the total velocity $U=U_{o}+U_{w}$.
The viscosity ratio $\bar{\mu}$ as well as the fractional flow $f_{f}$
are kept constant at 2 and 0.25, respectively, and the reference length
is the cell's width denoted $l=5\times10^{-4}$ m .

\subsection{Geometry and boundary conditions}

We study a model system consisting of an array of cylinders in a thin
square cuboid, resembling a Hele-Shaw-cell with obstacles, as shown
in Fig.~\ref{fig:model}. The two fluids are injected from the left
at a constant normal velocity. The outlet boundary condition for the
flow is a constant pressure. Initially the wetting fluid occupy the
entire geometry. The different boundary conditions are listed in \ref{tab:Boundary-conditions}.

\begin{figure}
\begin{centering}
\include{figures/pdflatex/model}
\par\end{centering}
\caption{The geometry used is an array of 5 cylinders in a thin square cuboid,
resembling a Hele-Shaw-cell where both fluids are injecting from left
to right, initially the square cuboïd are saturated with wetting fluid
(in blue), the length of one cell (UC) is one millimetre.\label{fig:model}}
\end{figure}

\begin{table}
\begin{centering}
\begin{tabular}{cccc}
\toprule 
Boundary & $\mathbf{u}$ & $p$ & $\ensuremath{\phi}$\tabularnewline
\midrule
\midrule 
$A_{ws}$ & 0 & - & $\mathbf{n}\cdot\mathbf{u}\phi=0$\tabularnewline
\midrule 
Outlet & - & 0 & $\mathbf{n}\cdot\boldsymbol{\nabla}\phi=0$\tabularnewline
\midrule 
Inlet $o$ & $U_{o}^{x}=\mathrm{cst}$ & - & 0\tabularnewline
\midrule 
Inlet $w$ & $U_{w}^{x}=\mathrm{cst}$ & - & 1\tabularnewline
\bottomrule
\end{tabular}
\par\end{centering}
\caption{Boundary conditions\label{tab:Boundary-conditions}}
\end{table}

\section{Results\label{sec:Results}}

Before we turn to the study of traction terms, we computed the absolute
permeability of our model, remember that it is expected to be smaller
than the $12/h^{2}$ value since the sandwiched cylinders contribute
to the pressure drop along with the cell's plates. The absolute permeability
of the model is varied by modifying the aperture between the two plates.
We used five different values for the aperture and the correspondence
in terms of absolute permeability, obtained from single-phase simulations,
is given in Tab.~\ref{tab:permeability}. When $l/h\rightarrow\infty$,
it is expected that the absolute permeability tends towards the permeability
obtained for purely 2D flow, this is observed here since the absolute
permeability for pure 2D case is $1.54\times10^{-4}$ darcy which
is very close to what we obtained for the largest aperture $h=2.5\times10^{-3}$
m. Thus, the range of absolute permeability values is typical of what
is encountered for high permeability porous media.

\begin{table}
\caption{Equivalence between the value of the cell plates aperture and the
absolute permeability of the model obtained either from the $12/h^{2}$
value for flow between two plates or from single-phase simulation.\label{tab:permeability}}

\centering{}%
\begin{tabular}{ccc}
\toprule 
$h^{*}=h/l$ & $h^{2}/12$ (darcy) & $K$ (darcy)\tabularnewline
\midrule
\midrule 
5 & $5.2\times10^{5}$ & $1.5\times10^{4}$\tabularnewline
\midrule 
0.5 & $5.2\times10^{3}$ & $2.8\times10^{3}$\tabularnewline
\midrule 
0.25 & $1.3\times10^{3}$ & $8.6\times10^{2}$\tabularnewline
\midrule 
0.125 & $3.3\times10^{2}$ & $2.3\times10^{2}$\tabularnewline
\midrule 
0.05 & $5.2\times10^{1}$ & $3.9\times10^{1}$\tabularnewline
\bottomrule
\end{tabular}
\end{table}

For the selected parameters, $f_{f}=0.25$ and $0.05<h^{*}<5$, we
observed only a film-flow regime without any phase rupture. The wetting
fluid saturation in one cuboid, the third one in the rest of the paper
unless otherwise specified, as a function of the dimensionless aperture
and for different capillary numbers is given in Fig.~\ref{fig:Saturation}
along with corresponding selected $\phi$-fields. Wetting fluid saturation
varies between 0.3 and 0.6 and increases as the aperture increases.
The capillary number also impacts saturation by promoting the hold-up
of the wetting fluid. As seen in the selected $\phi$-fields, the
smaller the capillary the flatter the interface. For a wetting fluid
layer thickness equivalent at the plumb end of the cylinder, the saturation
is logically higher.

\begin{figure}
\begin{centering}
\input{figures/pdflatex/saturation.tex}
\par\end{centering}
\caption{Wetting fluid saturation in UC3 as a function of the dimensionless
aperture between the Hele-Shaw plates and for different capillary
numbers at steasy-state\label{fig:Saturation}}

\end{figure}

Let us now take a closer look at the dynamic capillary pressure as
expressed in Eq.~ and the pressure fields. Here, as can be seen in
Fig.~\ref{fig:Pc}, the difference in intrinsic mean pressure of
the fluids is entirely controlled by the pressure jump caused by the
out-of-plane meniscus, even for the larger aperture, and does not
depends on the capillary number at all. How the pressure jump compared
to the pressure gradient across a cell? Based on the pressure fields
in Fig.~\ref{fig:pField} one can answer that for $Ca=1$ the pressure
gradient across the cell is larger than the pressure jump and, interestingly,
it becomes larger and larger than the pressure jump as the aperture
between the plates is reduced. It is quite different when the capillary
number decreases, and for $Ca=0.01$ combined with a large aperture
the pressure gradient across the cell is negligible in front of the
pressure jump due to surface tension effects. As the aperture decreases,
the pressure gradient increases and becomes comparable to the pressure
jump, but the pressure itsel is differently distributed in the cell
compared to higher capillary number cases with a higher pressure in
the wetting fluid.

\begin{figure}
\begin{centering}
\input{figures/pdflatex/Pc.tex}
\par\end{centering}
\caption{Dynamic capillary pressure in UC3 as a function of the dimensionless
aperture between the Hele-Shaw cell plates, for different capillary
numbers and pressure drop across the fluid-fluid from the out-of-plane
meniscus at steady-state.\label{fig:Pc}}

\end{figure}

\begin{figure}
\begin{centering}
\include{figures/pdflatex/pressureField}
\par\end{centering}
\caption{Pressure field in UC3 for two different capillary number and two different
dimensionless apertures, interface position as the iso-$\phi$=0.5
contour (white line) and isobar (black lines)\label{fig:pField}}

\end{figure}

We now turn to the study of the traction exerted upon the solid-fluid
boundaries $\mathbf{T}_{s}$, i.e. fluid $i$ upon the plates and
wetting fluid upon the cylinder, along with the traction at the fluid-fluid
interace $\mathbf{T}_{f}$. We focus on the magnitude of the traction
exerted at the fluid-fluid interface compared to the traction exerted
upon the solid boundaries. In the following we discuss only the $x$-component
of the traction vector $T^{x}$, i.e. aligned according to the inlet
velocity.

The traction exerted upont the plates by the fluids was obtained as
described in the theoretical backgound section. Traction on the cylinder
boundary was obtained by calculating the integral Eq.~\ref{eq:tractionIntegral}.
We computed the fluid-fluid traction $\mathbf{T}_{f}$ in two steps.
The first step concerns the viscous part of the stress tensor, which
was calculated directly from the velocity gradients available on the
interface contour. The second step concerns the pressure part, which
was obtained by applying the divergence theorem on each phase into
a cell (UC3).

We assessed that the traction exerted on the fluid-fluid boundary
reaches between 10 and 50\% of the traction exerted upon the solid-fluid
boundaries, as can be seen in Fig.~\ref{fig:ratioTraction}, on which
we plot the ratio $T_{f}^{x}/T_{s}^{x}$ as a function of the dimensionless
aperture between the cell plates. Interestingly, the ratio varies
non-monotonically with $h$ and the maximum value is obtained for
the smallest aperture which then decreases when the aperture increases.
The capillary number accentuates the decrease of $T_{f}^{x}$ compared
to $T_{s}^{x}$ when it decreases, as can be seen in Fig.~\ref{fig:ratioTractionCa}.

\begin{figure}
\begin{centering}
\input{figures/pdflatex/ratioDrag.tex}
\par\end{centering}
\caption{Ratio of traction ($x$-component) on fluid-fluid interface over traction
on solid-fluid interfaces in UC3 as a function of the dimensionless
aperture between the Hele-Shaw plates and for different capillary
numbers at steady-state.\label{fig:ratioTraction}}
\end{figure}

\begin{figure}
\begin{centering}
\input{figures/pdflatex/ratioDragCa.tex}
\par\end{centering}
\caption{Ratio of traction ($x$-component) on fluid-fluid interface over traction
on solid-fluid interfaces in UC3 as a function of the capillary number
and for different apertures at steady-state.\label{fig:ratioTractionCa}}
\end{figure}

Consider now the total traction on the fluid-fluid interface and the
total traction exerted upon the solid boundaries plotted separately
against the dimensionless aperture in Fig.~\ref{fig:tractionTotal}.
Between $h^{*}=0.5$ and $h^{*}=0.05$ the traction exerted upon the
solid scale as $h^{-2}$, in contrast the traction at the fluid-fluid
interface increases faster with $h$ over the same interval. This
explains why $T_{f}^{x}/T_{s}^{x}$ increases as the aperture between
the plates decreases. To go further we studied separately the pressure
part and the viscous contribution in Eq.\ref{eq:tractionIntegral}
at the fluid-fluid interface as a function of the dimensionless aperture,
see Fig.~\ref{fig:tractionFluid}. For $h^{*}<0.5$ the pressure
part largely dominates the viscous contribution and the evolution
of the traction is therefore controlled by it. The pressure part scales
as $h^{-2}$ for $Ca=1$ but increases faster when the capillary number
decreases.

\begin{figure}
\begin{centering}
\input{figures/pdflatex/tractionTotal.tex}
\par\end{centering}
\caption{$x$-component of the traction exerted upon fluid-fluid interface
and upon the solid-fluid interface in UC3 as a function of the dimensionless
aperture between the Hele-Shaw plates and for different capillary
numbers at steady-state.\label{fig:tractionTotal}}
\end{figure}

\begin{figure}
\begin{centering}
\input{figures/pdflatex/multiplotTractionFluidInterface.tex}
\par\end{centering}
\caption{Viscous and pressure part of $x$-component of the traction exerted
upon the fluid-fluid interface in UC3 as a function of the dimensionless
aperture between the Hele-Shaw plates at steady-state for different
capillary numbers.\label{fig:tractionFluid}}
\end{figure}

Due to the proximity of the plates, the frictional forces are high,
and they become higher and higher as they get closer. However, the
traction exerted at the fluid/fluid interface does not become negligible
compared to that exerted by the fluid upon the solid. Clearly, what
is happening here is that the pressure increases due to the combination
of constant inlet velocity and lower absolute permeability have a
stronger effect on the traction term exerted at the fluid-fluid boundary. -

\newpage{}

\bibliographystyle{elsarticle-harv}
\bibliography{biblio}

\newpage{}

\appendix

\section{Model validation}

The code is validated by comparison with a Boundary-Element Method,
which relies on a surface discretization of the interface and a pseudo-analytical
formulation in the bulk of the phases. This allows us to precisely
locate the interface, even in the case of very thin film flow, and
to carefully analyze the choice of parameters in Eq.~\ref{eq:advecPhi}.
\end{document}

### 2020-07-11 article.tex
last update

%% LyX 2.3.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english,5p]{elsarticle}
\usepackage{amsmath}
\usepackage{fontspec}
\usepackage{unicode-math}
\setmainfont[Mapping=tex-text]{Georgia}
\usepackage{xcolor}
\usepackage{pdfcolmk}
\usepackage{array}
\usepackage{float}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{setspace}
\PassOptionsToPackage{normalem}{ulem}
\usepackage{ulem}
\doublespacing

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}
\providecolor{lyxadded}{rgb}{0,0,1}
\providecolor{lyxdeleted}{rgb}{1,0,0}
%% Change tracking with ulem
\DeclareRobustCommand{\lyxadded}[3]{{\color{lyxadded}{}#3}}
\DeclareRobustCommand{\lyxdeleted}[3]{{\color{lyxdeleted}\lyxsout{#3}}}
\DeclareRobustCommand{\lyxsout}[1]{\ifx\\#1\else\sout{#1}\fi}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{xcolor}
\usepackage{lineno,hyperref}
\modulolinenumbers[5]
\usepackage{fontspec}
\usepackage{unicode-math}

\setmathfont{STIX2Math}[
Extension={.otf},
Path=STIX2fonts/,
Scale=1.1]
\setmainfont{STIX2Text}[
Extension={.otf},
Path=STIX2fonts/,
UprightFont={*-Regular},
BoldFont={*-Bold},
ItalicFont={*-Italic},
BoldItalicFont={*-BoldItalic}]

\makeatother

\usepackage{babel}
\begin{document}

\begin{frontmatter}{}

\title{Numerical investigation of two-phase flows in highly-permeable porous
media: Effect of the permeability on the drag force between fluid
phases}

\author{Maxime Cochennec, Hossein Davarzani, Yoahn Davit, Michel Quintard}
\begin{abstract}
The macroscopic description of two-phase flows in porous media requires
accurate modelling of the drag forces between the two fluids and the
solid phase. In standard porous media, where capillarity is often
dominant, the fluid-solid interactions are well-known and the fluid-fluid
drag force is treated in a similar way as the drag between fluids
and solid in the momentum transport equation. Two-phase flows in highly
permeable porous media, however, are often characterized by a larger
area of the interface between the two fluids and the development of
thin films. In such cases, the fluid-fluid interaction is not necessarily
negligible and may play an important role in the momentum transport
equations. Here, we use computational methods to study two-phase flows
in a microfluidic device made of an array of cylinders squeezed between
two plates in a Hele-Shaw flow cell. The idea is to keep the geometry
in the cell plane unmodified whereas the aperture $h$ between the
plates is changed to explore different ranges of permeability. This
is done by solving depth-averaged flow equations, taking advantage
of the quasi-planar nature of the flow. We reproduce a film-flow regime
and show that the fluid-fluid drag is non-negligible as it reaches
between 5\% and 60\% of the solid-fluid drag. XXXOur results demonstrate
that the fluid-fluid drag force should not be neglected in momentum
transport equations and behaves differently than the drag upon solids,
confirming that a no-flow condition between fluids is not suitable.
This is an important part of the physics that has strong implications
relevant for upscaling and modelling two-phase flows in microfluidic
devices or highly permeable porous media.
\end{abstract}

\end{frontmatter}{}

\section{Introduction\label{sec:Introduction}}

An accurate description of two-phase flows in high-perme\lyxadded{Maxime}{Wed Jul  8 13:49:08 2020}{}\lyxadded{Maxime}{Wed Jul  8 13:49:08 2020}{\-}ability
porous media is of major importance in several practical applications.
This includes soil remediation in sandy or gravely soils \citep{fetter2017contaminant},
nuclear safety \citep{clavier2017modeling}, or catalytic fixed bed
reactors \citep{Santos1991}. However, most of the literature on two-phase
flows in porous media is focused on low-permeability porous media.
For low-permeability porous media in the limit of creeping flows,
surface tension forces often dominate the flow; thus the capillary,
Bond, and Weber numbers are low. In that case, the fluid repartition
is well described as two independent flow paths for each phase \citep{dullien2012porous,blunt2017multiphase}.
The two fluids are segregated, the non-wetting fluid flows into the
larger pores, whereas the wetting fluid occupies the smaller pores.
One of the consequences is that the area of the fluid-fluid interface
is small (Fig.~\ref{fig:fluidPatterns} (a)) and there is little
drag between the fluid phases. In contrast, for high-permeability
porous media, the flow is the result of a complex interaction between
capillary, gravity, viscous, and inertial forces \citep{davit2018one}.
Capillary effects may no longer dominate and the capillary, Bond,
and Weber numbers may be large. The distribution of fluids in the
pore space can be schematically decomposed in two modes, even though
the reality is often a lot more complex. Either the non-wetting fluid
is continuous and flows in the center of the pores surrounded by the
wetting fluid flowing as a thin film in contact with the solid (Fig.~\ref{fig:fluidPatterns}
(b)), or the non-wetting fluid is discontinuous and flows in the center
of the pores in the form of droplets or ganglias (Fig.~\ref{fig:fluidPatterns}
(c)). In both cases, the surface area between the fluids is large
and the drag forces between the fluids are not negligible compared
with the solid-fluids drag forces. This is in strong contrast with
capillarity-dominated flow and it is important because, as discussed
in the following, modelling of the drag forces is the basis of any
attempt to establish continuous relationships on a macroscopic scale
starting from the pore scale.

\begin{figure}
\begin{centering}
\includegraphics[scale=0.6]{figures/pdf/INTRODUCTION_fluidRepartitionSketch}
\par\end{centering}
\caption{Schematics of possible distributions of fluids in a 2D porous network
with solid phase in black, the non-wetting fluid (fluid 1) in light
grey, and the wetting fluid (fluid 2) in white, (a) the two fluids
are flowing in different channels separated by numerous meniscus and
the fluid-fluid interface extent is small, (b) the wetting and non-wetting
fluids are flowing together in most of the pores as two continuous
streams and (c) both fluids are flowing together in most of the pores
and the non-wetting phase is discontinuous. - Adapted from \citep{dullien2012porous}\label{fig:fluidPatterns}}
\end{figure}

Models used to describe two-phase flows in porous media are often
based on a direct extension of Darcy's equations for one-phase flow
\citep{wyckoff1936flow,muskat1938flow}. This generalization depends
on the introduction of relative permeabilities that essentially account
for the division of the available void space between the fluids; thus
each fluid phase acts as a supplementary solid regarding the other
one and no interaction between the phases is taken into account \citep{dullien2012porous,blunt2017multiphase}.
As a consequence, it is commonly assumed that the relative permeability
only depends on the saturation \citep{brooks1964hydrau,van1980closed}
and we know that this is not always accurate as relative permeabilities
may also depend on the capillary number \citep{Li2005}, the flow
regime \citep{Avraam1995} or the viscosity ratio \citep{yuster1951theoretical,ehrlich1993viscous,Yiotis2007}.
Since the early 1980s, numerous works have aimed at improving the
generalized Darcy equations on a sound physical basis. Using upscaling
techniques, several authors proposed additional coupling terms that
correspond to stresses at the fluid-fluid interface and yield coupling
permeability tensors \citep{MARLE1982643,auriault1986remarques,Whitaker1986a,Lasseux1996}.
The importance of these coupling terms in the overall flow process
is not clear \citep{ayub1999interfacial}. These additional terms
can be calculated analytically in a two-phase annular cocurrent flow
in a cylindrical capillary tube, and are of the same order as the
dominant relative permeabilities \citep{bacri1990modele}. However,
other studies considered configurations for which the surface between
fluids was smaller and concluded that coupling terms should not be
as important \citep{scott1953explanation,rakotomalala1995viscous}.
\citet{zarcone1994determination}, \citet{Dullien1996} and \citet{ramakrishnan2015measurement}\lyxadded{Maxime}{Fri Jun 26 12:11:51 2020}{
}directly measured the coupling permeability terms in natural media
by performing steady-state cocurrent two-phase flows. \citet{Rose1988}
proposed to indirectly measure the coupling relative permeability
terms by performing two different types of experiments. This technique
was also used for both cocurrent and countercurrent experiments \citep{bourbiaux1990experimental,bentsen1993use}.
Each of these authors, except Zarcone and Lenormand, found that the
coupling relative permeabilities are significant. Recently, \citet{clavier2017modeling}
performed experiments of inertial two-phase flows in coarse non-consolidated
porous media and proposed constitutive models. However, results suffer
from a major shortcoming. Indeed, it is impossible to know which type
of flow regime dominates at the pore scale and thus the exact link
between the physics at the pore-scale and the macroscale model.

Micromodels can be used to better understand two-phase flows in porous
media \citep{karadimitriou2012review}, e.g. transitions between the
flow regimes or the onset and development of displacement instabilities
\citep{lenormand1988numerical,zhang2011influence}. Therefore, micromodels
allow overcoming the major issue we have just risen. Such apparatus
were employed to measure the relative coupling permeabilites for different
flow regimes \citep{Avraam1995} or study the impact of the fluid-fluid
drag on the flow characteristics \citep{Heshmati2018,Roman2019}.
\citet{Rothman1990} used numerical simulations in a 2D micromodel
geometry and found that coupling permeabilities are comparable in
magnitude with the case of the annular flow in a capillary tube. Fig.~\ref{fig:comparison_Kr_litterature}
shows Rothman's results along with some of the previously mentioned
results on relative coupling permeabilities. Hele-Shaw cells are one
of the simplest example of micromodels as they consist of two parallel
plates in between which the fluids flow. Governing equations for the
flow in such cells are similar to Darcy's equation, and thus the permeability
can be easily modified by increasing or decreasing the aperture between
the plates \citep{saffman1958penetration}. Recently, this characteristic
has been used to investigate the flow paths of a two-phases flow as
a function of the micromodel thickness \citep{liu2019preferential}
or to study the stability of an immiscible two-phases displacements
in a Hele-Shaw cell with a spatially varying thickness \citep{jackson2017stability}.
Here, we study the influence of micromodel's absolute permeability
on the fluid-fluid drag and solid-fluids drag for different capillary
numbers. The original idea of the article is to solve the appropriate
depth-averaged equations for two-phase flows in a Hele-Shaw cell to
study the effect of the varying absolute permeability on the drag
forces without having to change the in-plane geometry.

\begin{figure}
\begin{centering}
\includegraphics[scale=1.5]{figures/pdf/INTRODUCTION_littPerm}
\par\end{centering}
\caption{Normalized coupling relative permeabilities $K_{ij}^{*}$ as a function
of the wetting-fluid saturation $S_{w}$ from experimental work \citep{Dullien1996,zarcone1994determination},
numerical simulations \citep{Rothman1990,Li2005} and analytical solution
for a steady-state annular two-phase flow in a circular capillary
tube (dashed line) \citep{bacri1990modele}. The capillary theoretical
case provides an upper limit in terms of permeability and extent of
the interfacial surface area between the fluids. \label{fig:comparison_Kr_litterature}}

\end{figure}

\section{Pore-scale, depth-averaged and surface-averaged flow equations}

In this section, we present the derivation of the averaged flow equations
for two-phases flows in a Hele-Shaw cell, starting from the three-dimensional
Stokes equations. Then, we average the momentum equations spatially
to derive the unclosed form of the macroscopic momentum transport
equations. The system under consideration is depicted in Fig.~\ref{fig:Schematic_heleSahw}
which represents a quasi-planar cocurrent two-phase flow between two
parallel plates with a wedge of circular cross-section as a solid
obstacle. The transverse dimension of the cell is noted $L$ and $h$
is the length of the aperture between the plates.

\begin{figure}
\begin{centering}
\includegraphics{figures/pdf/THEORY_heleShawSystemSketch}
\par\end{centering}
\caption{Schematic view of a cocurrent two-phase flow in a Hele-Shaw cell parallel
to the $x-y$ plane with a wedge of circular cross-section as a solid
obstacle. The transverse dimension of the cell is noted $L$ and $h$
stands for the aperture between the plates. The boundary between the
wetting-fluid $w$ and the cylinder (in red) is noted $\Gamma_{wc}$
and the boundary between the two fluids (in blue) is noted $\Gamma_{ow}$.
No dynamic films along the plates are considered here and $\theta$
stands for the non-zero contact angle between the two fluids and the
plates. \label{fig:Schematic_heleSahw}}
\end{figure}

\paragraph{Pore-scale flow equations}

Three-dimensional continuity and Stokes equations for a Newtonian
fluid in the absence of external forcing read, respectively,

\begin{equation}
\nabla^{*}\cdot\mathbf{u}=0,\qquad-\nabla^{*}p+\mu\nabla^{*2}\mathbf{u}=0,\label{eq:3dStokes}
\end{equation}

\noindent where the superscript {*} indicates that the derivative
operators are three-dimensional.

\paragraph{Depth-averaged flow equations}

\noindent The starting point is to consider an aparatus such as depicted
in Fig.~\ref{fig:Schematic_heleSahw} for which $h$ is very small
compared to the transverse length of the cell. It follows that the
$z$-component of the velocity can be neglected and

\begin{equation}
\mathbf{u}=\left(u(x,y,z),v(x,y,z),0\right)^{T}=f(z)\bar{\mathbf{u}}(x,y),
\end{equation}

\noindent where we introduce the depth-averaged velocity defined as
$\bar{\mathbf{u}}\equiv\frac{1}{h}\int_{-h/2}^{h/2}\mathbf{u}\:\mathrm{d}z$.
The in-plane version of Eq.~\ref{eq:3dStokes} then read

\begin{equation}
\nabla\cdot\bar{\mathbf{u}}=0,\qquad-\nabla p+\mu\left(\nabla^{2}\bar{\mathbf{u}}+\bar{\mathbf{u}}\frac{\partial^{2}f(z)}{\partial z^{2}}\right)=0.\label{eq:inPlaneStokes}
\end{equation}

\noindent From the condition $\int_{-h/2}^{h/2}f(z)\:\mathrm{d}z=h$,
arising from the definition of the depth-averaged velocity, along
with the no-slip boundary condition at $z\pm h/2$, we find that $f(z)=\frac{3}{2}(1-4\frac{z^{2}}{h^{2}})$.
Then,

\begin{equation}
\nabla\cdot\bar{\mathbf{u}}=0,\qquad\mu\left(\nabla^{2}\bar{\mathbf{u}}-k^{2}\bar{\mathbf{u}}\right)=\nabla p,\label{eq:amalgamStokesDarcy}
\end{equation}

\noindent are the continuity and momentum transport equations for
the depth-averaged flow of one fluid and $k=\sqrt{12}/h$. It has
been shown that the velocity profile for a flow in a rectangular channel
with Eq.~\ref{eq:amalgamStokesDarcy} lead to a reasonable approximation,
up to aspect ratios $h/L=1$, of the result obtained from the three-dimensional
Stokes equations \citep{nagel2015boundary}. In the case of two-phase
flows, these equations have to be written for each fluid and boundary
conditions at the fluid-fluid interfaces are required. Continuity
of the depth-averaged velocities across the interface and a jump of
interface normal stress are sufficient if the surface tension is constant
along the interface \citep{park1984two}. These conditions are expressed

\begin{equation}
\bar{\mathbf{u}}_{o}-\bar{\mathbf{u}}_{w}=0\;\textrm{at}\;\text{\ensuremath{\Gamma_{ow}}},
\end{equation}

\begin{equation}
\left(\bar{\boldsymbol{\sigma}}_{w}-\boldsymbol{\bar{\sigma}}_{o}\right)\cdot\mathbf{n}_{ow}=\gamma\left(\frac{\pi}{4}\kappa_{\parallel}+\frac{2}{h}\cos\theta\right)\mathbf{n}_{ow}\;\textrm{at}\;\text{\ensuremath{\Gamma_{ow}}},\label{eq:BCHeleShawStress}
\end{equation}

\noindent where $\bar{\boldsymbol{\sigma}}_{i}$ is the in-plane stress
tensor of fluid $i$, $\mathbf{n}_{ow}$ is the in-plane normal vector
at the fluid interface pointing toward the fluid $w$, $\gamma$ is
the surface tension, $\kappa_{\parallel}$ is the in-plane interface
curvature and $\theta$ denotes the contact angle between the fluid
interface and the plates (Fig.~\ref{fig:Schematic_heleSahw}). The
meniscus in the $z$-direction is approximated as a half-circle of
radius $h/2$ and the $\pi/4$ correction for the in-plane curvature
was derived by \citet{park1984two}. In Eq.~\ref{eq:BCHeleShawStress},
we neglected the additional terms that pertain to the formation of
the dynamic film \citep{park1984two} which scaled non-linearly with
the capillary number. We rather considered a non-zero contact angle,
and consequently, the absence of such thin films.

\paragraph{Volume-averaged flow equations}

Here, we proceed to the spatial averaging of the in-plane momentum
transport equations. We recall that all the flow variables and differential
operators have components only in the transverse direction. According
to the volume averaging framework \citep{Whitaker2013} and acknowledging
that Eq.~\ref{eq:amalgamStokesDarcy} are two-dimensional, the traditional
averaging theorem for the depth-averaged quantity $\bar{\omega}_{i}$
associated with the fluid $i$ reads

\begin{equation}
\langle\nabla\bar{\omega}_{i}\rangle=\nabla\langle\bar{\omega}_{i}\rangle+\frac{1}{S}\int_{\Gamma_{ic}}\mathbf{n}_{ic}\bar{\omega}_{i}\:\mathrm{d}\Gamma+\frac{1}{S}\int_{\Gamma_{ij}}\mathbf{n}_{ij}\bar{\omega}_{i}\:\mathrm{d}\Gamma,
\end{equation}

\noindent where,

\begin{equation}
\langle\bar{\omega}_{i}\rangle=\frac{1}{S}\int_{S_{i}}\bar{\omega}_{i}\:\mathrm{d}S,
\end{equation}

\noindent is the superficial surface average and $S$ is the surface
of a representative elementary cell. Applying the superficial surface
average of Eq.~\ref{eq:amalgamStokesDarcy} along with the averaging
theorem and using traditional length-scale arguments \citep{Whitaker2013}
we obtain

\begin{subequations}

\begin{equation}
\nabla\cdot\langle\bar{\mathbf{u}}_{i}\rangle=0,\quad i,j=o,w,\:i\neq j,
\end{equation}

\begin{multline}
\frac{1}{S}\int_{\Gamma_{ic}}\mathbf{n}_{ic}\cdot\left(-p_{i}\mathbf{I}+\mu_{i}\left(\nabla\bar{\mathbf{u}}_{i}+\left(\nabla\bar{\mathbf{u}}_{i}\right)^{T}\right)\right)\:\mathrm{d}\Gamma+\\
+\frac{1}{S}\int_{\Gamma_{ij}}\mathbf{n}_{ij}\cdot\left(-p_{i}\mathbf{I}+\mu_{i}\left(\nabla\bar{\mathbf{u}}_{i}+\left(\nabla\bar{\mathbf{u}}_{i}\right)^{T}\right)\right)\:\mathrm{d}\Gamma-\\
-\mu_{i}k^{2}\langle\bar{\mathbf{u}}_{i}\rangle=\varepsilon_{i}\nabla\langle p_{i}\rangle^{i}+\langle p_{i}\rangle^{i}\nabla\varepsilon_{i},\quad i,j=o,w,\:i\neq j,\label{eq:averagedFull}
\end{multline}

\end{subequations}

\noindent where $\mathbf{I}$ is the $2\times2$ identity matrix and
$\langle p_{i}\rangle^{i}$ $\left(\langle p_{i}\rangle^{i}=\langle p_{i}\rangle/\varepsilon_{i}\right)$
is the intrinsic surface average pressure of fluid $i$, with $\varepsilon_{i}$
the volume fraction of fluid $i$. The first integral is the drag
force exerted upon the cylinders boundary by fluid $i$ and the second
integral pertains for the drag force exerted upon fluid $j$ by fluid
$i$. Here, we consider that the contour of the fluid-fluid interface
in the $x-y$ plane can be identically translated along the $z$-direction,
which is an approximation since the meniscus is a half-circle for
small $h/L$ ratio. However, as shown in the Appendix~\ref{sec:Approximation-made},
using three-dimensional flow simulations in microchannels, these approximations
remain reasonable.

If the variation of the saturation in space is negligible and acknowledging
that, as illustrated in Fig.~\ref{fig:Schematic_heleSahw}, only
the wetting fluid $w$ is in contact with the wedge, a more compact
form of Eq.~\ref{eq:averagedFull} reads

\vspace{-12mm}

\begin{subequations}

\begin{alignat}{1}
0 & =-\varepsilon_{w}\nabla\langle p_{w}\rangle^{w}-\mu_{w}k^{2}\langle\bar{\mathbf{u}}_{w}\rangle+\mathbf{d}_{wc}+\mathbf{d}_{wo},\\
0 & =-\varepsilon_{o}\nabla\langle p_{o}\rangle^{o}-\mu_{o}k^{2}\langle\bar{\mathbf{u}}_{o}\rangle+\mathbf{d}_{ow},
\end{alignat}

\end{subequations}

\noindent Here, $\mathbf{d}_{ij}$ ($\mathbf{d}_{ij}=\int_{\Gamma_{ij}}\boldsymbol{\sigma}_{i}\cdot\mathbf{n}_{ij}\:\mathrm{d}\Gamma$)
denotes the drag forces per unit surface area exerted upon phase $j$
by phase $i$ and which must be computed or modeled to obtain closed
macroscopic equations. In the following, we are working on the direct
calculation of each drag force terms summarized in Tab.~\ref{tab:Summary-of-each-drag}.

\begin{table}
\begin{centering}
\begin{tabular}{cccc}
\toprule 
\begin{tabular}{c}
Drag of\tabularnewline
upon\tabularnewline
\end{tabular} & Fluid $o$ & Fluid $w$ & \tabularnewline
\midrule
\midrule 
Plates & $-\mu_{o}\langle\bar{\mathbf{u}}_{o}\rangle\frac{12}{h^{2}}$ & $-\mu_{w}\langle\bar{\mathbf{u}}_{w}\rangle\frac{12}{h^{2}}$ & \multirow{2}{*}{$\Sigma=\mathbf{d}_{s}$}\tabularnewline
\cmidrule{1-3} \cmidrule{2-3} \cmidrule{3-3} 
Wedge & - & $\mathbf{d}_{wc}$ & \tabularnewline
\midrule 
Fluid $o$ & - & $\mathbf{d}_{wo}$ & \multirow{2}{*}{$\Sigma=\mathbf{d}_{f}$}\tabularnewline
\cmidrule{1-3} \cmidrule{2-3} \cmidrule{3-3} 
Fluid $w$ & $\mathbf{d}_{ow}$ & - & \tabularnewline
\bottomrule
\end{tabular}
\par\end{centering}
\caption{Summary of each drag force terms involved in the averaged momentum
transport equation for two-phase flow in a Hele-Shaw cell.\label{tab:Summary-of-each-drag}}
\end{table}

\section{Direct numerical simulations}

In this section, we introduce the standard Level Set method to capture
the moving free interface between the fluids, along with the flow
equations, both solved with a Finite Element solver.

\subsection{Equations}

\paragraph{Level Set model}

The Level Set method is an Eulerian method that easily handles the
topological phases changes, in contrast with Lagrangian methods. Here,
the fluid phases are identified with a phase color function that goes
smoothly from 0 to 1 across the fluid interface with the manifold
defined as the iso-level $\phi=0.5$. Transport of the level set function
$\phi$ is governed by

\begin{equation}
\frac{\partial\phi}{\partial t}+\nabla\cdot(\bar{\mathbf{u}}\phi)=\tau\nabla\cdot\left(\psi\boldsymbol{\nabla}\phi-\phi(1-\phi)\frac{\boldsymbol{\nabla}\phi}{\vert\nabla\phi\vert}\right),\label{eq:advecPhi}
\end{equation}

\noindent where $\bar{\mathbf{u}}$ is the depth-averaged velocity
field and $\tau$ and $\psi$ are two numerical parameters that control
the diffuse interface thickness and the amount of initialization of
$\phi$ function, respectively \citep{Olsson2007}. We investigated
the accuracy of the implicit definition of the interface as well as
the effect of the value of the initialization parameter on the interface
position in Appendix~\ref{sec:Comparison-with-BEM} by comparing
to a boundary element method \citet{nagel2015boundary}.

The governing flow equations read

\vspace{-12mm}

\begin{subequations}

\begin{alignat}{1}
0 & =\nabla\cdot\bar{\mathbf{u}}\\
0 & =-\nabla p+\mu(\phi)\left(\nabla^{2}\bar{\mathbf{u}}-\frac{12}{h^{2}}\bar{\mathbf{u}}\right)+\gamma\left(\frac{\pi}{4}\nabla\cdot\left(\frac{\boldsymbol{\nabla}\phi}{\vert\boldsymbol{\nabla}\phi\vert}\right)-\frac{2}{h}\right)\delta(\phi)\mathbf{n},\label{eq:Momentum-1}
\end{alignat}

\end{subequations}

\noindent where $\delta$ is the Dirac delta function localized on
the interface and $\mathbf{n}$ denotes the unit normal to the interface,
respectively defined as,

\begin{equation}
\delta(\phi)=6\left|\boldsymbol{\nabla}\phi\right|\left|\phi\left(1+\phi\right)\right|,\quad\textrm{and}\quad\mathbf{n}=\frac{\boldsymbol{\nabla}\phi}{\left|\boldsymbol{\nabla}\phi\right|}.
\end{equation}

\noindent We introduce the following reference and dimensionless quantities,

\begin{equation}
\bar{\mathbf{u}}=\bar{\mathbf{u}}'\times U_{r},\;p=p'\times\frac{\mu_{r}U_{r}}{L},\;\mathbf{x}=\mathbf{x}'\times L,\label{eq:referenceQuantities}
\end{equation}

\noindent and the dimensionless continuity and momentum transport
equations are

\vspace{-12mm}

\begin{subequations}

\begin{alignat}{1}
0 & =\nabla'\cdot\bar{\mathbf{u}}'\\
0 & =-\nabla'p'+\frac{\mu(\phi)}{\mu_{r}}\left(\nabla'^{2}\bar{\mathbf{u}}'-\frac{12}{\left(h/L\right)^{2}}\bar{\mathbf{u}}'\right)+\frac{\gamma}{\mu_{r}U_{r}}\left(\frac{\pi}{4}\nabla'\cdot\left(\frac{\boldsymbol{\nabla'}\phi}{\vert\boldsymbol{\nabla'}\phi\vert}\right)-\frac{2}{h/L}\right)\delta'(\phi)\mathbf{n},\label{eq:dimensionlessMomentum}
\end{alignat}

\end{subequations}

\noindent with $\delta'(\phi)=6\left|\boldsymbol{\nabla'}\phi\right|\left|\phi\left(1+\phi\right)\right|$.
We therefore have three dimensionless numbers: the viscosity ratio
$M(\phi)=\frac{\mu(\phi)}{\mu_{r}}$, the capillary number $Ca^{-1}=\frac{\gamma}{\mu_{r}U_{r}}$
and the aspect ratio $h^{*}=h/L$.

\paragraph{Geometry, boundary conditions and simulation parameters}

Our macroscopic model is a Hele-Shaw cell with wedges of cylindrical
cross-section. This system is subdivided into seven unit-cell (UC)
subdomains encompassing one wedge, as depicted in Fig.~\ref{fig:model}.
Taking advantage of the symmetry, we studied only the upper half of
a row. Each fluid flows from left to right ($x$-direction) and the
inlet boundary conditions for both fluids are a constant normal inlet
velocity $u_{i}$. The outlet boundary condition for the flow is a
reference pressure. The boundary conditions used are summarized in
Tab.\ref{tab:Boundary-conditions}. We choose the total inlet velocity
as the reference velocity; thus the dimensionless inlet velocities
can be expressed as a fractional flow $f_{f}$,

\begin{equation}
f_{f}=\frac{u_{w}}{U_{t}},\quad u_{o}=1-f_{f},\quad\textrm{with}\quad U_{t}=u_{w}+u_{o}.
\end{equation}

\noindent The non-wetting viscosity is taken as the reference viscosity
and the respective value of each dimensionless parameters is in Tab.~\ref{tab:Boundary-conditions}.
The initialization parameter value is equal to the maximum inlet velocity
value, which yields maximum accuracy (Appendix~\ref{sec:Comparison-with-BEM}).
As a reference, we conducted numerical simulations of one-phase flows
and found that the range of aperture corresponds to a range absolute
permeabilities between $1.5\times10^{4}$ and $40$ darcy. 
\begin{figure}
\begin{centering}
\includegraphics{figures/pdf/DNS_model}
\par\end{centering}
\caption{Schematics of the geometry and initial condition. We consider the
superior half of an array of five cylindrical wedges inside five cuboids
where both fluids are injected from left to right. Initially, the
model is saturated with wetting fluid (red), and the width $L$ of
one Unit Cell (UC) is $5\times10^{-4}$ m. Symmetry boundary conditions
are used on both length sides and the no-flow boundary condition is
imposed at the wedge boundary. \label{fig:model}}
\end{figure}

\begin{table}
\caption{Boundary conditions for flow variables and the Level Set function
(left) and simulation parameters (right).\label{tab:Boundary-conditions}}

\centering{}%
\begin{tabular}{cccc}
\toprule 
Boundary & $u$ & $p$ & $\ensuremath{\phi}$\tabularnewline
\midrule
\midrule 
Outlet & - & $0$ & $\mathbf{n}\cdot\boldsymbol{\nabla}\phi=0$\tabularnewline
\midrule 
Inlet $o$ & $u_{o}$ & - & $0$\tabularnewline
\midrule 
Inlet $w$ & $u_{w}$ & - & $1$\tabularnewline
\bottomrule
\end{tabular}\hfill{}%
\begin{tabular}{cc}
\toprule 
Parameters & Value\tabularnewline
\midrule
\midrule 
$Ca=\frac{U_{t}\mu_{o}}{\gamma}$ & from $0.5$ to $0.02$\tabularnewline
\midrule 
$M_{w}=\frac{\mu_{w}}{\mu_{o}}$ & 1\tabularnewline
\midrule 
$f_{f}=\frac{u_{w}}{U_{t}}$ & 1/4\tabularnewline
\midrule 
$h^{*}=h/L$ & from $5$ to $1/20$\tabularnewline
\bottomrule
\end{tabular}
\end{table}

\paragraph{Mesh convergence study}

Here, we study the mesh convergence of the various drag forces. The
dimensionless numbers for this study are $Ca=0.5$, $h^{*}=1/4$,
$f_{f}=0.25$ and $M=1$. In Fig.~\ref{fig:Mesh-convergence-study}
(a) the results are normalized with respect to the finer mesh result
and are given as a function of the total number of degree of freedom
in the whole model. The fluid-fluid interface position for three different
mesh densities is given in Fig.~\ref{fig:Mesh-convergence-study}
(b). We obtain these results in the fourth unit-cell (UC4) at steady-state.
The drag terms are not very sensitive to the mesh density and the
interface between the fluids is correct even for a coarse mesh and
converges quickly. In the following simulations, we use a mesh that
corresponds to $3.7\times10{{}^5}$ degree of freedoms

\begin{figure}[H]
\begin{centering}
\includegraphics{figures/pdf/DNS_meshConv}
\par\end{centering}
\caption{Mesh convergence study of (a) drag force exerted upon the wedge (red),
drag force exerted upon the fluid-fluid boundary (blue) and intrinsic
average velocity of fluid $o$ (green) . All the results are normalized
with respect to the result obtained with the finer mesh, at steady-state
and in UC4. (b) Fluid-fluid interface position in UC4 at steady-state
for different number of degree of freedoms in the whole geometry.
Flow parameters are $Ca=0.5$, $h^{*}=1/4$, $f_{f}=0.25$ and $M=1$.
In the following we use a mesh that corresponds to $3.7\times10{{}^5}$
degree of freedoms.\label{fig:Mesh-convergence-study}}
\end{figure}

\section{Results and discussion}

Before presenting the results for drag force terms, we briefly discuss
the two-phase flow regime observed along with the variation of the
saturation with time. The flow regime is fundamental as it drives
the extent of the interfacial surface area between the fluids.

\paragraph{Two-phase flow regimes}

We observe that the two fluids remain continuous at all times and
for the entire range of tested capillary numbers and aperture. The
interface between the fluids becomes stationary and a stationary state
is reached for every capillary number and aperture values. Fig.~\ref{fig:Fluid-repartion-along}
shows the initial, intermediate, and final configurations of the fluid
repartition for the case $Ca=1\times10^{-1}$ and $h^{*}=1/8.$ At
steady-state, the fluid-fluid interface is periodic on the five central
unit cells whereas the interface is slightly deformed at the inlet
and outlet cells, under the influence of boundary conditions.

We could have expected a break-up of the invading phase either by
snap off phenomena, or by the shear action of the wetting fluid flow.
However, the pore throat is large and the fractional flow is low,
which favours the continuity of the fluid phases. The flow regime
observed is therefore a film-flow regime within the limit of the parameters
chosen for this study. We obtained all the following results in the
fourth unit-cell (UC4) in order to be far from the influence of the
inlet and outlet boundary conditions.

\begin{figure}
\begin{centering}
\includegraphics[scale=1.2]{figures/pdf/RESULTS_dynamic}
\par\end{centering}
\caption{Fluids repartion along the superior half-row at (a) the initial time,
(b) for an intermediate time and (c) final time (steady-state reached)
for $Ca=0.1$, $f_{f}=0.25$, $M=1$ and $h^{*}=1/8$. At steady-state
the fluid-fluid interface taken on the central unit cells is periodic
whereas it is slightly deformed in the first and last UC under the
influence of the boundary conditions. \label{fig:Fluid-repartion-along}}
\end{figure}

\paragraph{Saturation}

Wetting fluid saturation at steady-state decreases, in average, from
0.6 to 0.4 as the aperture between the plates decreases from 5 to
1/20. This can be seen in Fig.~\ref{fig:wettingFluidSaturation},
which presents the saturation in UC4 as a function of the dimensionless
aperture and for different capillary numbers. The saturation does
not depend on the capillary number for the smallest apertures. However,
as the aperture between the plates increases, the fluid-fluid interface
becomes flatter and the saturation of the wetting fluid increases
until it converges to the saturation obtained for the 2D limit case
which values slightly depend on the capillary number (see saturation
fields insets in Fig.~\ref{fig:wettingFluidSaturation}). Fig.~\ref{fig:Comparison-of-interface}
compares the position of the fluid-fluid interface in UC4 for different
aperture and capillary numbers. For $Ca=5\times10^{-1}$ the interface
is symmetric regardless the aperture's value, whereas the interface
is non-symmetric for low $Ca$ and small aperture.

\begin{figure}
\begin{centering}
\includegraphics[scale=1.5]{figures/pdf/RESULTS_saturation}
\par\end{centering}
\caption{Fluid saturation at steady-state in UC4 as a function of the dimensionless
aperture and for different capillary numbers. Fields of the level-set
function at steady-state in UC4 are given for the selected value of
the dimensionless aperture and $Ca=5\times10^{-1}$ and $Ca=2\times10^{-2}$.
As the aperture increases the interface flattens and the wetting fluid
saturation increases. A decrease in the capillary number acts in the
same way, albeit less strongly. The 2D limit case is given with plain
black markers. \label{fig:wettingFluidSaturation}}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics{figures/pdf/RESULTS_interfacePosition}
\par\end{centering}
\caption{Comparison of interface position for different aperture (a) $Ca=5\times10^{-1}$and
(b) $Ca=2\times10^{-2}$ (in UC4 for $f_{f}=0.25$ and $M=1$).\label{fig:Comparison-of-interface}}
\end{figure}

\paragraph{Drag force}

We compute the fluid-fluid drag according to Eq.\ref{eq:averagedFull}.
The viscous part is obtained through the velocity gradient available
at the $\phi=0.5$-contour and the pressure part is obtained by application
of the divergence theorem. In the following, we are interested in
the drag force component in the $x$-direction, i.e., the main flow
direction. The drag force exerted upon the solid-fluids boundaries,
denoted $d_{s}^{x}$, is the sum of the drag upon the wedge and upon
the cell plates by both fluids, as defined in Tab~\ref{tab:Summary-of-each-drag}.
The fluid-fluid drag, denoted $d_{f}^{x}$, is the sum $d_{f}^{x}=d_{ow}^{x}+d_{wo}^{x}$.
In the following, the $x$–component of the drag forces are expressed
per unit surface area of unit-cell.

Fig.~\ref{fig:ratioDragForceTotal} shows the ratio of the fluid-fluid
drag over the solid-fluid drag as a function of the aperture between
the plates and different values of capillary number. \lyxadded{Maxime}{Wed Jul  1 11:01:27 2020}{Our
results show that the fluid-fluid drag is non-negligible compared
to the solid-fluid drag. At most, the fluid-fluid drag reaches 80\%
of the value of the solid-fluid drag and this occur for high $Ca$
and large aperture. The results from the purely 2d simulations ($h^{*}\rightarrow\infty)$
indicate that for $h/L=5$, the larger aperture tested, the contribution
of the drag upon the plates is already insignificant. Then, by narrowing
the aperture, the fluid-fluid drag decreases compared to the solid-fluid
drag, until $h^{*}=1/2$. From $h^{*}=1/2$ and below the fluid-fluid
drag either remains constant, for the higher Ca, or increases. In
the end, for the smallest aperture, the importance of the fluid-fluid
drag does not depend on the capillary number and is 40\% of the value
of the solid-fluid drag.}

\lyxadded{Maxime}{Wed Jul  1 11:02:24 2020}{Regarding the limit case
of infinite aperture the capillary number is important}

\begin{figure}
\begin{centering}
\includegraphics[scale=1.5]{figures/pdf/RESULTS_ratioDrag}
\par\end{centering}
\caption{Ratio of fluid-fluid drag force over solid-fluid drag force as a function
of the dimensionless aperture and for different capillary numbers
(in UC4 for $f_{f}=0.25$ and $M=1$). The 2D limit case is given
with plain black markers. \label{fig:ratioDragForceTotal}}
\end{figure}

Fig.~\ref{fig:Fluid-fluid-and-solid-fluid} shows the fluid-fluid
ant the solid-fluid drag forces as a function of the aperture and
for two capillary numbers. Here again, the dimensionless aperture
$h^{*}=1/2$ signal a change in the drag behavior. For $1/2<h^{*}$,
the fluid-fluid and solid-fluid drag forces barely change with the
aperture whereas, for $h^{*}<1/2$ both drag forces scale approximately
as $h^{-2}.$

\begin{figure}
\begin{centering}
\includegraphics[scale=1.5]{figures/pdf/RESULTS_Drag}
\par\end{centering}
\caption{Comparison of fluid-fluid and solid-fluid drag force as a function
of the dimensionless aperture for two capillary numbers (in UC4 for
$f_{f}=0.25$ and $M=1$).\label{fig:Fluid-fluid-and-solid-fluid}}
\end{figure}

Fig.~\ref{fig:Proportion-solid--drag} shows the proportion of the
solid-fluid drag which is due to the drag exerted upon the wedge as
a function of the aperture. Here the 2D limit case is clear since
the solid-fluid drag is only due to the drag upon the wedge in that
case. The importance of the drag upon the wedge decreases as the aperture
decrease, however it remains non-negligible (40\% of the solid-fluid
drag) even for the smallest aperture $h^{*}=1/20$.

\begin{figure}
\begin{centering}
\includegraphics[scale=1.5]{figures/pdf/RESULTS_solidDrag}
\par\end{centering}
\caption{Proportion of the solid-fluid drag force due to the contribution of
the drag exerted upon the wedge, as a function of the dimensionless
aperture and for different capillary numbers (in UC4 for $f_{f}=0.25$
and $M=1$). \label{fig:Proportion-solid--drag}}
\end{figure}

Fig.~\ref{fig:Proportion-of-pressureTerm} shows the proportion of
the fluid-fluid and solid-fluid drag upon the wedge which is due to
the pressure part of the stress-tensor. The pressure part produces
most of the drag force, either regarding the fluid-fluid or the solid-fluid
drag upon the wedge, for $h^{*}<1/2$. For larger aperture, the fluid-fluid
drag is mainly produced by the viscous part, whereas the pressure
part has a minimum contribution as high as 60\% of $d_{wc}$.

\begin{figure}
\begin{centering}
\includegraphics[scale=1.3]{figures/pdf/RESULTS_decompose}
\par\end{centering}
\caption{Proportion of the fluid/fluid drag (a) or the solid-fluid drag upon
the wedge (b) due to the pressure term contribution, as a function
of the dimensionless aperture and for different capillary numbers
(in UC4 for $f_{f}=0.25$ and $M=1$) \label{fig:Proportion-of-pressureTerm}}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics{figures/pdf/RESULTS_details}
\par\end{centering}
\caption{Comparison of the drag produced by the pressure for}
\end{figure}

\section{Conclusion}

In this study we conducted direct simulations of depth-averaged two-phases
flows, and we investigated the effect of the permeability on the drag
forces exerted upon the different phase interfaces. The permeability
was changed by varying the Darcean term which arising from the depth-averaging,
thus without altering the in-plane geometry. These drag terms have
to be modelled to obtain the macroscopic momentum transport equations
but the drag exerted upon the the fluid-fluid interface is commonly
neglected for flow driven by capillarity forces. Here we focused on
film-flow regime pourencountered in two-phases flows in high permeability
porous media or in microfluidic devices. We found that the drag exerted
upon the fluid-fluid interface should not be neglected into the momentum
transport equations for film-flow regimes. Lower permeability does
not make the drag force terms at the interface negligible as long
as the flow regime remains a film regime. On the contrary, the fluid-fluid
drag force increases faster than the drag upon the fluid-solid interfaces,
as the aperture between the plates decreases.

\newpage{}

\bibliographystyle{elsarticle-harv}
\bibliography{biblio}

\newpage{}

\appendix

\section{Approximation made on the drag force when considering a fluid-fluid
flat interface\label{sec:Approximation-made}}

We conduct three-dimensional one-phase flow simulations into a microchannel
to determine the approximation made on the drag calculation when considering
that the fluid-fluid contour can be translated along the $z$-direction.

One side of the microchannel is a half-circle wall to mimic a fluid-fluid
interface whereas the opposite side is a flat wall (Fig.~\ref{fig:Microchannel}).
We obtain the curved side by extruding a cylinder; thus a special
treatment is done to correctly meshed the very thin part left, and
we build very small chamfers (Fig.~\ref{fig:Microchannel}). We compute
the drag force per unit surface area on each sidewall of the microchannel
and plot in Fig.~\ref{fig:Drag-force-exerted} the drag force exerted
upon the curved side normalized with respect to the flat wall. The
drag per unit surface area exerted upon the curved wall is roughly
70\% of the drag per unit surface area upon the flat wall. Giving
the greater extent of surface area for the curved wall, the difference
regarding the drag value is negligible.

\begin{figure}
\begin{centering}
\includegraphics[scale=0.7]{figures/pdf/APPENDIX_microchannel}
\par\end{centering}
\caption{Microchannel with one side is a half-circle wall and the opposite
side is a flat wall with $h/L=1/16$ (left) and mesh detail at the
small chamfer build after the cylinder extrusion (right).\label{fig:Microchannel}}
\end{figure}

\begin{figure}
\begin{centering}
\includegraphics[scale=0.7]{figures/pdf/APPENDIX_test_half_circle}
\par\end{centering}
\caption{Drag force and drag force per unit surface area exerted upon the curved
wall of the microchannel normalized with respect to the drag force
exerted upon the flat wall. The drag exerted upon the curved solid
wall is almost equal to the drag exerted upon the flat wall, since
the smaller drag per unite surface area is almost fully compensated
by the greater surface area. \label{fig:Drag-force-exerted}.}
\end{figure}

\section{Comparison with a Boundary Element Method\label{sec:Comparison-with-BEM}}

Here we validate the Level Set (LS) code by comparison with a Boundary-Element
Method (BEM) \citep{nagel2015boundary}, which relies on a surface
discretization of the interface and a pseudo-analytical formulation
in the bulk of the phases. This allows us to precisely locate the
interface, even in the case of very thin film flow, and to carefully
analyze the choice of parameters in Eq.~\ref{eq:advecPhi}.

The test case resembles the case study (Fig.~\ref{fig:Test-case}).
The model is initially fully saturated with fluid $w$, with the exception
of the channel through which fluid $o$ is injected. We test either
the $\phi=0.5$ contour is an appropriate definition for the fluid-fluid
interface and in what extent the value of the initialization parameter
$\psi$ can change the interface position. In Fig.~\ref{fig:Interface-position}
we present the half-part of the model and the interface between the
fluids at time $t=2.5$, for an aspect ratio $h/L$, where $L$ is
the width of the fluid $o$ inlet channel. We present three different
interfaces obtained with the LS method, depending on the initialization
parameter value. This parameter is normalized with respect to the
inlet velocity of fluid $o$, which is three times greater than the
fluid $w$ inlet velocity. One can clearly observe that the interface
position obtained with the LS method is almost identical to that obtained
with the BEM, regardless of value of $\psi$. However, the best results,
especially regarding the interface tip position, is obtained for $\psi=U_{o}^{\mathrm{inlet}}$.

\begin{figure}
\begin{centering}
\includegraphics{figures/pdf/APPENDIX_compare_model_BEM}
\par\end{centering}
\caption{Test case for comparison between Level Set and Boundary Element methods.
The flow is cocurrent from left to right and the viscosity is the
same for both fluids.\label{fig:Test-case}}

\end{figure}

\begin{figure}
\begin{centering}
\includegraphics[scale=0.75]{figures/pdf/APPENDIX_interface_BEMvsLS}
\par\end{centering}
\caption{Interface position at time $t=2.5\:\mathrm{s}$ obtained for an aspect
ratio $h/L=0.25$ with a Boundary Element method and the Level Set
method for different values of the dimensionless initialization parameter
$\psi^{*}=\psi/U_{o}^{\mathrm{inlet}}$. \label{fig:Interface-position}}

\end{figure}

\end{document}

### 2020-07-11 presentation/presentation.tex
last update

% Created 2020-07-09 jeu. 20:51
% Intended LaTeX compiler: pdflatex
\documentclass[presentation]{beamer}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{palatino}
\usepackage{newtxmath}
\usetheme{Boadilla}
\author{Maxime Cochennec}
\date{\today}
\title{Résultats de l'article 1}
\hypersetup{
 pdfauthor={Maxime Cochennec},
 pdftitle={Résultats de l'article 1},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 26.3 (Org mode 9.3.6)}, 
 pdflang={English}}
\begin{document}

\maketitle
\begin{frame}{Outline}
\tableofcontents
\end{frame}

\section{Introduction}
\label{sec:org491c0a8}
\begin{frame}[label={sec:orgd908bc7}]{Flow regimes and instabilities in Hele-Shaw cell}
\begin{itemize}
\item Flow regimes:

\begin{itemize}
\item \alert{continuous stream}

\item drop and/or ganglia flow
\end{itemize}
\end{itemize}

Break-up of the invading fluid by snap-off?

\begin{itemize}
\item Instabilities:

\begin{itemize}
\item Saffman-Taylor

\item Rayleigh-Plateau (see (b) bellow from \cite{cubaud2008capillary})
\end{itemize}
\end{itemize}

\centering
\includegraphics[scale=0.352]{cubaudMason.png}
\end{frame}

\begin{frame}[label={sec:org630fdfe}]{Boundary conditions and simulation parameters}
\begin{table}
\centering{}
\begin{tabular}{cccc}
\toprule 
Boundary & $u$ & $p$ & $\ensuremath{\phi}$\tabularnewline
\midrule
\midrule 
Outlet & - & $0$ & $\mathbf{n}\cdot\boldsymbol{\nabla}\phi=0$\tabularnewline

Inlet $o$ & $u_{o}$ & - & $0$\tabularnewline

Inlet $w$ & $u_{w}$ & - & $1$\tabularnewline
\bottomrule
\end{tabular}\hfill{}%
\begin{tabular}{cc}
\toprule 
Parameters & Value\tabularnewline
\midrule
\midrule 
$Ca=\frac{U_{t}\mu_{o}}{\gamma}$ & from $0.125$ to $0.005$\tabularnewline

$M_{w}=\frac{\mu_{w}}{\mu_{o}}$ & 1\tabularnewline

$f_{f}=\frac{u_{w}}{U_{t}}$ & 1/4\tabularnewline

$h^{*}=h/L$ & from $5$ to $1/20$\tabularnewline
\bottomrule
\end{tabular}
\caption{Boundary conditions (left) and simulation parameters (right)}
\end{table}

\centering
\includegraphics[scale=0.7]{DNS_model.pdf}
\end{frame}

\begin{frame}[label={sec:org1a3e78c}]{Equations}
If \(\nabla \epsilon_i \approx 0\), the momentum transport equations read

\begin{subequations}
\begin{align}
0&=-\varepsilon_{w}\nabla\langle p_{w}\rangle^{w}-\mu_{w}k^{2}\langle\bar{\mathbf{u}}_{w}\rangle+\mathbf{d}_{wc}+\mathbf{d}_{wo},\\
0&=-\varepsilon_{o}\nabla\langle p_{o}\rangle^{o}-\mu_{o}k^{2}\langle\bar{\mathbf{u}}_{o}\rangle+\mathbf{d}_{ow}.
\end{align}
\end{subequations}

where, \(\mathbf{d}_{ij}\) has dimensions \(\mathrm{Pa.m^{-1}}\) i.e. drag forces
per unit surface area (of unit-cell).

\begin{block}{Drag definition}
$\mathbf{d}_{ij}= \frac{1}{S} \int_{\Gamma_{ij}}\sigma_i \cdot \mathbf{n}_{ij} \:
\mathrm{d} \Gamma$, 
\begin{itemize}
\item $\sigma_i$ is the stress-tensor for a Newtonian fluid $i$,
\item $S$ is the unit-cell's surface
\item $\mathbf{n}_{ij}$ is the unit normal vector pointing toward the $j$-phase.
\end{itemize}
\end{block}
\end{frame}
\begin{frame}[label={sec:org20ba60f}]{Drag}
\begin{table}
\begin{centering}
\begin{tabular}{cccc}
\toprule 
\begin{tabular}{c}
Drag of...\tabularnewline
upon...\tabularnewline
\end{tabular} & Fluid $o$ & Fluid $w$ & \tabularnewline
\midrule
\midrule 
Plates & $-\mu_{o}\langle\bar{\mathbf{u}}_{o}\rangle\frac{12}{h^{2}}$ & $-\mu_{w}\langle\bar{\mathbf{u}}_{w}\rangle\frac{12}{h^{2}}$ & \multirow{2}{*}{$\Sigma=\mathbf{d}_{s}$}\tabularnewline
\cmidrule{1-1}
Wedge & - & $\mathbf{d}_{wc}$ & \tabularnewline
\midrule 
Fluid $o$ & - & $\mathbf{d}_{wo}$ & \multirow{2}{*}{$\Sigma=\mathbf{d}_{f}$}\tabularnewline
\cmidrule{1-1} 
Fluid $w$ & $\mathbf{d}_{ow}$ & - & \tabularnewline
\bottomrule
\end{tabular}
\caption{Summary of each drag force terms involved in the averaged momentum
transport equations for two-phase flows in a Hele-Shaw cell.\label{tab:Summary-of-each-drag}}
\par\end{centering}
\end{table}

\begin{alertblock}{Information}
In the following we are interested in the x-component of the drag
(i.e. component align with the main flow direction).
\end{alertblock}
\end{frame}

\section{Results}
\label{sec:org70e6955}

\begin{frame}[label={sec:org11bbfbb}]{Results: flow regimes}
\end{frame}
\begin{frame}[label={sec:orgbf39abd}]{Results: saturation}
\begin{figure}
\centering
\includegraphics[scale=0.8]{RESULTS_saturation.pdf}
\caption{Saturation in wetting fluid as a function of the dimensionless gap between the plates.}
\end{figure}
\end{frame}

\begin{frame}[label={sec:org98afb82}]{Results: fluid-fluid interface}
\end{frame}

\begin{frame}[label={sec:orgb51f6ba}]{Results: solid-fluid drag force (1)}
\begin{figure}
\centering
\includegraphics[scale=0.8]{RESULTS_drag.pdf}
\caption{Comparison of the solid-fluid drag upon the wedge and upon the Hele-Shaw plates as a function of the
dimensionless gap.}
\end{figure}
\end{frame}

\begin{frame}[label={sec:orgeed5a19}]{Results: solid-fluid drag force (2)}
\begin{alertblock}{Drag upon the plates}
\begin{enumerate}
\item constant inlet velocity whereas the gap between the plates is narrowing
\item drag : $-\mu_i \langle u_i \rangle \frac{12}{h^2}$
\item the drag upon the plates scales as $h^{-2}$
\end{enumerate}
\end{alertblock}

\begin{exampleblock}{Drag upon the wedge}

\begin{enumerate}
\item constant geometry
\item velocity gradient depends on the fluid-fluid interface position
\item pressure increases as the gap is narrowing
\end{enumerate}

\end{exampleblock}
\end{frame}

\begin{frame}[label={sec:org75b4b07}]{Results: fluid-fluid drag force (1)}
\begin{alertblock}{Fluid-fluid drag}
\begin{enumerate}
\item interface is changing (slightly)
\item pressure gradient increases as the gap is narrowing since the inlet velocity is constant
\end{enumerate}
\end{alertblock}
\end{frame}

\begin{frame}[label={sec:org76545b5}]{Results: fluid-fluid drag force (2)}
\end{frame}
\begin{frame}[label={sec:orgfc92227}]{Results: drag ratio}
\begin{figure}
\centering
\includegraphics[scale=0.8]{RESULTS_dragRatio.pdf}
\caption{Ratio of fluid-fluid drag over solid-fluid drag as a function of the dimensionless gap
 between the plates.}
\end{figure}
\end{frame}

\begin{frame}[label={sec:orgbc5cfbd}]{Results: drag}
\begin{figure}
\centering
\includegraphics[scale=0.8]{RESULTS_dragSepare.pdf}
\caption{Comparison of fluid-fluid and solid-fluid drag force for differrent capillary numbers as a
function of the dimensionless gap.}
\end{figure}
\end{frame}

\begin{frame}[label={sec:orgab07557}]{Results: Flat interface}
What happens if the fluid-fluid interface is flat for \(1<h/L\)?

\begin{enumerate}
	\item we keep the drag upon the plates in the momentum transport equations
	\item the contact angle is 0 and the pressure jump across the interface is only due to the $x-y$ curvature
\end{enumerate}
\end{frame}

\begin{frame}[label={sec:org497bd82}]{Results: Dynamic film formation}
By following \cite{park1984two}, the thickness of the wetting fluid film
scales, at leading order, as 
\begin{equation}
\frac{h}{6}
\end{equation}

for \(Ca= 1.25 \times 10^{-1}\).
\end{frame}

\begin{frame}[label={sec:org00a4ea3}]{References}
\bibliographystyle{unsrt}
\bibliography{article}
\end{frame}
\end{document}

### 2020-09-20 validation_pressureDrag/20-09-09.validation_intPressureDrag.tex
last commit

% Created 2020-09-18 ven. 09:30
% Intended LaTeX compiler: pdflatex
\documentclass[justified]{tufte-handout}
 \usepackage{multicol}
\usepackage{units}
\usepackage[french, english]{babel}

  \usepackage{graphicx}
  \usepackage[colorlinks=true]{hyperref}
\author{Maxime Cochennec}
\date{\today}
\title{Vérification calcul pressure drag}
\hypersetup{
 pdfauthor={Maxime Cochennec},
 pdftitle={Vérification calcul pressure drag},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 26.3 (Org mode 9.3.6)}, 
 pdflang={English}}
\begin{document}

\maketitle
\tableofcontents

\section{Problème}
\label{sec:org62e5345}

On cherche à calculer l'intégrale \footnote{La composante selon \(x\) de l'intégrande est nul si symétrie
d'axe \(y\). Dans le cas d'une double symétrie d'axe \(x\) et \(y\), par
ex. une bulle parfaitement cylindrique, l'intégrande est nul.}
\begin{equation}\label{eq:int}
\frac{1}{S}\int_{\Gamma_{wo}}\gamma\left(\frac{\pi}{4}\kappa_{\parallel}-\frac{2}{h}\cos\theta\right)\mathbf{n}_{ow}  \:\mathrm{d}l,
\end{equation}
en particulier on s'intéresse à la composante selon \(x\) afin de
discuter la différence entre les forces de trainées \(d_{ow} \vert_x -
d_{wo}\vert_x\).

\vspace{0.8cm}
\noindent
L'intégrale \ref{eq:int} est calculée en
dehors de Comsol comme suit :
\begin{itemize}
\item export de l'interface,
\item reconstruction par fitting avec un polynôme,
\item calcul sur l'interface reconstruite.
\end{itemize}
\noindent
On note que
\begin{equation}
\int_{\Gamma_{wo}}(p_{w}-p_{o})\mathbf{n}_{ow}\cdot\mathbf{e}_{x}\:\mathrm{d}l=\int_{\Gamma_{wo}}\gamma\left(\frac{\pi}{4}\kappa_{\parallel}-\frac{2}{h}\cos\theta\right)\mathbf{n}_{ow}\cdot\mathbf{e}_{x}\:\mathrm{d}l,
\end{equation}
\noindent puisque la viscosité des fluides est identique. On vérifie
d'abord que l'intégrale \ref{eq:int} est bien nulle dans le cas d'une
interface avec une symétrie d'axe \(y\). Ensuite, on présente les résultats.

\section{Vérification avec interface symétrique}
\label{sec:org7ecbd00}

\begin{marginfigure}
  \includegraphics[width=\linewidth]{symInt.pdf}
  \caption{Interface symétrique utilisée pour l'exemple.}
  \label{fig:marginfig1}
\end{marginfigure}

On utilise une interface symétrique (Fig.\ref{fig:marginfig1}) afin de
vérifier si le calcul de l'intégrale \ref{eq:int} renvoie bien un
résultat nul. Les résultats sont donnés dans la Table
\ref{tab:org7327f8c}. L'intégrale est bien nulle.

\begin{table}[htbp]
\caption{\label{tab:org7327f8c}Courbure moyenne dans le plan et résultat de l'intégrale pour l'interface symétrique Fig.\ref{fig:marginfig1}}
\centering
\begin{tabular}{rl}
\(\langle \kappa\)\textsubscript{\(\parallel\)}\(\rangle\) (1/m) & Intégrale (N/m)\\
\hline
-308 & 3\texttimes{}10\textsuperscript{-17}\\
\end{tabular}
\end{table}

\section{Résultats}
\label{sec:org88b19f6}
On calcule maintenant la courbure moyenne et l'intégrale \ref{eq:int}
en fonction de l'entrefer et pour deux nombres capillaire. Les
résultats sont donnés dans les Tables \ref{tab:orge94ce7b} et \ref{tab:org961c12c}.

\begin{table}[htbp]
\caption{\label{tab:orge94ce7b}Courbure moyenne dans le plan et résultat de l'intégrale en fonction de l'entrefer (\(Ca=0.5\)).}
\centering
\begin{tabular}{rrrrr}
\(h^*\) & \(\langle \kappa\)\textsubscript{\(\parallel\)}\(\rangle\) (1/m) & Intégrale (N/m\textsuperscript{3}) & \(\int \mathbf{n} \cdot \mathbf{e}_x \: \mathrm{d} l\) & \(d_{ow}\) (N/m\textsuperscript{3})\\
\hline
5 & 25 & -4.65e-3 & -2.79E-06 & 15.2\\
2 & 29 & -1.83e-3 & -3.21E-06 & 16.3\\
1 & 33 & 8.54e-3 & -4.90E-06 & 21.2\\
0.5 & 39 & 3.44e-2 & -5.92E-06 & 43.7\\
0.25 & 43 & 6.05e-2 & -4.26E-06 & 138.5\\
0.125 & 53 & 5.05e-2 & -1.67E-06 & 490.2\\
0.05 & 60 & 7.28e-2 & -9.24E-07 & 2484.0\\
\end{tabular}
\end{table}

\begin{table}[htbp]
\caption{\label{tab:org961c12c}Courbure moyenne dans le plan et résultat de l'intégrale en fonction de l'entrefer (\(Ca=0.025\)).}
\centering
\begin{tabular}{rrrrr}
\(h^*\) & \(\langle \kappa\)\textsubscript{\(\parallel\)}\(\rangle\) (1/m) & Intégrale (N/m\textsuperscript{3}) & \(\int \mathbf{n} \cdot \mathbf{e}_x \: \mathrm{d} l\) & \(d_{ow}\) (N/m\textsuperscript{3})\\
\hline
5 & -11 & 2.48e-2 & 3.70E-05 & 13.3\\
2 & -12 & -2.84e-4 & 2.96E-05 & 13.5\\
1 & -17 & 1.00e-1 & 2.14E-07 & 14.6\\
0.5 & -23 & 6.10e-1 & -4.55E-05 & 24.0\\
0.25 & -1 & 5.93e+0 & -2.93E-04 & 109.9\\
0.125 & 56 & 1.31e+1 & -3.61E-04 & 431\\
0.05 & 61 & 8.32e+0 & -1.07E-04 & 2502\\
\end{tabular}
\end{table}

\noindent Les résultats de l'intégrale \ref{eq:int} sont très petits
au regard de la force de trainée calculée dans le papier. Ces
résultats confirment le résultat du papier, c-à-d, \(d_{ow} \vert_x \approx
d_{wo}\vert_x\).

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{Int.pdf}
  \caption{Position de l'interface fluide-fluide en fonction de l'entefer pour $Ca=0.025$. Une symétrie
  d'axe $y$ apparaît pour un entrefer très petit.}
  \label{fig:Int}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{kappa.pdf}
  \caption{Courbure dans le plan de l'interface le long de son abscisse curviligne ($Ca=0.025$ et $h^*=1/20$).}
  \label{fig:marginkappa}
\end{figure}

\clearpage
\section{Code}
\label{sec:orgc0eea86}
\begin{verbatim}
import matplotlib
import numpy as np
matplotlib.use("Agg")
import matplotlib.pyplot as plt
from matplotlib import rc
from scipy.optimize import curve_fit
from scipy.integrate import trapz

matplotlib.rcParams["mathtext.fontset"] = "stix"
matplotlib.rcParams["font.family"] = "STIXGeneral"

# constant
#gamma = 2.23e-6
h = 5e-4 * 0.05
Surf = 5e-4 * 1e-3
gamma = 4.46e-5

xReconstruct = np.linspace(3e-3, 4e-3, num=100000)

# artificial symmetric interface
yArtificial = -5e-5 * np.sin(xReconstruct * np.pi / 1e-3) + 3.5e-4

xy = np.array(data, float)
z = np.polyfit(xy[:, 0], xy[:, 1], 8)
yReconstruct = np.polyval(z, xReconstruct)

# ----COMPUTE GRADIENT
dx = np.gradient(xReconstruct)
ddx = np.gradient(dx)
dy = np.gradient(yArtificial)
ddy = np.gradient(dy)

# ----COMPUTE NORMAL VECTOR
nx = np.divide(-dy, np.sqrt(dx ** 2 + dy ** 2))
ny = np.divide(dx, np.sqrt(dx ** 2 + dy ** 2))

# ----COMPUTE ELEMENT LENGTH
ds = np.sqrt(dx ** 2 + dy ** 2)
sumds = np.cumsum(ds)

# ----COMPUTE CURVATURE
num = np.multiply(dx, ddy) - np.multiply(ddx, dy)
denom = (np.multiply(dx, dx) + np.multiply(dy, dy)) ** (3 / 2)

kappa = np.divide(num, denom)

L = np.sum(ds)

intnx = trapz(nx[2:-2], xReconstruct[2:-2])
intny = trapz(ny[2:-2], xReconstruct[2:-2])

meankappa = trapz(kappa[2:-2], xReconstruct[2:-2]) / L
meankappa1 = np.sum(np.multiply(kappa, ds)) / L

integral = gamma * trapz(nx[2:-2] * (kappa[2:-2] * np.pi / 4 - (2 / h)), xReconstruct[2:-2]) / Surf
integralnx = gamma * trapz(nx[2:-2], xReconstruct[2:-2]) / Surf 
plt.clf()

fig1 = plt.figure(figsize=(3, 3))
plt.plot(xReconstruct[2:-2], yArtificial[2:-2])
plt.xlabel(r"$x$ (m)", fontsize=14)
plt.ylabel(r"$y$ (m)", fontsize=14)
plt.ticklabel_format(axis="x", style="sci", scilimits=(0, 0))
plt.ticklabel_format(axis="y", style="sci", scilimits=(0, 0))
fig1.tight_layout()
# plt.savefig('symInt.pdf')
plt.close()

fig2 = plt.figure(figsize=(3, 3))
plt.plot(sumds[3:-3],kappa[3:-3])
plt.xlabel(r'x curv (m)',fontsize=14)
plt.ylabel(r'$\kappa_{\parallel}$ (1/m)',fontsize=14)
plt.ticklabel_format(axis="x", style="sci", scilimits=(0, 0))
fig2.tight_layout()
#plt.savefig('kappa.pdf')

print("%2d,%2d,%10.2E,%10.2E"% (meankappa,meankappa1,integral,integralnx))
\end{verbatim}
\end{document}
